### 2024-03-14 manuscript.Rmd
init

---
title: ""
date: "`r Sys.Date()`"
author: ""
output: 
  # bookdown::html_document2:
  #   toc: true
  #   number_sections: FALSE
  bookdown::word_document2:
    reference_docx: [./custom-reference-doc.docx]
    number_sections: false
    fig_caption: yes
bibliography: [C:/Users/ST/Documents/repo/anu_thesis_texts/zotero_library.bib]
csl: evolution.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE,
                      message=FALSE,
                      warning=FALSE,
                      dpi=300)
# knitr::opts_knit$set(root.dir=normalizePath("..")) #use grandparent folder working directory
# for including graphics we need full path with knitr
WD <- getwd()
```

```{r packages}
library(dplyr); library(readr); library(janitor); 
library(ape); library(phytools);
library(kableExtra)
```

**Title** TBD

**Authors**: Sarin Tiatragul^1^; Andrew Amey^2^; Scott Hocknull^2^;  Scott Keogh^1^; Emma Sherratt^3^

^1^Division of Ecology & Evolution, Research School of Biology, The Australian National University, Canberra 2601, ACT, Australia 
^2^Queensland Museum, Hendra, QLD, 4011, Australia
^3^School of Biological Sciences, The University of Adelaide, Adelaide, SA 5005, Australia

**Corresponding author**: Sarin Tiatragul (sarin.tiatragul@anu.edu.au)

<!-- **Author contributions**:  -->
Putter:

<!-- ST and ES conceptualised the study; ST collected morphology data;  -->
<!-- ST obtained X-rays from SAM, WAM, and MAGNT. SH and AM provided X-rays Queensland Museum specimens. -->
<!-- ST and ES analysed the data; ST prepared the figures; ST, SK and ES wrote the manuscript. -->

<!-- **Acknowledgements (including funding information)**: -->

<!-- We to thank museum curators and collection managers from the  -->
<!-- South Australian Museum  (R Foster, S Donnellan), Western Australian Museum (P Doughty), Queensland Museum (A Amey and P Couper), Australian National Wildlife Collection (T Haff, C Wilson, and L Joseph), Museum and Art Gallery of the Northern Territory (G Dally), Museums Victoria (J Sumner, J Melville, and R Aguilar), Florida Museum of Natural History (J Gray, CM Sheehy, and ZS Randall), and the Natural History Museum Berlin (K Mahlow and F Tillack).  -->
<!-- SK thanks the Australian Research Council for ongoing support. ST PhD scholarship is supported by the Australian Government Research Training Program Scholarship at the Australian National University.  -->

**Conflict of Interest statement**
The authors declare no Conflict of Interest.

**Data Availability Statement**
Upon acceptance, associated data will be archived on Zenodo/Dryad.

\newpage

#Abstract

\newpage

# Introduction

# Material and Methods

## Sampling

We examined 341 specimens representing 44 lineages *Anilios* blindsnakes including 42/47 formally described species [@uetzQuarter2021]. Our sample size per species ranged from 1 to 25 specimens (median=7 specimens). We sourced specimens from South Australian Museum (SAMA), Western Australian Museum (WAM), Queensland Museum (QM), Museum and Art Gallery of the Northern Territory (MAGNT), and a private collection. Where possible, we chose relatively large non-fragile specimens from both males and females with well-preserved heads for X-ray. Sex for each specimen were determined by inspecting the gonads, or assigned based on relative tail lengths folowing (Tiatragul et al. unpublished).

## Phylogeny
We extracted 37 lineages of Australian blindsnakes from the time-calibrated phylogeny of Australian blindsnakes and outgroups presented in [@tiatragulAustralia2023]. The subset phylogeny represents 35/47 formally described species of _Anilios_ along with two additional genetically distinct lineages: one eastern population of "_A. grypus_" and one eastern population of "_A. ligatus_" [@tiatragulAustralia2023]. This phylogeny was inferred using the shortcut coalescent method in ASTRAL III [@zhangASTRALIIIPolynomialTime2018] based on 4930 maximum likelihood gene trees estimated using IQTREE [@minhIQTREE2020]. Divergence dating was estimated using MCMCTree from the PAML package [@yangPAML2007] based on 27 squamate reptile fossil calibrations [@tiatragulAustralia2023].

## Morphological data

We collected body size measurements and vertebral counts from museum specimens examined. To quantify body size and body width ratios, we measured snout-to-vent length (SVL), tail lengths (TL) and midbody diameter (MBD). We measured SVL and TL by stretching the specimen along a ruler to the nearest 0.5 mm. For MBD, we used a digital vernier calliper (ABS Coolant Proof Caliper, Mitutoyo, Japan) to the nearest 0.1 mm. Total body length (TBL) is the sum of SVL + TL. 
<!-- To obtain  mean body shape ratios [@mosimannSize1970] for each species, we divided each MBD by SVL and log-transformed for further analyses. -->

To obtain vertebral counts, we X-ray scanned the specimens at various institutions including: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n = 87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n = 129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (ANU) (n = 90), and Queensland X-Ray (n = 35).

<!-- Resolution for ANU CTLab is calculated by lenth / pixel * 1000 -->

We used the Fiji distribution of ImageJ2 [@schindelinFiji2012] to digitally visualise X-ray images and counted the number of vertebrae using Cell Counter module. Precloacal and postcloacal vertebrae were distinguished based on vertebral morphology.

Additionally, we quantified the vertebral profile for two representative blindsnake species (*A. grypus* and *A. ligatus*) by measuring the length of each vertebra from the first pre-cloacl vertebra to the last tail vertebra. Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 where landmarks were placed medially along the vertebral column at the anterior limit of the centrum of each vertebra.

## PICK UP HERE $*)#$@

## Environmental data

Occurrence data were based on locality records of preserved specimens from museum databases and verified human observations from Atlas of the Living Australia (ALA). We modified some records to reflect the latest taxonomic changes based on updated distribution range and ST's examination of specimens. We discarded records with identical latitude and longitude, and records with uncertain taxonomic identities. In total we retained 3896 verified records representing every species of Australian *Anilios* for our analyses (See data availability statement).

We selected soil bulk density, annual mean temperature, and aridity index as the most biologically relevant data to describe blindsnake niches. 
Soil bulk density measures the penetrability of soil, which is relevant for strictly fossorial reptiles [camaitiMacroecological2022]. Bulk density is calculated by taking the weight of dry soil divided by the total soil volume (g/cm<sup>3</sup>). Values > 1.6 g/cm<sup>3</sup> tend to restrict root growth, while loose soil have densities < 0.5 g/cm<sup>3</sup> [@mckenzieSoil2002]. Mean annual temperature serves as a climate proxy and is commonly used to describe niche for squamate reptiles [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013]. Finally, aridity index reflects precipitation availability over atmospheric water demand and is categorised as follows: hyper arid, < 0.03; arid, 0.03 – 0.2; semi-arid, 0.2-0.5; dry sub-humid, 0.5 – 0.65; humid, >0.65 [@zomerVersion2022]. 
From raster files that contained modelled approximations at the resolution of 3 arc-seconds, we extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html), mean annual temperature from WorldClim v.2 [@fickWorldClim2017], and aridity index from @zomerVersion2022 for each georeferenced record using the "extract" function in the R package `raster` v.3.6-3 [@hijmansRaster2022]. We then calculated the maximum soil bulk density, mean annual temperature, and mean aridity index for each species. Correlation analyses revealed that none of the background environmental variables were strongly correlated (Pearson's *r* < |0.75|).

#### Statistical analyses

Phylogenetic generalised least squares (PGLS). Body size as dependent variable regressed onto independent vertebral counts. Use

Use partial least squares to test correlation between body width and vertebrae ratio.

# Discussion

## Miscellaneous 
Traditionally, ventral scale counts have been used as a proxy for vertebral number as they correlate for most alethinophidian snakes [@alexanderPattern1966]. We found that this is (not) the case in blindsnakes.

### 2024-03-14 code/03_visualising_data.R
code

@@ -10,6 +10,7 @@
 library(ggplot2)
 # library(dplyr)
 library(phytools); 
+library(geiger);
 library(ape); 
 library(geomorph); 
 # library(geiger)
@@ -75,7 +76,7 @@ dimorph_df <- vert_df[vert_df$sex %in% c('f', 'm') &
 # Vector of species where sample size is at least 3 for males and females
 include_species <- dimorph_df %>% 
   dplyr::group_by(species, sex) %>% 
-  dplyr::summarise(n = n()) %>% 
+  dplyr::summarise(n = dplyr::n()) %>% 
   dplyr::filter(n >= 3) %>% dplyr::distinct(species) %>% 
   dplyr::filter(species %notin% c("Anilios_systenos")) %>%
   c()
@@ -88,7 +89,7 @@ source('code/utility/func_sexual_dimorphism_tester.R')
 d_vert_ratio <- sex_dimorphic_tester("vert_ratio ~ sex + species + sex:species", subset_dimorph)
 d_vert_ratio
 
-## ANSWER: 4/29 species are sexually dimorphic. 
+## ANSWER: 4/29 species are sexually dimorphic in vertebrae/length ratio.
 d_vert_ratio$dimorph_sp
 
 
@@ -169,7 +170,7 @@ vert_data <- as.data.frame(vert_data)
 rownames(vert_data) <- vert_data$species
 vert_data$ratio <- vert_data$mean_tot_vert/ vert_data$mean_total_length
 
-check_data <- name.check(phy = sub_phy, data = vert_data)
+check_data <- geiger::name.check(phy = sub_phy, data = vert_data)
 vert_data <- vert_data[which(rownames(vert_data) %notin% check_data$data_not_tree),]
 
 plotTree.barplot(sub_phy, setNames(vert_data$ratio, rownames(vert_data)), 
@@ -185,14 +186,15 @@ plotTree.barplot(sub_phy, setNames(vert_data$ratio, rownames(vert_data)),
 
 # Calculate width ratio
 width_ratio_cex <- vert_data$body_ratio*90
-width_ratio_cex <- setNames(obj = width_ratio, rownames(vert_data))
+width_ratio_cex <- setNames(obj = width_ratio_cex, rownames(vert_data))
 
 dev.off()
-plot(mean_total_length ~ mean_tot_vert, cex = width_ratio_cex, data = vert_data, bty="n", pch = 19)
[…]

### 2024-03-14 code/00_stats_for_paper.R
stats for paper

@@ -76,6 +76,8 @@ mean(table(vert_df$species))
 ## Institution we measured from
 unique(vert_df$institution)
 
+## Number of scans by institution
+table(vert_df$institution)
 
 
 

### 2024-03-25 manuscript.Rmd
update methods

@@
 Precloacal and postcloacal vertebrae were distinguished based on vertebral morphology.
-
-Additionally, we quantified the vertebral profile for two representative blindsnake species (*A.
+Additionally, we quantified intracolumnar length of each vertebrae for two representative blindsnake species (*A.
 grypus* and *A.
-ligatus*) by measuring the length of each vertebra from the first pre-cloacl vertebra to the last tail vertebra.
-Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 where landmarks were placed medially along the vertebral column at the anterior limit of the centrum of each vertebra.
-
-## PICK UP HERE $*)#$@
+ligatus*).
+Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra.
+We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R.
+Measurements were repeated three times to acccount for measurement errors.
 
@@
 
-Occurrence data were based on locality records of preserved specimens from museum databases and verified human observations from Atlas of the Living Australia (ALA).
-We modified some records to reflect the latest taxonomic changes based on updated distribution range and ST's examination of specimens.
-We discarded records with identical latitude and longitude, and records with uncertain taxonomic identities.
-In total we retained 3896 verified records representing every species of Australian *Anilios* for our analyses (See data availability statement).
+<!-- NEED TO PUT IN THE CORRECT VALUES HERE -->
+We used the Australian blindsnake occurrence records presented in (Tiatragul et al.
+unpublished).
+These records were compiled from museum databases and verified human observations from Atlas of the Living Australia (ALA).
+A total of XXXX records (min = , median = , max = , median per species) representing XX species of Australian *Anilios* for our analyses.
 
+###<!-- PICK UP HERE -->###
 We selected soil bulk density, annual mean temperature, and aridity index as the most biologically relevant data to describe blindsnake niches. 

### 2024-03-25 code/utility/func_vert_profile_plotter.R
function to plot profile

@@ -1,0 +1,47 @@
+# func_vert_profile_plotter.R
+
+# Function to plot vertebral profile
+
+vert_profile_plotter <- function(.data1, .data2, .data3, .title = NULL){
+  
+  rep1 <- .data1
+  rep2 <- .data2
+  rep3 <- .data3
+  
+  # Calculate the lengths of each vertebra from the landmark coordinates
+  lindist1 <- diag(as.matrix(dist(rep1))[-1,]) 
+  lindist2 <- diag(as.matrix(dist(rep2))[-1,])
+  lindist3 <- diag(as.matrix(dist(rep3))[-1,])
+  
+  # Combine into a single matrix
+  lindist <- cbind(lindist1, lindist2, lindist3)
+  
+  # Average the vertebrae lengths across the three replicates
+  avg <- apply(lindist, 1, mean)
+  
+  # Calculate the standard deviation for each vertebra length
+  sdev <- apply(lindist, 1, sd)
+  
+  # Calculate the standard error for each vertebra length
+  sdevm <- sdev / sqrt(nrow(lindist))
+  mean(sdevm) # calculate your repeatability score (closest to 0 wins!)
+  
+  # Calculate the body length, which is the sum of the measurements you've taken
+  sum(avg)
+  
+  # Get the number of vertebrae you've measured
+  length(avg)
+  
+  # Plot a graph to show the pattern of intracolumnar variation in vertebrae size
+  plot(avg, pch=19, xlab = "vertebra number", 
+       ylab= "vertebra length (mm)", bty = "n",
+       ylim=c(0, max(avg+sdev)))
+  # Add error bars
+  arrows(c(1:length(avg)), avg-sdevm, 
+         c(1:length(avg)), avg+sdevm, 
+         length=0.05, angle=90, code=3)
+  # Add a title
+  title(paste(.title)) 
+  # change "My Name & Species name & body length & Number Vertebrae" to your info
+  
+}

### 2024-03-25 code/03_vert_profile.R
init

@@ -1,0 +1,26 @@
+# 03_vert_profile.R
+# 2024-03-25
+# Adapted from Emma's script for Measuring and analysing vertebral length of sea snakes along the axial column
+# October 2021
+
+
+### READ ME! ###
+# Comments are instructions
+# make changes where indicated
+
+##### DO FIRST! ######
+# Set working directory
+# setwd(" ") # paste between the quotation the path to working directory, OR
+
+
+# Data --------------------------------------------------------------------
+# Read in the three sets of replicate measurements for a single specimen
+lig1 <- read.csv("data/x_ray/lig_R00050_ST_1.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+lig2 <- read.csv("data/x_ray/lig_R00050_ST_2.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+lig3 <- read.csv("data/x_ray/lig_R00050_ST_3.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+
+# Plot  -------------------------------------------------------------------
+source('code/utility/func_vert_profile_plotter.R')
+vert_profile_plotter(lig1, lig2, lig3, "A. ligatus R00050")
+
+

### 2024-03-28 code/03_visualising_data.R
barplot for aspet ratio

@@ -52,7 +52,7 @@ linear_df$species <- genus_species
 # Check vertebrae by length -----------------------------------------------
 
 # Filter out rows with NA or 0 values in precloacal_vert
-vert_df <- linear_df[!is.na(linear_df$precloacal_vert) & linear_df$precloacal_vert != 0, ]
+vert_df <- linear_df[!is.na(linear_df$precloacal_vert) & linear_df$precloacal_vert != 0 & linear_df$reg_no != "R152714", ]
 
 # Calculate total vertical length and vertical ratio
 vert_df$total_vert <- vert_df$precloacal_vert + vert_df$postcloacal_vert
@@ -97,7 +97,7 @@ d_vert_ratio$dimorph_sp
 
 ## Question: How does number of vertebrae vary with total body length by species and sex
 
-filtered_vert_df <- vert_df %>% dplyr::group_by(species) %>% dplyr::filter(n() > 4, sex %in% c("f","m")) 
+filtered_vert_df <- vert_df %>% dplyr::group_by(species) %>% dplyr::filter(n() > 4, sex %in% c("f","m"), reg_no != "R152714") 
 
 ggplot(filtered_vert_df, aes(x = total_vert, y = total_length, colour = sex)) +
   geom_point() + 
@@ -164,8 +164,12 @@ vert_data <- vert_df %>% dplyr::group_by(species) %>%
                    mean_preclo = mean(precloacal_vert),
                    mean_post = mean(postcloacal_vert),
                    mean_tot_vert = mean(precloacal_vert + postcloacal_vert),
-                   body_ratio = mean(midbody_diameter/svl))
+                   body_ratio = mean(midbody_diameter/svl),
+                   aspect = mean(total_length/midbody_diameter))
+
+
 vert_df
+
 vert_data <- as.data.frame(vert_data)
 rownames(vert_data) <- vert_data$species
 vert_data$ratio <- vert_data$mean_tot_vert/ vert_data$mean_total_length
@@ -175,6 +179,30 @@ vert_data <- vert_data[which(rownames(vert_data) %notin% check_data$data_not_tre
 
 plotTree.barplot(sub_phy, setNames(vert_data$ratio, rownames(vert_data)), 
                  args.barplot=list(xlab="Vertebrae/Length ratio"))
+
+# Aspect ratio as boxplot
+aspect_v <- setNames(vert_df$total_length / vert_df$midbody_diameter, nm = vert_df$species)
[…]

### 2024-03-28 code/03_vert_profile.R
plot new profile

@@ -19,8 +19,15 @@ lig1 <- read.csv("data/x_ray/lig_R00050_ST_1.csv", header=TRUE, row.names = 1)[,
 lig2 <- read.csv("data/x_ray/lig_R00050_ST_2.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
 lig3 <- read.csv("data/x_ray/lig_R00050_ST_3.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
 
+
+gry1 <- read.csv("data/x_ray/gry_R146673_ST_1.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+gry2 <- read.csv("data/x_ray/gry_R146673_ST_2.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+gry3 <- read.csv("data/x_ray/gry_R146673_ST_3.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+
 # Plot  -------------------------------------------------------------------
 source('code/utility/func_vert_profile_plotter.R')
-vert_profile_plotter(lig1, lig2, lig3, "A. ligatus R00050")
 
-
+par(mfrow=c(1,2))
+vert_raw_plotter(lig1, lig2, lig3, "A. ligatus R00050")
+vert_profile_plotter(lig1, lig2, lig3, "A. ligatus R00050", .error = TRUE)
+vert_profile_plotter(gry1, gry2, gry3, "A. grypus R146673", .error = TRUE)

### 2024-03-28 code/utility/func_plotTreeboxplot.R
functions

@@ -1,0 +1,114 @@
+# func_plotTreeboxplot.R
+# This is a working version from: http://blog.phytools.org/2016/07/new-function-to-print-box-plot-next-to.html
+
+plotTree.boxplot<-function(tree,x,args.plotTree=list(),
+                           args.boxplot=list()){
+  cw<-reorder(tree)
+  if(!is.list(x)&&class(x)!="formula"){
+    obj<-setNames(
+      lapply(cw$tip.label,function(x,y) y[which(names(y)==x)],
+             y=x),cw$tip.label)
+  } else obj<-x
+  if(class(x)=="formula") 
+    args.boxplot$formula<-obj else args.boxplot$x<-obj
+    args.boxplot$horizontal<-TRUE
+    args.boxplot$axes<-FALSE
+    args.boxplot$names.arg<-""
+    args.boxplot$xlim<-c(1,Ntip(cw))
+    args.boxplot$ylim<-NULL
+    if(is.null(args.boxplot$space)) args.boxplot$space<-0.7
+    if(is.null(args.boxplot$mar)) 
+      args.boxplot$mar<-c(5.1,0,2.1,1.1)
+    else args.boxplot$mar[2]<-0.1
+    args.plotTree$tree<-cw
+    if(is.null(args.plotTree$mar)) 
+      args.plotTree$mar<-c(5.1,1.1,2.1,0)
+    else {
+      args.plotTree$mar[4]<-0
+    }
+    if(args.plotTree$mar[1]!=args.boxplot$mar[1])
+      args.plotTree$mar[1]<-args.boxplot$mar[1]
+    if(args.plotTree$mar[3]!=args.boxplot$mar[3])
+      args.plotTree$mar[3]<-args.boxplot$mar[3]
+    if(is.null(args.plotTree$ftype)) args.plotTree$ftype<-"i"
+    if(is.null(args.plotTree$lwd)) args.plotTree$lwd<-1
+    par(mfrow=c(1,2))
+    do.call(plotTree,args.plotTree)
+    par(mar=args.boxplot$mar)
+    ii<-which(names(args.boxplot)%in%c("formula","x"))
+    args.boxplot<-c(args.boxplot[ii],args.boxplot[-ii])
+    obj<-do.call(boxplot,args.boxplot)
+    axis(1)
+    if(!is.null(args.boxplot$xlab)) title(xlab=args.boxplot$xlab)
+    else title(xlab="x")
+    invisible(obj)
[…]

### 2024-03-28 code/utility/func_vert_profile_plotter.R

@@ -2,7 +2,7 @@
 
 # Function to plot vertebral profile
 
-vert_profile_plotter <- function(.data1, .data2, .data3, .title = NULL){
+vert_raw_plotter <- function(.data1, .data2, .data3, .title = NULL){
   
   rep1 <- .data1
   rep2 <- .data2
@@ -45,3 +45,60 @@ vert_profile_plotter <- function(.data1, .data2, .data3, .title = NULL){
   # change "My Name & Species name & body length & Number Vertebrae" to your info
   
 }
+
+
+vert_profile_plotter <- function(.data1, .data2, .data3, .title = NULL, .error = FALSE){
+  
+  rep1 <- .data1
+  rep2 <- .data2
+  rep3 <- .data3
+  
+  # Calculate the lengths of each vertebra from the landmark coordinates
+  lindist1 <- diag(as.matrix(dist(rep1))[-1,]) 
+  lindist2 <- diag(as.matrix(dist(rep2))[-1,])
+  lindist3 <- diag(as.matrix(dist(rep3))[-1,])
+  
+  # Combine into a single matrix
+  lindist <- cbind(lindist1, lindist2, lindist3)
+  
+  # Average the vertebrae lengths across the three replicates
+  avg <- apply(lindist, 1, mean)
+  
+  # Calculate the standard deviation for each vertebra length
+  sdev <- apply(lindist, 1, sd)
+  
+  # Calculate the standard error for each relative vertebra length
+  sdevm <- (sdev/sum(sdev)) / sqrt(nrow(lindist))
+  
+  # Calculate relative size
+  relative_size <- avg / sum(avg)
+  
+  
+  # Fit polynomial 4 degree
+  polynomial_fit <- lm(relative_size ~ poly(1:length(relative_size), degree = 4, raw = TRUE))
+  
+  # Generate equally spaced points along the x-axis
+  x_values <- seq(1, length(relative_size), length.out = 100)
+  
+  # Predict corresponding y-values using the fitted polynomial
+  predicted_points <- predict(polynomial_fit, newdata = data.frame(x_values))
+  
+  # Plot the profile
+  plot(relative_size, xlab = "Vertebra Number (index)", ylab = "Relative Vertebra Size")
+  title(paste(.title)) 
+  
+  # Add error bars if .error is TRUE
+  if (.error) {
+    arrows(c(1:length(relative_size)), relative_size-sdevm, 
[…]

### 2024-03-28 code/03_visualising_data.R
update analyses pgls

@@ -35,6 +35,8 @@ PC_log <- read.csv('data/script_generated_data/log_body_shape_ratio_pc_all.csv')
 
 # Full individual data
 linear_df <- read.csv('data/script_generated_data/blindsnake_body_traits.csv')
+linear_df <- linear_df[linear_df$reg_no %notin% c("R.8515","R.127956"),] ### Dropped these samples because mis-id as troglodytes but in Queensland. Probably torresianus or broomi
+
 
 # Summary data
 anilios_data <- read.csv(file = 'data/script_generated_data/anilios_summary_data.csv', row.names = 1)
@@ -49,6 +51,7 @@ anilios_tree$tip.label
 genus_species <- paste(linear_df$genus, linear_df$species, sep = "_")
 linear_df$species <- genus_species
 
+
 # Check vertebrae by length -----------------------------------------------
 
 # Filter out rows with NA or 0 values in precloacal_vert
@@ -62,6 +65,68 @@ plot(x = vert_df$total_length, y = vert_df$total_vert, xlab = "Total vertebrae #
 
 # No clear pattern of vertebrae and total length, indicates there may be some species that elongate their vertebrae and others that shorten their vertebrae relative to the mean. 
 
+
+# Total number of vertebrae -----------------------------------------------
+
+source('code/utility/func_plotTreeboxplot.R')
+
+# Total number of vertebrae
+total_vertebrae <- setNames(vert_df$total_vert, nm = vert_df$species)
+total_vertebrae <- total_vertebrae[which(names(total_vertebrae) %in% sub_phy$tip.label)]
+geiger::name.check(phy = sub_phy, data = total_vertebrae)
+spp<-factor(names(total_vertebrae),untangle(ladderize(sub_phy),"read.tree")$tip.label)
+
+pdf(file = "output/tree_boxplot_totalVert.pdf", width = 11.7, height = 8.3, paper = "a4r")
+plotTree.boxplot(sub_phy,x=total_vertebrae~spp,
+                 args.boxplot = list(xlab="Total number of vertebrae",
+                                     # ylim=c(20,170),
+                                     col="grey99"))
+dev.off()
+
+summary(vert_df$total_vert)
+vert_df[order(vert_df$total_vert), c("species", "total_vert")]
+
+## Summary SE for species
+Vertebra_summary_table <- vert_df %>% 
+  dplyr::group_by(species) %>% 
[…]

### 2024-03-28 manuscript.Rmd
added intro, methods, and results

@@
 
+Blindsnakes are
+
+Scolecophidians comprise XXX species of snakes with highly specialised phenotypes for a strictly fossorial life-style [@mirallesMolecular2018].
+
+Although all blindsnakes are fossorial, body size and shape vary significantly among species (Tiatragul et al.
+unpublished; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993).
+Leptophlopids are generally thought to be more narrow and long compared to typhlopids, which are wider and more robust for a given snout-vent length [@herrelBurrowing2021].
+These differences in body size and proportions determine how much push force the snakes can generate when burrowing.
+
+Their morphological differences led to the idea that leptyphlopids generally use existing burrows while typhlopids create their own burrow.
+
+Compared to other snakes, the vertebral column of fossorial snakes tend to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001].
+This is reflected in the relative short tails of blindsnakes (CITE).
+
+The Australian blindsnake radiation comprise 47 described species [@tiatragulAustralia2023] with exceptionally diverse morphology from slender (aspect ratio = ) to robust.
+Body proportion exhibit strong phylogenetic signal but in some instances spread out across the phylogeny (CITE).
+
+In this study, we investigate the interspecific variation in the number of pre- and post-cloacal vertebrae to test the hypothesis that changes in developmental patterns underpin the evolution of different body proportions in Australian typhlopid snakes.
+
+Specifically, we test two predictions:
+(1) narrow-and-long species develop from postnatal ontogenetic changes where vertebral numbers are not significantly different between narrow-and-long species compared to wide-and-robust species; but are elongated.
+
+(2) Narrow-and-long blindsnakes develop via changes in the embryonic and postnatal stages where narrow-and-long species have greater number of vertebrae than wide-and-robust species.
+
+To test these predictions, we quantified the number of pre- and post-clocal vertebra across 37 species of Australian blindsnakes.
+
 # Material and Methods
@@
 
-To obtain vertebral counts, we X-ray scanned the specimens at various institutions including: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n = 87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n = 129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (ANU) (n = 90), and Queensland X-Ray (n = 35).
+To obtain vertebral counts, we X-ray scanned the specimens at various institutions including: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n=87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n=129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (ANU) (n=90), and Queensland X-Ray (n=35).
 
@@
 We used the Fiji distribution of ImageJ2 [@schindelinFiji2012] to digitally visualise X-ray images and counted the number of vertebrae using Cell Counter module.
-Precloacal and postcloacal vertebrae were distinguished based on vertebral morphology.
+Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023].
 Additionally, we quantified intracolumnar length of each vertebrae for two representative blindsnake species (*A.
@@
 We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R.
-Measurements were repeated three times to acccount for measurement errors.
+Measurements were repeated three times to account for measurement errors.
 
-## Environmental data
-
+<!-- ## Environmental data  -->
 <!-- NEED TO PUT IN THE CORRECT VALUES HERE -->
-We used the Australian blindsnake occurrence records presented in (Tiatragul et al.
+<!-- We used the Australian blindsnake occurrence records presented in (Tiatragul et al.
 unpublished).
@@
 
-###<!-- PICK UP HERE -->###
 We selected soil bulk density, annual mean temperature, and aridity index as the most biologically relevant data to describe blindsnake niches. 
@@
 Correlation analyses revealed that none of the background environmental variables were strongly correlated (Pearson's *r* < |0.75|).
-
+ -->
 #### Statistical analyses
 
-Phylogenetic generalised least squares (PGLS). Body size as dependent variable regressed onto independent vertebral counts. Use
+All statistical analyses were performed in R v.4.31 [@rcoreteamLanguage2023].
+To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Browninan motion implemented in `geomorph` v.4.0.6 [@adamsGeomorph2023].
+First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes.
+We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum SVL from Tiatragul et al.
+(unpublished).
+Then, we test whether the total vertebrae number to total body length ratio correlates with body shape, defined as the ratio between midbody diameter divided by SVL, also from Tiatragul et al.
+(unpublished) data.
 
-Use partial least squares to test correlation between body width and vertebrae ratio.
+# Results
+
+Total number of vertebrae varies among the 39 species of Australo-Papuan blindsnakes in this study (Fig.
+\@ref(fig:fig-variation-totalvert)), ranging from 146 in *A.
+tovelli* to 400 in *A.
+grypus* ($$Table 1$$).
+
+We found no relationship between indicating maximum number of total vertebrae and maximum SVL (PGLS, slope = -0.08, y=225, R^2^=0.02, F=0.9, Z=0.40, P=0.361), indicating an absence of pleomerism in Australian blindsnakes.
+
+Vertebrae ratio is negatively correlated with body width ratio (PGLS, slope = -21.4, y=1.4, R^2^=0.2, F=8.8,, Z=2.3, P=0.007), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig.
+\@ref(fig:fig-vert-width-ratios)).
+
+```{r fig-variation-totalvert, fig.cap= 'Interspecific variation in total number of vertebrae among Australo-Papuan blindsnake and close relatives.
+The boxplot represent the distribution of the data, with vertical inside representing the median.
+The extent of the box represent the lower (left) and upper (right) quartile, with the whiskers showing 1.5 times the interquartile range.
+Outliers beyond this range are depicted as individual points.', out.width="165mm", out.height= "126mm"}
+
+knitr::include_graphics(paste0(WD, '/assets/figures/tree-boxplot-totalVert.jpg'))
+```
+
+<!-- Figure 2: Pleomerism and ratio correlations -->
+```{r fig-vert-width-ratios, fig.cap= 'Relationship between mean vertebrae ratio and mean body width ratio among Australian blindsnakes.
+Vertebrae ratio is calculated by total vertebrae/total length, and body width ratio is midbody diameter/total body length.', out.width="165mm", out.height= "126mm"}
+
+knitr::include_graphics(paste0(WD, '/assets/figures/vert-width-ratios.jpg'))
+```
+
+<!-- ### Variation in vertebrae number -->
+
+Closely related species tend to have similar number of total vertebrae. 
+Males tend to have greater number of postcloacal vertebrae
+
+<!-- FIGURE 1 -->
+
+There is a statistically significant correlation between vertebrae ratio and fineness of the body.
+Species that have more vertebrae per length (vertebrae ratio > 1) are usually more fine, whereas more robust species have lower number of vertebrae.
 

### 2024-03-29 manuscript.Rmd
update analysis and results

@@
 
-The Australian blindsnake radiation comprise 47 described species [@tiatragulAustralia2023] with exceptionally diverse morphology from slender (aspect ratio = ) to robust.
-Body proportion exhibit strong phylogenetic signal but in some instances spread out across the phylogeny (CITE).
+The Australian blindsnake radiation comprise 47 described species [@tiatragulAustralia2023] with considerably diverse morphology from narrow-and-long (aspect ratio >= 100) to robust-and-wide (aspect ratio < 60).
 
-In this study, we investigate the interspecific variation in the number of pre- and post-cloacal vertebrae to test the hypothesis that changes in developmental patterns underpin the evolution of different body proportions in Australian typhlopid snakes.
+In this study, we investigate the interspecific variation in the number of vertebrae to test the hypothesis that changes in developmental patterns underpin the evolution of different body shapes in Australian typhlopid snakes.
 
@@
 (1) narrow-and-long species develop from postnatal ontogenetic changes where vertebral numbers are not significantly different between narrow-and-long species compared to wide-and-robust species; but are elongated.
-
 (2) Narrow-and-long blindsnakes develop via changes in the embryonic and postnatal stages where narrow-and-long species have greater number of vertebrae than wide-and-robust species.
@@
 ## Phylogeny
+
 We extracted 37 lineages of Australian blindsnakes from the time-calibrated phylogeny of Australian blindsnakes and outgroups presented in [@tiatragulAustralia2023].
@@
 Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023].
-Additionally, we quantified intracolumnar length of each vertebrae for two representative blindsnake species (*A.
-grypus* and *A.
-ligatus*).
+Additionally, we quantified intracolumnar length of each vertebrae for three species representing narrow-and-wide phenotypes *A.
+grypus*, wide-and-robust phenotypes (*A.
+pinguis*), and intermediate phenotypes (*A.
+centralis*).
 Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra.
@@
  -->
-#### Statistical analyses
+
+## Statistical analyses
 
@@
 First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes.
-We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum SVL from Tiatragul et al.
+We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum total body length from Tiatragul et al.
 (unpublished).
-Then, we test whether the total vertebrae number to total body length ratio correlates with body shape, defined as the ratio between midbody diameter divided by SVL, also from Tiatragul et al.
+Then, we test whether the vertebrae to body length ratio is related to maximum aspect ratio, defined by total body length / body width, also from Tiatragul et al.
 (unpublished) data.
@@
 
-We found no relationship between indicating maximum number of total vertebrae and maximum SVL (PGLS, slope = -0.08, y=225, R^2^=0.02, F=0.9, Z=0.40, P=0.361), indicating an absence of pleomerism in Australian blindsnakes.
+We found a weakly positive but non statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, y=256, R^2^=0.02, F=0.72, Z=0.29, P=0.42), indicating an absence of pleomerism in Australian blindsnakes (Fig.
+\@ref(fig: fig-tbl-vert)).
 
-Vertebrae ratio is negatively correlated with body width ratio (PGLS, slope = -21.4, y=1.4, R^2^=0.2, F=8.8,, Z=2.3, P=0.007), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig.
+However, we found that vertebrae ratio is postively correlated with aspect ratio (PGLS, slope=-0.0048, y=0.73, R^2^=0.17, F=7.09,, Z=2.2, P=0.011), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig.
 \@ref(fig:fig-vert-width-ratios)).
@@
 
-<!-- Figure 2: Pleomerism and ratio correlations -->
-```{r fig-vert-width-ratios, fig.cap= 'Relationship between mean vertebrae ratio and mean body width ratio among Australian blindsnakes.
-Vertebrae ratio is calculated by total vertebrae/total length, and body width ratio is midbody diameter/total body length.', out.width="165mm", out.height= "126mm"}
+<!-- Fig. Vertebral counts vs max body length  -->
+```{r fig-tbl-vert, fig.cap= 'Maximum vertebral counts plotted against maximum total body length for Australian typhlopids.
+Trend line represent PGLS regression for Australian species.', out.width="165mm", out.height= "126mm"}
 
-knitr::include_graphics(paste0(WD, '/assets/figures/vert-width-ratios.jpg'))
+knitr::include_graphics(paste0(WD, '/assets/figures/tbl-vert.jpg'))
+```
+
+<!-- Fig: Pleomerism and ratio correlations -->
+```{r fig-vert-aspect-ratios, fig.cap= '(B) Relationship between mean vertebrae and aspect ratios among Australian blindsnake species.
+Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter.
+Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and wide', out.width="165mm", out.height= "126mm"}
+
+knitr::include_graphics(paste0(WD, '/assets/figures/vert-aspect.jpg'))
 ```

### 2024-03-29 code/03_visualising_data.R

@@ -59,7 +59,7 @@ vert_df <- linear_df[!is.na(linear_df$precloacal_vert) & linear_df$precloacal_ve
 
 # Calculate total vertical length and vertical ratio
 vert_df$total_vert <- vert_df$precloacal_vert + vert_df$postcloacal_vert
-vert_df$vert_ratio <- vert_df$total_vert / vert_df$svl
+vert_df$vert_ratio <- vert_df$total_vert / vert_df$total_length
 
 plot(x = vert_df$total_length, y = vert_df$total_vert, xlab = "Total vertebrae #", ylab = "Total length (mm)", bty = "n")
 
@@ -120,12 +120,6 @@ Vertebra_summary_table_by_sex <- vert_df %>%
             )
 
 Vertebra_summary_table_by_sex
-
-
-# Pleomerism --------------------------------------------------------------
-
-
-
 
 # Q1 Test for sexual dimorphism ----------------------------------------------
 
@@ -238,8 +232,12 @@ vert_data <- vert_df %>% dplyr::group_by(species) %>%
                    mean_preclo = mean(precloacal_vert),
                    mean_post = mean(postcloacal_vert),
                    mean_tot_vert = mean(precloacal_vert + postcloacal_vert),
-                   body_ratio = mean(midbody_diameter/svl),
-                   aspect = mean(total_length/midbody_diameter))
+                   body_ratio = mean(midbody_diameter/total_length),
+                   aspect = mean(total_length/midbody_diameter),
+                   max_total_vert = max(total_vert),
+                   max_preclo = max(precloacal_vert),
+                   max_svl = max(svl),
+                   max_tbl = max(total_length))
 
 
 vert_df
@@ -309,7 +307,7 @@ anilios_data <- anilios_data[rownames(anilios_vert), ]
 anilios_data$pre_cloa <- round(anilios_vert$mean_preclo, 0)
 anilios_data$tot_vert <- anilios_vert$mean_tot_vert
 anilios_data$ver_rati <- anilios_vert$ratio
-anilios_data$width_ratio <- anilios_data$mean_width/anilios_data$svl
+anilios_data$width_ratio <- anilios_data$mean_width/anilios_data
 
 # Fit and compare what evolutionary model fit best
 fit.phylolm.ev <- function(trait, .data, .phy){
[…]

### 2024-04-01 manuscript.Rmd
structuring question and some discussion

@@
 
-In this study, we investigate the interspecific variation in the number of vertebrae to test the hypothesis that changes in developmental patterns underpin the evolution of different body shapes in Australian typhlopid snakes.
+In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes.
+The specific questions we addressed are:
 
-Specifically, we test two predictions:
-(1) narrow-and-long species develop from postnatal ontogenetic changes where vertebral numbers are not significantly different between narrow-and-long species compared to wide-and-robust species; but are elongated.
-(2) Narrow-and-long blindsnakes develop via changes in the embryonic and postnatal stages where narrow-and-long species have greater number of vertebrae than wide-and-robust species.
+1. How prevalent is pleomerism among blindsnakes?
 
-To test these predictions, we quantified the number of pre- and post-clocal vertebra across 37 species of Australian blindsnakes.
+2. Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes?
+
+3. Do vertebral number and body size vary with mean annual temperature?
+
+<!-- Specifically, we test two predictions: -->
+<!-- (1) narrow-and-long species develop from postnatal ontogenetic changes where vertebral numbers are not significantly different between narrow-and-long species compared to wide-and-robust species.
+-->
+<!-- (2) Narrow-and-long blindsnakes develop via changes in the embryonic and postnatal stages where narrow-and-long species have greater number of vertebrae than wide-and-robust species.  -->
+
+To answer these questions we gathered data on the number of vertebrae across XX species of Australian blindsnakes.
 
@@
 Divergence dating was estimated using MCMCTree from the PAML package [@yangPAML2007] based on 27 squamate reptile fossil calibrations [@tiatragulAustralia2023].
+This phylogeny is was used for all phylogenetic comparative analyses.
 
@@
 
+### Prevalence of pleomerism among blindsnakes
+
 Total number of vertebrae varies among the 39 species of Australo-Papuan blindsnakes in this study (Fig.
@@
 grypus* ($$Table 1$$).
-
-We found a weakly positive but non statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, y=256, R^2^=0.02, F=0.72, Z=0.29, P=0.42), indicating an absence of pleomerism in Australian blindsnakes (Fig.
-\@ref(fig: fig-tbl-vert)).
-
-However, we found that vertebrae ratio is postively correlated with aspect ratio (PGLS, slope=-0.0048, y=0.73, R^2^=0.17, F=7.09,, Z=2.2, P=0.011), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig.
-\@ref(fig:fig-vert-width-ratios)).
 
@@
 ```
+
+For the Australian blindsnakes, we found a weakly positive but non statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, y=256, R^2^=0.02, F=0.72, Z=0.29, P=0.42), indicating an absence of pleomerism in Australian blindsnakes (Fig.
+\@ref(fig: fig-tbl-vert)).
+This suggest that maximum length is acheived by elongating each vertebrae rather than by adding additional vertebrae
+
+### Relationship between pleomerism and body shape
+However, we found that vertebrae ratio is postively correlated with aspect ratio (PGLS, slope=-0.0048, y=0.73, R^2^=0.17, F=7.09,, Z=2.2, P=0.011), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig.
+\@ref(fig:fig-vert-width-ratios)).
+This is consistent with colubrids and viperids, where species that are more elongated tend to have greater number of vertebrae than stouter ones [@lindellEvolution1994].
 
@@
 
+### Correlation with mean annual temperature among Australian blindsnakes
+
 <!-- ### Variation in vertebrae number -->
@@
 
+Summary about what the paper found
+
+Not so surprising but shows that pleomerism does not explain variation in body size in this continental radiation of blindsnakes.
+Only real giant taxa like *Rhinotyphlops* and *Afrotyphlops* tend to break the rule.
+However, when fitted in our PGLS, we found that even they are not exceptional???
+
+Body shape varied with number of vertebrae in similar ways as previously reported in vipers and colubrids [@lindellEvolution1994].
+We can speculate that species that are narrow-and-long may be using pre-existing burrows rather than digging their own burrow.
+Narrow-and-wide phenotypes are common among Leptotyphlopidae, which are thought to use existing burrows rather than dig their own burrow [@herrelBurrowing2021].
+The Australian group is quite exceptional because within the same genus, their is great variation in aspect ratio so species that evolved more narrow-and-long bodies may be converging on the lifestyle with other blindsnakes that have this narrow-and-long phenotype.
+On the other hand, large robust-and-wide species are expected to be able to generate much more force and make their own burrows.
+A previous study (Tiatragul et al.
+unpublished) found that species that robust-and-wide species tend to occupy areas in Australia that have easier to penetrate soil compared to narrow-and-long species that tend to occupy areas with more difficult to penetrate soil.
+However, more detailed microhabitat observations is required to determine if body shape is related to the environment in which they burrow.
+
+Temperature seem to have XXX effect.
+Common garden experiments for different species of snakes can tell us if incubation temperature and thus rate of development influences number of vertebrae.
+We would expect higher temperature to accelerate development (CITE anolis papers, turtles, etc), so less time for embryonic development.
+
 ## Miscellaneous 
@@
 We found that this is (not) the case in blindsnakes.
+Blindsnakes are generally small enough to fit in an X-ray scanner so more people should do it.

### 2024-04-04 code/03_visualising_data.R
include option to slice the data so we just get the max svl individual

@@ -233,17 +233,30 @@ vert_data <- vert_df %>% dplyr::group_by(species) %>%
                    mean_post = mean(postcloacal_vert),
                    mean_tot_vert = mean(precloacal_vert + postcloacal_vert),
                    body_ratio = mean(midbody_diameter/total_length),
+                   # aspect = max(total_length/midbody_diameter),
                    aspect = mean(total_length/midbody_diameter),
                    max_total_vert = max(total_vert),
                    max_preclo = max(precloacal_vert),
                    max_svl = max(svl),
                    max_tbl = max(total_length))
 
-
-vert_df
-
 vert_data <- as.data.frame(vert_data)
 rownames(vert_data) <- vert_data$species
+
+
+
+### Another way is to use the data from the biggest total length individual in that species
+slice_df <- vert_df %>% dplyr::group_by(species) %>% 
+  dplyr::slice(which.max(total_length)) %>% 
+  dplyr::ungroup()
+  
+slice_df$ratio <- slice_df$total_vert/slice_df$total_length
+slice_df$aspect <- slice_df$total_length/slice_df$midbody_diameter
+
+
+
+
+
 # Calculating vertebrae number to length ratio
 vert_data$ratio <- vert_data$mean_tot_vert/ vert_data$mean_total_length
 
@@ -400,7 +413,6 @@ aspect_ratio <- anilios_vert$aspect; names(aspect_ratio) <- anilios_data$species
 ### Using geomorph::procD.pgls 
 vert.pgls <- geomorph::procD.pgls(vert_ratio ~ aspect_ratio, phy = anilios_tree)
 summary(vert.pgls)
-predict(vert.pgls)
 coefficients(vert.pgls)
 
 dev.off()
@@ -408,11 +420,28 @@ dev.off()
 plot(vert_ratio ~ aspect_ratio, bty="n", 
      # cex = width_ratio_cex[names(width_ratio)], 
      pch = 19,
-     xlab = "Aspect ratio", ylab = "Vertebrae ratio")
-text(x = width_ratio, y = vert_ratio, labels = sp_labs, pos = 1, cex = 0.7)
+     xlab = "Mean aspect ratio", ylab = "Mean vertebrae ratio")
+text(x = aspect_ratio, y = vert_ratio, labels = sp_labs, pos = 1, cex = 0.7)
[…]

### 2024-04-04 manuscript.Rmd
update

@@
 Compared to other snakes, the vertebral column of fossorial snakes tend to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001].
-This is reflected in the relative short tails of blindsnakes (CITE).
+This is reflected in the relatively short tails of blindsnakes (CITE).
 
@@
 
-3. Do vertebral number and body size vary with mean annual temperature?
+3. Does vertebral ratio vary with soil compactness or mean annual temperature?
 
@@
 
-<!-- ## Environmental data  -->
+## Environmental data 
 <!-- NEED TO PUT IN THE CORRECT VALUES HERE -->
+
+We used the Australian blindsnake occurrence records presented in (Tiatragul et al.
+unpublished).
+These records were compiled from museum databases and verified human observations from Atlas of the Living Australia (ALA).
+A total of XXXX records (min = , median = , max = , median per species) representing XX species of Australian *Anilios* for our analyses.
+
 <!-- We used the Australian blindsnake occurrence records presented in (Tiatragul et al.
@@
 (unpublished).
-Then, we test whether the vertebrae to body length ratio is related to maximum aspect ratio, defined by total body length / body width, also from Tiatragul et al.
+Next, we test whether the vertebrae to body length ratio is related to maximum aspect ratio, defined by total body length / body width, also from Tiatragul et al.
 (unpublished) data.
+Finally, we tested if variation in vertebral ratio can be explained by mean annual temperature.
 
@@
 
-### Relationship between pleomerism and body shape
+### Relationship between vertebrae ratio and body shape
 However, we found that vertebrae ratio is postively correlated with aspect ratio (PGLS, slope=-0.0048, y=0.73, R^2^=0.17, F=7.09,, Z=2.2, P=0.011), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig.
@@
 
-<!-- ### Variation in vertebrae number -->
+We found a positive relationship between vertebrae ratio and species mean annual temperature (PGLS, slope=0.038, y=0.19, R^2^=0.21, F=9.58,, Z=2.32, P=0.009), indicating species that have greater amount of vertebrae for body size are generally found in places that are warmer.
 
-Closely related species tend to have similar number of total vertebrae. 
-Males tend to have greater number of postcloacal vertebrae
+<!-- Fig: Pleomerism and ratio correlations -->
+```{r fig-vert-aspect-ratios, fig.cap= '(B) Relationship between mean vertebrae and aspect ratios among Australian blindsnake species.
+Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter.
+Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and wide', out.width="165mm", out.height= "126mm"}
 
-<!-- FIGURE 1 -->
-
-There is a statistically significant correlation between vertebrae ratio and fineness of the body.
-Species that have more vertebrae per length (vertebrae ratio > 1) are usually more fine, whereas more robust species have lower number of vertebrae.
+knitr::include_graphics(paste0(WD, '/assets/figures/vert-ratio-to-temperature.jpg'))
+```
 
@@
 
-Temperature seem to have XXX effect.
+Mean annual temperature was a significant predictor of vertebrae ratio. Temperature could be a proxy for several ecological factors such ...
 Common garden experiments for different species of snakes can tell us if incubation temperature and thus rate of development influences number of vertebrae.

### 2024-04-05 code/03_vert_profile.R
update script for model fit and data viz

@@ -15,19 +15,29 @@
 
 # Data --------------------------------------------------------------------
 # Read in the three sets of replicate measurements for a single specimen
+
+# Ligatus
 lig1 <- read.csv("data/x_ray/lig_R00050_ST_1.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
 lig2 <- read.csv("data/x_ray/lig_R00050_ST_2.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
 lig3 <- read.csv("data/x_ray/lig_R00050_ST_3.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
 
-
+# Grypus
 gry1 <- read.csv("data/x_ray/gry_R146673_ST_1.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
 gry2 <- read.csv("data/x_ray/gry_R146673_ST_2.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
 gry3 <- read.csv("data/x_ray/gry_R146673_ST_3.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
 
+# Diversus
+div1 <- read.csv("data/x_ray/div_R51126_ST_1.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+div2 <- read.csv("data/x_ray/div_R51126_ST_2.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+div3 <- read.csv("data/x_ray/div_R51126_ST_3.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+
 # Plot  -------------------------------------------------------------------
 source('code/utility/func_vert_profile_plotter.R')
 
-par(mfrow=c(1,2))
-vert_raw_plotter(lig1, lig2, lig3, "A. ligatus R00050")
-vert_profile_plotter(lig1, lig2, lig3, "A. ligatus R00050", .error = TRUE)
-vert_profile_plotter(gry1, gry2, gry3, "A. grypus R146673", .error = TRUE)
+pdf(file = "output/vertebral_profile.pdf", width = 8.3, height = 11.7)
+par(mfrow=c(3,1))
+# vert_raw_plotter(lig1, lig2, lig3, "A. ligatus R00050")
+vert_profile_plotter(lig1, lig2, lig3, "A. ligatus R00050", .error = F)
[…]

### 2024-04-05 code/03_vertebrae_profile.R

@@ -1,38 +1,0 @@
-# 03_vertebrae_profile.R
-# Putter Tiatragul
-# March 2024
-# Visualise vertebrae length by numnber of vertebrae
-
-# load data ---------------------------------------------------------------
-
-ligatus_1 <- read.csv('data/x_ray/vert_lengths.csv')
-grypus_1 <- read.csv('data/x_ray/vert_length_grypus.csv')
- 
-
-# Processing --------------------------------------------------------------
-
-ligatus_1 <- ligatus_1[c("Label", "Length")]
-ligatus_1$Label <- sub("\\.dcm$", "", ligatus_1$Label)
-ligatus_1$vert_no <- ave(seq_along(ligatus_1$Label), ligatus_1$Label, FUN = seq_along)
-ligatus_1$rel_length <- ligatus_1$Length / mean(ligatus_1$Length)
-
-
-grypus_1 <- grypus_1[c("Label", "Length")]
-grypus_1$Label <- sub("\\.dcm$", "", grypus_1$Label)
-grypus_1$rel_length <- grypus_1$Length / mean(grypus_1$Length)
-grypus_1$vert_no <- ave(seq_along(grypus_1$Label), grypus_1$Label, FUN = seq_along)
-
-lig_1 <- ligatus_1[ligatus_1$Label == unique(ligatus_1$Label)[1], ]
-gry_1 <- grypus_1[grypus_1$Label == unique(grypus_1$Label)[1], ]
-
-
-
-# Plotting ----------------------------------------------------------------
-par(mfrow=c(1,2))
-
-plot(rel_length ~ vert_no, data = lig_1, pch = 20, xlab = "Vertebrae no", ylab = "Rel. vert. length", bty = "n",
-     ylim = c(0, 1.6))
-
-plot(rel_length ~ vert_no, data = gry_1, pch = 20, xlab = "Vertebrae no", ylab = "Rel. vert. length", bty = "n",
-     ylim = c(0, 1.6))
-

### 2024-04-05 code/03_visualising_data.R

@@ -8,7 +8,7 @@
 # Load packages with two lines
 
 library(ggplot2)
-# library(dplyr)
+library(dplyr)
 library(phytools); 
 library(geiger);
 library(ape); 
@@ -64,7 +64,6 @@ vert_df$vert_ratio <- vert_df$total_vert / vert_df$total_length
 plot(x = vert_df$total_length, y = vert_df$total_vert, xlab = "Total vertebrae #", ylab = "Total length (mm)", bty = "n")
 
 # No clear pattern of vertebrae and total length, indicates there may be some species that elongate their vertebrae and others that shorten their vertebrae relative to the mean. 
-
 
 # Total number of vertebrae -----------------------------------------------
 
@@ -277,14 +276,14 @@ spp<-factor(names(aspect_v),untangle(ladderize(sub_phy),"read.tree")$tip.label)
 
 ## This is using old code where we can't adjust the ylim
 
-pdf(file = "output/tree_boxplot_aspectRatio.pdf", width = 11.7, height = 8.3, paper = "a4r")
+# Figure 1: phylo + boxplots 
+
+pdf(file = "output/tree_boxplot_aspectRatio.pdf", width = 10, height = 7.5)
 plotTree.boxplot(sub_phy,x=aspect_v~spp,
-                 args.boxplot = list(xlab="Aspect ratio (total length/body width",
+                 args.boxplot = list(xlab="Aspect ratio (total length/body width)",
                                      # ylim=c(20,170),
                                      col="grey99"))
 dev.off()
-
-
 
 plotTree.barplot(sub_phy,setNames(vert_data$aspect, rownames(vert_data)),
                  args.boxplot = list(xlab="Aspect ratio (total length/body width"))
@@ -383,23 +382,26 @@ coefficients(pleomerism.pgls)
 
 sp_labs <- gsub(pattern = "Anilios_", replacement = "", x = anilios_data$species)
 
+# Log max body length against log max number of vertebrae
+
+pdf(file = "output/tbl-vert.pdf", width = 10, height = 7.5)
 plot(max_tbl~max_vert, bty="n", pch = 19,
-     ylab = "Max. total body length (mm)", xlab = "Maximum total vertebrae",
[…]

### 2024-04-05 code/utility/func_vert_profile_plotter.R
change axis labels

@@ -84,7 +84,7 @@ vert_profile_plotter <- function(.data1, .data2, .data3, .title = NULL, .error =
   predicted_points <- predict(polynomial_fit, newdata = data.frame(x_values))
   
   # Plot the profile
-  plot(relative_size, xlab = "Vertebra Number (index)", ylab = "Relative Vertebra Size")
+  plot(relative_size, xlab = "Vertebra number", ylab = "Relative vertebra size", ylim=c(0,0.01), xlim=c(0, 400))
   title(paste(.title)) 
   
   # Add error bars if .error is TRUE

### 2024-04-05 manuscript.Rmd
update rough draft

@@
 unpublished; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993).
-Leptophlopids are generally thought to be more narrow and long compared to typhlopids, which are wider and more robust for a given snout-vent length [@herrelBurrowing2021].
+Leptophlopids are generally thought to be more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@herrelBurrowing2021].
 These differences in body size and proportions determine how much push force the snakes can generate when burrowing.
@@
 In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes.
-The specific questions we addressed are:
-
-1. How prevalent is pleomerism among blindsnakes?
-
-2. Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes?
-
-3. Does vertebral ratio vary with soil compactness or mean annual temperature?
+The specific questions we addressed are: (1) How prevalent is pleomerism among blindsnakes?
+(2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes?
+(3) Does vertebral ratio vary with soil compactness or mean annual temperature?
 
@@
 Where possible, we chose relatively large non-fragile specimens from both males and females with well-preserved heads for X-ray.
-Sex for each specimen were determined by inspecting the gonads, or assigned based on relative tail lengths folowing (Tiatragul et al.
+Sex for each specimen were determined by inspecting the gonads or assigned based on relative tail lengths following (Tiatragul et al.
 unpublished).
@@
 Divergence dating was estimated using MCMCTree from the PAML package [@yangPAML2007] based on 27 squamate reptile fossil calibrations [@tiatragulAustralia2023].
-This phylogeny is was used for all phylogenetic comparative analyses.
+This phylogeny was used for all phylogenetic comparative analyses.
 
@@
 Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023].
-Additionally, we quantified intracolumnar length of each vertebrae for three species representing narrow-and-wide phenotypes *A.
+Additionally, we quantified intracolumnar length of each vertebra for three species representing narrow-and-wide phenotypes *A.
 grypus*, wide-and-robust phenotypes (*A.
@@
 Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra.
-We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R.
+We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R v.4.3.1 [@rcoreteamLanguage2023].
 Measurements were repeated three times to account for measurement errors.
@@
 We used the Australian blindsnake occurrence records presented in (Tiatragul et al.
-unpublished).
+unpublished) including a total of 3898 verified records (min=1, median=23, mean=81, max=525 per species).
 These records were compiled from museum databases and verified human observations from Atlas of the Living Australia (ALA).
-A total of XXXX records (min = , median = , max = , median per species) representing XX species of Australian *Anilios* for our analyses.
 
-<!-- We used the Australian blindsnake occurrence records presented in (Tiatragul et al.
-unpublished).
-These records were compiled from museum databases and verified human observations from Atlas of the Living Australia (ALA).
-A total of XXXX records (min = , median = , max = , median per species) representing XX species of Australian *Anilios* for our analyses.
+We selected soil bulk density and annual mean temperature, as we are interested in the potential effects these may have on the distribution of species.
+Soil bulk density measures the penetrability of soil, which is a proxy for the type of microhabitat each species are most likely living in as they are all strictly fossorial.
+Bulk density values > 1.6 g/cm<sup>3</sup> indicate soil that restrict penetration by plant roots, while values <0.5 indicate looser, more organic soil [@mckenzieSoil2002].
+Mean annual temperature is a climate proxy and is commonly used to describe niche for squamate reptiles [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013].
 
-We selected soil bulk density, annual mean temperature, and aridity index as the most biologically relevant data to describe blindsnake niches. 
-Soil bulk density measures the penetrability of soil, which is relevant for strictly fossorial reptiles [camaitiMacroecological2022].
-Bulk density is calculated by taking the weight of dry soil divided by the total soil volume (g/cm<sup>3</sup>).
-Values > 1.6 g/cm<sup>3</sup> tend to restrict root growth, while loose soil have densities < 0.5 g/cm<sup>3</sup> [@mckenzieSoil2002].
-Mean annual temperature serves as a climate proxy and is commonly used to describe niche for squamate reptiles [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013].
-Finally, aridity index reflects precipitation availability over atmospheric water demand and is categorised as follows: hyper arid, < 0.03; arid, 0.03 – 0.2; semi-arid, 0.2-0.5; dry sub-humid, 0.5 – 0.65; humid, >0.65 [@zomerVersion2022].
-From raster files that contained modelled approximations at the resolution of 3 arc-seconds, we extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html), mean annual temperature from WorldClim v.2 [@fickWorldClim2017], and aridity index from @zomerVersion2022 for each georeferenced record using the "extract" function in the R package `raster` v.3.6-3 [@hijmansRaster2022].
-We then calculated the maximum soil bulk density, mean annual temperature, and mean aridity index for each species.
-Correlation analyses revealed that none of the background environmental variables were strongly correlated (Pearson's *r* < |0.75|).
- -->
+From raster files that contained modelled approximations at the resolution of 3 arc-seconds, we extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html), mean annual temperature from WorldClim v.2 [@fickWorldClim2017] for each georeferenced record using the "extract" function in the R package `raster` v.3.6-3 [@hijmansRaster2022].
+We then calculated the species maximum soil bulk density and species mean annual temperature.
 
@@
 
-All statistical analyses were performed in R v.4.31 [@rcoreteamLanguage2023].
-To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Browninan motion implemented in `geomorph` v.4.0.6 [@adamsGeomorph2023].
+All statistical analyses were performed in R.
+To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in `geomorph` v.4.0.6 [@adamsGeomorph2023].
 First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes.
@@
 tovelli* to 400 in *A.
-grypus* ($$Table 1$$).
+grypus* (**Table S1**).
 
 ```{r fig-variation-totalvert, fig.cap= 'Interspecific variation in total number of vertebrae among Australo-Papuan blindsnake and close relatives.
-The boxplot represent the distribution of the data, with vertical inside representing the median.
-The extent of the box represent the lower (left) and upper (right) quartile, with the whiskers showing 1.5 times the interquartile range.
+The boxplot represents the distribution of the data, with vertical inside representing the median.
+The extent of the box represents the lower (left) and upper (right) quartile, with the whiskers showing 1.5 times the interquartile range.
 Outliers beyond this range are depicted as individual points.', out.width="165mm", out.height= "126mm"}
@@
 
-For the Australian blindsnakes, we found a weakly positive but non statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, y=256, R^2^=0.02, F=0.72, Z=0.29, P=0.42), indicating an absence of pleomerism in Australian blindsnakes (Fig.
-\@ref(fig: fig-tbl-vert)).
-This suggest that maximum length is acheived by elongating each vertebrae rather than by adding additional vertebrae
+For the Australian blindsnakes, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, y=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig.
+\@ref(fig:fig-tbl-vert)).
+This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae.
 
 ### Relationship between vertebrae ratio and body shape
-However, we found that vertebrae ratio is postively correlated with aspect ratio (PGLS, slope=-0.0048, y=0.73, R^2^=0.17, F=7.09,, Z=2.2, P=0.011), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig.
-\@ref(fig:fig-vert-width-ratios)).
+However, we found that vertebrae ratio is positively correlated with aspect ratio (PGLS, slope=-0.0048, y=0.73, R^2^=0.17, F=7.09, Z=2.2, P=0.011), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig.
+\@ref(fig:fig-vert-aspect-ratios) and \@ref(fig:fig-vertebral-profile)).
 This is consistent with colubrids and viperids, where species that are more elongated tend to have greater number of vertebrae than stouter ones [@lindellEvolution1994].
@@
 <!-- Fig: Pleomerism and ratio correlations -->
-```{r fig-vert-aspect-ratios, fig.cap= '(B) Relationship between mean vertebrae and aspect ratios among Australian blindsnake species.
+```{r fig-vert-aspect-ratios, fig.cap= 'Relationship between mean vertebrae and aspect ratios among Australian blindsnake species.
 Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter.
-Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and wide', out.width="165mm", out.height= "126mm"}
+Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.', out.width="165mm", out.height= "180mm"}
 
@@
 
+<!-- Fig: Vertebral profile -->
+```{r fig-vertebral-profile, fig.cap= 'Relative vertebral length plotted against vertebra number for three example species with fitted 4° polynomial curves overlaid.', out.width="126mm", out.height= "160mm"}
+
+knitr::include_graphics(paste0(WD, '/assets/figures/vertebral-profile.jpg'))
+```
+
 ### Correlation with mean annual temperature among Australian blindsnakes
 
-We found a positive relationship between vertebrae ratio and species mean annual temperature (PGLS, slope=0.038, y=0.19, R^2^=0.21, F=9.58,, Z=2.32, P=0.009), indicating species that have greater amount of vertebrae for body size are generally found in places that are warmer.
+We found a positive relationship between vertebrae ratio and species mean annual temperature (PGLS, slope=0.038, y=0.19, R^2^=0.21, F=9.58, Z=2.32, P=0.009), indicating species that have greater number of vertebrae for body size are generally found in places that are warmer (Fig.
+\@ref(fig:fig-vert-ecology)A).
+We also found a significant negative relationship between vertebrae ratio and species maximum soil bulk density (PGLS, slope=-1.00, y=2.52, R^2^=0.12, F=4.8, Z=1.7, P=0.043), that have greater amount of vertebrae for body size are generally found in places that have harder to penetrate soil (Fig.
+\@ref(fig:fig-vert-ecology)B).
 
-<!-- Fig: Pleomerism and ratio correlations -->
-```{r fig-vert-aspect-ratios, fig.cap= '(B) Relationship between mean vertebrae and aspect ratios among Australian blindsnake species.
-Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter.
-Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and wide', out.width="165mm", out.height= "126mm"}
+<!-- Fig: Ratio and temperature -->
+```{r fig-vert-ecology, fig.cap= 'Relationship between mean vertebrae ratios and ecological variables.
+(A) Mean vertebrae against species mean annual temperature.
+(B) Mean vertebrae ratio against species max soil bulk density.
+Mean vertebrae ratio is calculated by mean total vertebrae / mean total length for each species.
+Species mean annual temperature and max soil bulk density values are calculated based on occurrence data.
+Trend line represents slope from PGLS', out.width="165mm", out.height= "126mm"}
 
-knitr::include_graphics(paste0(WD, '/assets/figures/vert-ratio-to-temperature.jpg'))
+knitr::include_graphics(paste0(WD, '/assets/figures/vert-ecology.jpg'))
 ```
@@
 Narrow-and-wide phenotypes are common among Leptotyphlopidae, which are thought to use existing burrows rather than dig their own burrow [@herrelBurrowing2021].
-The Australian group is quite exceptional because within the same genus, their is great variation in aspect ratio so species that evolved more narrow-and-long bodies may be converging on the lifestyle with other blindsnakes that have this narrow-and-long phenotype.
+The Australian group is quite exceptional because within the same genus, there is great variation in aspect ratio so species that evolved more narrow-and-long bodies may be converging on the lifestyle with other blindsnakes that have this narrow-and-long phenotype.
 On the other hand, large robust-and-wide species are expected to be able to generate much more force and make their own burrows.
@@
 unpublished) found that species that robust-and-wide species tend to occupy areas in Australia that have easier to penetrate soil compared to narrow-and-long species that tend to occupy areas with more difficult to penetrate soil.
-However, more detailed microhabitat observations is required to determine if body shape is related to the environment in which they burrow.
+However, more detailed microhabitat observations are required to determine if body shape is related to the environment in which they burrow.
 
@@
 Blindsnakes are generally small enough to fit in an X-ray scanner so more people should do it.
+
+\newpage
+
+## References

### 2024-04-08 manuscript.Rmd
update dimensions of figures

@@
 ```{r fig-tbl-vert, fig.cap= 'Maximum vertebral counts plotted against maximum total body length for Australian typhlopids.
-Trend line represent PGLS regression for Australian species.', out.width="165mm", out.height= "126mm"}
+Trend line represent PGLS regression for Australian species.', out.width="165mm", out.height= "124mm"}
 
@@
 Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter.
-Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.', out.width="165mm", out.height= "180mm"}
+Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.', out.width="165mm", out.height= "124mm"}
 
@@
 <!-- Fig: Vertebral profile -->
-```{r fig-vertebral-profile, fig.cap= 'Relative vertebral length plotted against vertebra number for three example species with fitted 4° polynomial curves overlaid.', out.width="126mm", out.height= "160mm"}
+```{r fig-vertebral-profile, fig.cap= 'Relative vertebral length plotted against vertebra number for three example species with fitted 4° polynomial curves overlaid.', out.width="140mm", out.height= "197mm"}
 
@@
 We found that this is (not) the case in blindsnakes.
-Blindsnakes are generally small enough to fit in an X-ray scanner so more people should do it.
+Blindsnakes are generally small enough to fit in an X-ray scanner, so more people should do it.
 

### 2024-04-08 code/03_visualising_data.R
plot with total number of vertebrae across phy

@@ -266,6 +266,7 @@ plotTree.barplot(sub_phy, setNames(vert_data$ratio, rownames(vert_data)),
                  args.barplot=list(xlab="Vertebrae/Length ratio"))
 
 
+
 # Aspect ratio as boxplot
 aspect_v <- setNames(vert_df$total_length / vert_df$midbody_diameter, nm = vert_df$species)
 aspect_v <- aspect_v[which(names(aspect_v) %in% sub_phy$tip.label)]
@@ -284,6 +285,18 @@ plotTree.boxplot(sub_phy,x=aspect_v~spp,
                                      # ylim=c(20,170),
                                      col="grey99"))
 dev.off()
+
+
+
+# Fig vertebrae number boxplot
+vert_num <- setNames(vert_df$total_vert, nm = vert_df$species)
+vert_num <- vert_num[which(names(vert_num) %in% sub_phy$tip.label)]
+plotTree.boxplot(sub_phy,x=vert_num~spp,
+                 args.boxplot = list(xlab="Total number of vertebrae",
+                                     # ylim=c(20,170),
+                                     col="grey99"))
+
+
 
 plotTree.barplot(sub_phy,setNames(vert_data$aspect, rownames(vert_data)),
                  args.boxplot = list(xlab="Aspect ratio (total length/body width"))

### 2024-04-09 code/03_visualising_data.R
combo plot max number of vertebrae

@@ -252,10 +252,6 @@ slice_df <- vert_df %>% dplyr::group_by(species) %>%
 slice_df$ratio <- slice_df$total_vert/slice_df$total_length
 slice_df$aspect <- slice_df$total_length/slice_df$midbody_diameter
 
-
-
-
-
 # Calculating vertebrae number to length ratio
 vert_data$ratio <- vert_data$mean_tot_vert/ vert_data$mean_total_length
 
@@ -287,7 +283,6 @@ plotTree.boxplot(sub_phy,x=aspect_v~spp,
 dev.off()
 
 
-
 # Fig vertebrae number boxplot
 vert_num <- setNames(vert_df$total_vert, nm = vert_df$species)
 vert_num <- vert_num[which(names(vert_num) %in% sub_phy$tip.label)]
@@ -295,8 +290,6 @@ plotTree.boxplot(sub_phy,x=vert_num~spp,
                  args.boxplot = list(xlab="Total number of vertebrae",
                                      # ylim=c(20,170),
                                      col="grey99"))
-
-
 
 plotTree.barplot(sub_phy,setNames(vert_data$aspect, rownames(vert_data)),
                  args.boxplot = list(xlab="Aspect ratio (total length/body width"))
@@ -445,6 +438,35 @@ text(x = aspect_ratio, y = vert_ratio, labels = sp_labs, pos = 1, cex = 0.7)
 abline(a = coefficients(vert.pgls)[1], b = coefficients(vert.pgls)[2])
 dev.off()
 
+
+# Mean aspect ratio (y) against max. number of vertebrae
+ratio_maxvert.pgls <- geomorph::procD.pgls(max_vert ~ aspect_ratio, phy = anilios_tree)
+summary(ratio_maxvert.pgls)
+coefficients(ratio_maxvert.pgls)
+
+plot(max_vert ~ aspect_ratio, bty="n", pch = 19, 
+     ylab = "Max number of vertebrae", xlab = "Mean aspect ratio")
+text(y = max_vert, x = aspect_ratio, labels = sp_labs, pos = 1, cex = 0.7)
+abline(a = coefficients(ratio_maxvert.pgls)[1], b = coefficients(ratio_maxvert.pgls)[2])
+
+
+pdf(file = "output/vert-aspect.pdf", width = 10, height = 7.5)
+par(mfrow=c(2,1))
+plot(vert_ratio ~ aspect_ratio, bty="n", 
[…]

### 2024-04-12 manuscript.Rmd
outline intro, update results, outline discussions

@@
 ---
-title: ""
-date: "`r Sys.Date()`"
+title: "Dissociation between vertebrae number and body size in Australian blindsnakes"
+date: "Version `r Sys.Date()`"
 author: ""
@@
 
-**Title** TBD
+<!-- **Title** TBD -->
 
@@
 
-#Abstract
+# Abstract
+
+Keywords: Scolecophidia, vertebrae, pleomerism, snakes,
 
@@
 
-Blindsnakes are
+***[General background]***
+
+* Introduce the concept of pleomerism? Association between body size and number of segments in vertebrae [@lindseyPleomerism1975].
+* Introduce work on other vertebrates [@headDissociation2007;@pollyTesting2001;@headEvolution2015]
+* In marine snakes [@sherrattVertebral2019;@sherrattMacroevolution2022]
+
+***[Specific background]***
 
 Scolecophidians comprise XXX species of snakes with highly specialised phenotypes for a strictly fossorial life-style [@mirallesMolecular2018].
-
 Although all blindsnakes are fossorial, body size and shape vary significantly among species (Tiatragul et al.
@@
 These differences in body size and proportions determine how much push force the snakes can generate when burrowing.
-
 Their morphological differences led to the idea that leptyphlopids generally use existing burrows while typhlopids create their own burrow.
@@
 
+* Other relevant specific background related to the angle of the paper
+
+***[Knowledge gap]***
+
+A big review included number of vertebrae [@szyndlarIllustrated2023] for many non-aleth
+
 The Australian blindsnake radiation comprise 47 described species [@tiatragulAustralia2023] with considerably diverse morphology from narrow-and-long (aspect ratio >= 100) to robust-and-wide (aspect ratio < 60).
+However, information about the relationship between body size and vertebrae number is unknown.
+
+***[What we do in this paper]***
 
 In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes.
-The specific questions we addressed are: (1) How prevalent is pleomerism among blindsnakes?
+The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes?
 (2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes?
@@
 
-To answer these questions we gathered data on the number of vertebrae across XX species of Australian blindsnakes.
+<!-- To answer these questions we gathered data on the number of vertebrae across XX species of Australian blindsnakes.  -->
 
@@
 (unpublished).
-Next, we test whether the vertebrae to body length ratio is related to maximum aspect ratio, defined by total body length / body width, also from Tiatragul et al.
+Next, we test whether the vertebrae to body length ratio is related to the mean vertebrae number and mean aspect ratio, defined by total body length / body width, also from Tiatragul et al.
 (unpublished) data.
-Finally, we tested if variation in vertebral ratio can be explained by mean annual temperature.
+Finally, we tested if variation in vertebral ratio and vertebrae number can be explained by mean annual temperature and soil compactness.
 
@@
 Total number of vertebrae varies among the 39 species of Australo-Papuan blindsnakes in this study (Fig.
-\@ref(fig:fig-variation-totalvert)), ranging from 146 in *A.
+\@ref(fig:fig-tree-boxplots)), ranging from 146 in *A.
 tovelli* to 400 in *A.
@@
 
-```{r fig-variation-totalvert, fig.cap= 'Interspecific variation in total number of vertebrae among Australo-Papuan blindsnake and close relatives.
-The boxplot represents the distribution of the data, with vertical inside representing the median.
+```{r fig-tree-boxplots, fig.cap= 'Interspecific variation in aspect ratio and total number of vertebrae among Australo-Papuan blindsnake and close relatives.
+The boxplots represent the distribution of data, with vertical inside representing the median.
 The extent of the box represents the lower (left) and upper (right) quartile, with the whiskers showing 1.5 times the interquartile range.
-Outliers beyond this range are depicted as individual points.', out.width="165mm", out.height= "126mm"}
+Outliers beyond this range are depicted as individual points.
+Some species are represented by only the mean.', out.width="165mm", out.height= "92mm"}
 
-knitr::include_graphics(paste0(WD, '/assets/figures/tree-boxplot-totalVert.jpg'))
+knitr::include_graphics(paste0(WD, '/assets/figures/tree-boxplots.jpg'))
 ```
@@
 
-### Relationship between vertebrae ratio and body shape
-However, we found that vertebrae ratio is positively correlated with aspect ratio (PGLS, slope=-0.0048, y=0.73, R^2^=0.17, F=7.09, Z=2.2, P=0.011), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig.
+### Relationship between vertebrae ratio, number, and body shape
+However, we found that vertebrae ratio is positively correlated with aspect ratio (PGLS, slope=-0.0048, y=0.73, R^2^=0.17, F=7.09, Z=2.20, P=0.011), indicating that species that are more narrow-and-long usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig.
 \@ref(fig:fig-vert-aspect-ratios) and \@ref(fig:fig-vertebral-profile)).
-This is consistent with colubrids and viperids, where species that are more elongated tend to have greater number of vertebrae than stouter ones [@lindellEvolution1994].
+We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope=-1.8, y=120.95, R^2^=0.69, F=79.39, Z=5.30, P<0.001).
 
@@
 <!-- Fig: Pleomerism and ratio correlations -->
-```{r fig-vert-aspect-ratios, fig.cap= 'Relationship between mean vertebrae and aspect ratios among Australian blindsnake species.
+```{r fig-vert-aspect-ratios, fig.cap= 'Species mean aspect ratios against (A) mean vertebrae ratio and (B) mean number of vertebrae of Australian blindsnakes.
 Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter.
-Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.', out.width="165mm", out.height= "124mm"}
+Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.', out.width="124mm", out.height= "165mm"}
 
@@
 
-### Correlation with mean annual temperature among Australian blindsnakes
+### Correlation between vertebrae ratio, number, and ecological variables among Australian blindsnakes
 
@@
 \@ref(fig:fig-vert-ecology)A).
-We also found a significant negative relationship between vertebrae ratio and species maximum soil bulk density (PGLS, slope=-1.00, y=2.52, R^2^=0.12, F=4.8, Z=1.7, P=0.043), that have greater amount of vertebrae for body size are generally found in places that have harder to penetrate soil (Fig.
+We also found a significant negative relationship between vertebrae ratio and species maximum soil bulk density (PGLS, slope=-1.00, y=2.52, R^2^=0.12, F=4.8, Z=1.7, P=0.043), indicating species that have greater amount of vertebrae for body size are generally found in places that have harder to penetrate soil (Fig.
 \@ref(fig:fig-vert-ecology)B).
+When considering raw mean vertebrae number, we found non-statistically significant positive correlations with both mean annual temperature (PGLS, slope=4.13, y=142.45, R^2^=0.76, F=2.87, Z=1.31, P=0.09) and maximum soil bulk density (PGLS, slope=111.69, y=, R^2^=0.04, F=1.61, Z=0.85, P=0.22; Fig.
+\@ref(fig:fig-vert-ecology)C–D).
 
-<!-- Fig: Ratio and temperature -->
+<!-- Fig: Ratio and ecological variables -->
 ```{r fig-vert-ecology, fig.cap= 'Relationship between mean vertebrae ratios and ecological variables.
@@
 Species mean annual temperature and max soil bulk density values are calculated based on occurrence data.
-Trend line represents slope from PGLS', out.width="165mm", out.height= "126mm"}
+Trend line represents slope from PGLS', out.width="165mm", out.height= "117mm"}
 
@@
 
-Summary about what the paper found
+ <!-- The specific questions we addressed are: (1) How prevalent is pleomerism among blindsnakes?
+(2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes?
+(3) Does vertebral ratio vary with mean annual temperature or soil compactness?
+-->
 
-Not so surprising but shows that pleomerism does not explain variation in body size in this continental radiation of blindsnakes.
-Only real giant taxa like *Rhinotyphlops* and *Afrotyphlops* tend to break the rule.
-However, when fitted in our PGLS, we found that even they are not exceptional???
+***[Big summary]***
+Results from our study revealed that that pleomerism does not apply to Australian typhlopids, indicating a deviation from the positive correlation between body size and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975].
+Variation in body shape, however, is underpinned by the ratio between vertebrae number to body size.
+We show that species that are narrow-and-long have proportionally more vertebrae compared to species that are more robust-and-wide.
+This relationship between the proportion of vertebrae to body shape has been found in colubrids and viperids, where more elongated species tend to have a greater number of vertebrae than stouter ones [@lindellEvolution1994].
+Variation in vertebral ratio can also be explained by both mean annual temperature and soil compactness.
+These results have important implications for understanding the macroevolutionary dynamics of continental-scale radiation of Australian blindsnakes.
 
-Body shape varied with number of vertebrae in similar ways as previously reported in vipers and colubrids [@lindellEvolution1994].
+***[Discuss why body shape might influence vertebrae ratio or vice-versa]***
+
+Body shape varied with the number of vertebrae in similar ways as previously reported in vipers and colubrids [@lindellEvolution1994]. 
 We can speculate that species that are narrow-and-long may be using pre-existing burrows rather than digging their own burrow.
 Narrow-and-wide phenotypes are common among Leptotyphlopidae, which are thought to use existing burrows rather than dig their own burrow [@herrelBurrowing2021].
-The Australian group is quite exceptional because within the same genus, there is great variation in aspect ratio so species that evolved more narrow-and-long bodies may be converging on the lifestyle with other blindsnakes that have this narrow-and-long phenotype.
-On the other hand, large robust-and-wide species are expected to be able to generate much more force and make their own burrows.
-A previous study (Tiatragul et al.
-unpublished) found that species that robust-and-wide species tend to occupy areas in Australia that have easier to penetrate soil compared to narrow-and-long species that tend to occupy areas with more difficult to penetrate soil.
-However, more detailed microhabitat observations are required to determine if body shape is related to the environment in which they burrow.
+In this regard, the Australian typhlopid genus, *Anilios*, is quite exceptional because there is a great variation in body shape (Tiatragul et al.
+unpublished).
+Species that evolved more narrow-and-long bodies may be converging on the lifestyle with other scolecophidians that have similar phenotypes (i.e.
+leptotyphlopids).
+On the other hand, robust-and-wide phenotypes common among typhlopid species are expected to be able to generate much more force and make their own burrows [@herrelBurrowing2021].
 
-Mean annual temperature was a significant predictor of vertebrae ratio. Temperature could be a proxy for several ecological factors such ...
-Common garden experiments for different species of snakes can tell us if incubation temperature and thus rate of development influences number of vertebrae.
-We would expect higher temperature to accelerate development (CITE anolis papers, turtles, etc), so less time for embryonic development.
+***[Discuss/speculate why vertebrae ratio is correlated with annual mean temperature and soil bulk density]***
+
+Both mean annual temperature and soil bulk density were significant predictors of vertebrae ratio but in different directions.
+Mean annual temperature has been shown to be a proxy for several ecological factors such as rainfall and vegetation (CITE).
+However, temperature has also been shown to directly influence multiple axes of reptile development (cite).
+Vertebral development (somitogenesis) and homeotic (*Hox gene*) regulation may thus be affected by variations in incubation conditions [@martin-del-campoHox2018].
+However, we currently lack knowledge about egg-laying behaviour and how temperature affects the development of blindsnake embryos.
+Common garden experiments across different species of blindsnakes may be able to shed light on whether variation in incubation temperature and consequently, the rate of development may alter the number of vertebrae formation during development.
+<!-- We would expect higher temperature to accelerate development (CITE anolis papers, turtles, etc), so less time for embryonic development.  -->
+
+A previous study on Australian blindsnakes (Tiatragul et al.
+unpublished) found that robust-and-wide species tend to occupy areas in Australia that have easier-to-penetrate soil compared to narrow-and-long species that tend to occupy areas with more difficult-to-penetrate soil.
+A greater number of vertebrae in a narrow-and-long body may afford greater flexibility that enables these species to navigate in pre-existing burrows or termite galleries, typical across northern Australia, without exerting force to dig their own burrow.
+However, more detailed species-level microhabitat observations are required to determine if body shape is related to the environment in which they burrow.
 
@@
 We found that this is (not) the case in blindsnakes.
-Blindsnakes are generally small enough to fit in an X-ray scanner, so more people should do it.
+Blindsnakes are generally small enough to fit in an X-ray scanner, even in portable field X-ray machines.
+Collecting more data on this group of basal snakes can help us understand how vertebral morphology evolved in snakes.
 

### 2024-04-12 code/03_visualising_data.R
additional analyses and plots

@@ -46,6 +46,16 @@ load('data/script_generated_data/subset_mcmctree_shape.rda')
 anilios_tree <- ape::drop.tip(sub_phy, tip = c("Ramphotyphlops_multilineatus","Acutotyphlops_subocularis"))
 anilios_tree$tip.label
 
+dev.off()
+
+b_tree <- sub_phy
+b_tree$tip.label <- gsub(pattern = "Anilios_", replacement = "A. ", b_tree$tip.label)
+b_tree$tip.label <- gsub(pattern = "Acutotyphlops_", replacement = "Acu. ", b_tree$tip.label)
+b_tree$tip.label <- gsub(pattern = "Ramphotyphlops_", replacement = "R. ", b_tree$tip.label)
+
+plotTree(b_tree, ftype="i", lwd = 2, fsize=1, type = "fan")
+
+
 # Summarise linear measurement data ---------------------------------------
 
 genus_species <- paste(linear_df$genus, linear_df$species, sep = "_")
@@ -74,13 +84,6 @@ total_vertebrae <- setNames(vert_df$total_vert, nm = vert_df$species)
 total_vertebrae <- total_vertebrae[which(names(total_vertebrae) %in% sub_phy$tip.label)]
 geiger::name.check(phy = sub_phy, data = total_vertebrae)
 spp<-factor(names(total_vertebrae),untangle(ladderize(sub_phy),"read.tree")$tip.label)
-
-pdf(file = "output/tree_boxplot_totalVert.pdf", width = 11.7, height = 8.3, paper = "a4r")
-plotTree.boxplot(sub_phy,x=total_vertebrae~spp,
-                 args.boxplot = list(xlab="Total number of vertebrae",
-                                     # ylim=c(20,170),
-                                     col="grey99"))
-dev.off()
 
 summary(vert_df$total_vert)
 vert_df[order(vert_df$total_vert), c("species", "total_vert")]
@@ -120,6 +123,9 @@ Vertebra_summary_table_by_sex <- vert_df %>%
 
 Vertebra_summary_table_by_sex
 
+
+
+
 # Q1 Test for sexual dimorphism ----------------------------------------------
 
 ## Question: Which species show sexual dimorphism in vert/length ratio?
@@ -149,7 +155,6 @@ d_vert_ratio
 
[…]

### 2024-06-07 code/03_visualising_data.R
skull length

@@ -124,6 +124,17 @@ Vertebra_summary_table_by_sex <- vert_df %>%
 Vertebra_summary_table_by_sex
 
 
+## Head length x-ray
+skull_summary <- vert_df %>% 
+  dplyr::group_by(species, sex) %>% 
+  dplyr::filter(sex %in% c("f","m")) %>% 
+  summarise(sample_size = n(),
+            skull_length = mean(headlength_xray),
+            rel_skull = mean(headlength_xray/svl))
+
+write.csv(skull_summary, file = "assets/tables/skull_summary.csv", row.names = FALSE)
+
+
 
 
 # Q1 Test for sexual dimorphism ----------------------------------------------
@@ -279,10 +290,10 @@ spp<-factor(names(aspect_v),untangle(ladderize(sub_phy),"read.tree")$tip.label)
 
 # Figure 1: phylo + boxplots 
 
-pdf(file = "output/tree_boxplot_aspectRatio.pdf", width = 10, height = 7.5)
+pdf(file = "output/tree_boxplot_aspectRatio.pdf", width = 20, height = 12)
 plotTree.boxplot(sub_phy,x=aspect_v~spp,
                  args.boxplot = list(xlab="Aspect ratio (total length/body width)",
-                                     # ylim=c(20,170),
+                                     # ylim=c(0,200),
                                      col="grey99"))
 dev.off()
 
@@ -290,7 +301,7 @@ dev.off()
 vert_num <- setNames(vert_df$total_vert, nm = vert_df$species)
 vert_num <- vert_num[which(names(vert_num) %in% sub_phy$tip.label)]
 
-pdf(file = "output/tree_boxplot_totalVert.pdf", width = 10, height = 7.5)
+pdf(file = "output/tree_boxplot_totalVert.pdf", width = 10, height = 12)
 plotTree.boxplot(sub_phy,x=vert_num~spp,
                  args.boxplot = list(xlab="Total number of vertebrae",
                                      # ylim=c(20,170),

### 2024-06-07 manuscript.Rmd
more citations

@@
 ^1^Division of Ecology & Evolution, Research School of Biology, The Australian National University, Canberra 2601, ACT, Australia
+
 ^2^Queensland Museum, Hendra, QLD, 4011, Australia
+
 ^3^School of Biological Sciences, The University of Adelaide, Adelaide, SA 5005, Australia
@@
 <!-- **Author contributions**:  -->
-Putter:
 
@@
 
-***[General background]***
+* Axial evolution an important aspect of the vertebrate body plan evolution – facilitates diversity and influences locomotion 
+  <!-- Note the papers  that are about axial number relating to locomotion -->
+@galisFast2014 - mammals
+@vandammeSpeed2002 - lacertids
+@yamadaAdaptive2009;@forkerBendy2021 - gobies 
+@molnarMorphological2023 - chameleons
 
-* Introduce the concept of pleomerism? Association between body size and number of segments in vertebrae [@lindseyPleomerism1975].
-* Introduce work on other vertebrates [@headDissociation2007;@pollyTesting2001;@headEvolution2015]
-* In marine snakes [@sherrattVertebral2019;@sherrattMacroevolution2022]
+* Need a reference for mammals? Limited variation of a common theme that is constrained [@asherVariability2011;@naritaEvolution2005]
+@liDivergent2023 found vertebral counts to differ significantly between major groups of mammals and within groups. 
+* Reptiles, on the other hand, are more varied [@headEvolution2015]
+
+* Elongate animals have special properties of vertebral number known as pleomerism, where body size and number of segments in vertebrae are correlated [@lindseyPleomerism1975]. 
+Pleomerism has been observed in eels and fish [@wardEvolution2007;@wardDifferential2014;@reeceEvolutionary2013;], in squamates [@pollyTesting2001;@headDissociation2007;@headEvolution2015;@bergmannVertebral2012;@lindellEvolution1994;@cernanskyVertebral2019], in salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015]
+
+* Snakes are a good model to study axial evolution because they have a high number of vertebrae and a wide range of body shapes 
+* In elapid snakes [@sherrattVertebral2019;@sherrattMacroevolution2022;@hamptonInterspecific2019;@hamptonRegionalization2023]; colubrid snakes [@nash-hahnModification2024].
 
@@
 
-Scolecophidians comprise XXX species of snakes with highly specialised phenotypes for a strictly fossorial life-style [@mirallesMolecular2018].
+Scolecophidians comprise XXX species of snakes with highly specialised phenotypes for a strictly fossorial lifestyle [@mirallesMolecular2018;@uetzQuarter2021].
 Although all blindsnakes are fossorial, body size and shape vary significantly among species (Tiatragul et al.
@@
 
-Compared to other snakes, the vertebral column of fossorial snakes tend to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001].
+Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001].
 This is reflected in the relatively short tails of blindsnakes (CITE).
@@
 We selected soil bulk density and annual mean temperature, as we are interested in the potential effects these may have on the distribution of species.
-Soil bulk density measures the penetrability of soil, which is a proxy for the type of microhabitat each species are most likely living in as they are all strictly fossorial.
+Soil bulk density measures the penetrability of soil, which is a proxy for the type of microhabitat species is most likely living in as they are all strictly fossorial.
 Bulk density values > 1.6 g/cm<sup>3</sup> indicate soil that restrict penetration by plant roots, while values <0.5 indicate looser, more organic soil [@mckenzieSoil2002].
@@
 Outliers beyond this range are depicted as individual points.
-Some species are represented by only the mean.', out.width="165mm", out.height= "92mm"}
+Some species are represented by only the mean.', out.width="165mm", out.height= "129mm"}
 
@@
 \@ref(fig:fig-vert-ecology)A).
-We also found a significant negative relationship between vertebrae ratio and species maximum soil bulk density (PGLS, slope=-1.00, y=2.52, R^2^=0.12, F=4.8, Z=1.7, P=0.043), indicating species that have greater amount of vertebrae for body size are generally found in places that have harder to penetrate soil (Fig.
+We also found a significant negative relationship between vertebrae ratio and species maximum soil bulk density (PGLS, slope=-1.00, y=2.52, R^2^=0.12, F=4.8, Z=1.7, P=0.043), indicating species that have greater number of vertebrae for body size are generally found in places that have harder to penetrate soil (Fig.
 \@ref(fig:fig-vert-ecology)B).
@@
 Species mean annual temperature and max soil bulk density values are calculated based on occurrence data.
-Trend line represents slope from PGLS', out.width="165mm", out.height= "117mm"}
+Trend lines represent slopes from PGLS', out.width="165mm", out.height= "117mm"}
 
@@
 ***[Big summary]***
+
 Results from our study revealed that that pleomerism does not apply to Australian typhlopids, indicating a deviation from the positive correlation between body size and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975].

### 2024-06-12 manuscript.Rmd
update intro from ES edit

@@
 
-* Snakes are a good model to study axial evolution because they have a high number of vertebrae and a wide range of body shapes 
+* Snakes are a good model to study axial evolution because they have a high number of vertebrae and a wide range of body size and shape.
 * In elapid snakes [@sherrattVertebral2019;@sherrattMacroevolution2022;@hamptonInterspecific2019;@hamptonRegionalization2023]; colubrid snakes [@nash-hahnModification2024].
@@
 
-Scolecophidians comprise XXX species of snakes with highly specialised phenotypes for a strictly fossorial lifestyle [@mirallesMolecular2018;@uetzQuarter2021].
+Scolecophidians, often referred to in English as "blindsnakes", are characterised by their fossorial lifestyle, poor visual system, and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018; @strongDeconstructing2021; @gowerEyetranscriptome2021].
+Scolecophidians comprise five paraphyletic families, including the "core scolecophidians" — a monophyletic clade consisting of Typhlopidae (425 spp), Leptotyphlopidae (143 spp), Gerrhopilidae (29 spp), and Xenotyphlopidae (1 spp) — and the paraphyletic Anomalepididae (20 spp), which is sister to all other "typical snakes" (Alethinophidia) [@mirallesMolecular2018; @burbrinkInterrogating2020; @uetzQuarter2021].
+
 Although all blindsnakes are fossorial, body size and shape vary significantly among species (Tiatragul et al.
@@
 
-Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001].
-This is reflected in the relatively short tails of blindsnakes (CITE).
-
-* Other relevant specific background related to the angle of the paper
-
 ***[Knowledge gap]***
-
-A big review included number of vertebrae [@szyndlarIllustrated2023] for many non-aleth
 
@@
 However, information about the relationship between body size and vertebrae number is unknown.
+
+Among typhlopidae, gigantism have evolved multiple times but with different mechanisms [@headDissociation2007].
+For instance, @headDissociation2007 found that some of the largest species of blindsnakes in the African continent (e.g.
+*Afrotyphlops schlegelii*) have fewer than expected precloacal vertebrae than expected for their size.
+This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae.
+
+In another predominantly burrowing clade, caecilians, the number of vertebrae is positively correlated with body size [@renousBody1989], suggesting pleomerism is a mechanism for burrowing taxa to diversity in the underground environment.
+
+<!-- A big review included number of vertebrae [@szyndlarIllustrated2023] for many non-alethinophidian snakes, but  -->
+
+Compared to other snakes, the vertebral column of fossorial snakes tend to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001].
+This is reflected in the relatively short tails of blindsnakes (CITE).
 
@@
 
-Results from our study revealed that that pleomerism does not apply to Australian typhlopids, indicating a deviation from the positive correlation between body size and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975].
+Results from our study revealed that pleomerism does not apply to Australian typhlopids, indicating a deviation from the positive correlation between body size and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975].
 Variation in body shape, however, is underpinned by the ratio between vertebrae number to body size.

### 2024-06-14 manuscript.Rmd
make changes from ES comments

@@
 We examined 341 specimens representing 44 lineages *Anilios* blindsnakes including 42/47 formally described species [@uetzQuarter2021].
+We additionally included two close relatives, *Ramphotyphlops multilineatus* and *Acutotyphlops subocularis* [@tiatragulAustralia2023].
 Our sample size per species ranged from 1 to 25 specimens (median=7 specimens).
@@
 Where possible, we chose relatively large non-fragile specimens from both males and females with well-preserved heads for X-ray.
-Sex for each specimen were determined by inspecting the gonads or assigned based on relative tail lengths following (Tiatragul et al.
-unpublished).
+Sex for each specimen were determined by inspecting the gonads or assigned based on relative tail lengths following Tiatragul et al.
+(unpublished).
 
@@
 
-We used the Fiji distribution of ImageJ2 [@schindelinFiji2012] to digitally visualise X-ray images and counted the number of vertebrae using Cell Counter module.
+We used the Fiji distribution of ImageJ2 v.1.54j [@schindelinFiji2012;@ruedenImageJ22017] to digitally visualise X-ray images and counted the number of vertebrae using Cell Counter module.
 Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023].
-Additionally, we quantified intracolumnar length of each vertebra for three species representing narrow-and-wide phenotypes *A.
-grypus*, wide-and-robust phenotypes (*A.
-pinguis*), and intermediate phenotypes (*A.
+We calculated vertebrae ratio by dividing the total number of vertebrae by total body length.
+
+Additionally, we quantified intracolumnar length of each vertebra for three species representing 1) narrow-and-wide phenotypes (*A.
+grypus*), 2) wide-and-robust phenotypes (*A.
+pinguis*), and 3) intermediate phenotypes (*A.
 centralis*).
 Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra.
-We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R v.4.3.1 [@rcoreteamLanguage2023].
+We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R Statistical Environment v.4.3.1 [@rcoreteamLanguage2023].
 Measurements were repeated three times to account for measurement errors.
@@
 
-We used the Australian blindsnake occurrence records presented in (Tiatragul et al.
-unpublished) including a total of 3898 verified records (min=1, median=23, mean=81, max=525 per species).
+We used the Australian blindsnake occurrence records presented in Tiatragul et al.
+(unpublished) including a total of 3898 verified records (min=1, median=23, mean=81, max=525 per species).
 These records were compiled from museum databases and verified human observations from Atlas of the Living Australia (ALA).
@@
 
-From raster files that contained modelled approximations at the resolution of 3 arc-seconds, we extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html), mean annual temperature from WorldClim v.2 [@fickWorldClim2017] for each georeferenced record using the "extract" function in the R package `raster` v.3.6-3 [@hijmansRaster2022].
+From raster files that contained modelled approximations at the resolution of 3 arc-seconds, we extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html), mean annual temperature from WorldClim v.2 [@fickWorldClim2017] for each georeferenced record using the "extract" function in the R package *`raster`* v.3.6-3 [@hijmansRaster2022].
 We then calculated the species maximum soil bulk density and species mean annual temperature.
@@
 
-All statistical analyses were performed in R.
-To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in `geomorph` v.4.0.6 [@adamsGeomorph2023].
+All statistical analyses were performed in R Statistical Environment.
+To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in the R package *`geomorph`* v.4.0.6 [@adamsGeomorph2023].
 First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes.
@@
 (unpublished) data.
-Finally, we tested if variation in vertebral ratio and vertebrae number can be explained by mean annual temperature and soil compactness.
+Finally, we tested if variation in vertebrae ratio and vertebrae number can be explained by mean annual temperature and soil compactness.
+In all PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations.
+
+To visualise the relationship between vertebrae number and intracolumnar vertebral length, for each of the representative species, we plotted the mean vertebrae size (from three repeated measurements) against the vertebra number.
+Each vertebrae profile was overlaid with a fitted 4° polynomial curve.
 
@@
 grypus* (**Table S1**).
+Aspect ratios ranged from 20.8 in the robust-and-wide *A.
+pinguis* to 169.8 in the narrow-and-long *A.
+longissimus* (**Table S2**).
 
@@
 
-For the Australian blindsnakes, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, y=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig.
+For the Australian blindsnakes, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig.
 \@ref(fig:fig-tbl-vert)).
-This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae.
+This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae, while most species have many more vertebrae than expected for their size.
+For example, *A.
+grypus* "West" is an intermediate length (relative to other species) and has the most vertebrae, while *A.
+torresianus*, which is the longest, has relatively few vertebrae.
 
 ### Relationship between vertebrae ratio, number, and body shape
-However, we found that vertebrae ratio is positively correlated with aspect ratio (PGLS, slope=-0.0048, y=0.73, R^2^=0.17, F=7.09, Z=2.20, P=0.011), indicating that species that are more narrow-and-long usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig.
+We found that vertebrae ratio is positively correlated with aspect ratio (PGLS, slope=-0.0048, intercept=0.73, R^2^=0.17, F=7.09, Z=2.20, P=0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig.
 \@ref(fig:fig-vert-aspect-ratios) and \@ref(fig:fig-vertebral-profile)).
-We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope=-1.8, y=120.95, R^2^=0.69, F=79.39, Z=5.30, P<0.001).
+We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope=-1.8, intercept=120.95, R^2^=0.69, F=79.39, Z=5.30, P<0.001).
 

### 2024-06-14 code/03_visualising_data.R
change to phylomorphospace

@@ -285,17 +285,31 @@ aspect_v <- aspect_v[which(names(aspect_v) %in% sub_phy$tip.label)]
 geiger::name.check(phy = sub_phy, data = aspect_v)
 spp<-factor(names(aspect_v),untangle(ladderize(sub_phy),"read.tree")$tip.label)
 
+## Statistics about aspect ratio
+summary(aspect_v)
+aspect_v[(which.min(aspect_v))]
+aspect_v[(which.max(aspect_v))]
+
+
+table_s2 <- vert_df %>% 
+  dplyr::group_by(species) %>% 
+  dplyr::summarise(mean_aspect = mean(total_length/midbody_diameter),
+                   max_aspect = max(total_length/midbody_diameter),
+                   min_aspect = min(total_length/midbody_diameter),
+                   sample_n = n()) 
+
+write.csv(table_s2, file = "assets/tables/aspect_ratios_summary", row.names = F)
 
 ## This is using old code where we can't adjust the ylim
 
 # Figure 1: phylo + boxplots 
 
-pdf(file = "output/tree_boxplot_aspectRatio.pdf", width = 20, height = 12)
+# pdf(file = "output/tree_boxplot_aspectRatio.pdf", width = 20, height = 12)
 plotTree.boxplot(sub_phy,x=aspect_v~spp,
                  args.boxplot = list(xlab="Aspect ratio (total length/body width)",
                                      # ylim=c(0,200),
                                      col="grey99"))
-dev.off()
+# dev.off()
 
 # Fig vertebrae number boxplot
 vert_num <- setNames(vert_df$total_vert, nm = vert_df$species)
@@ -406,13 +420,28 @@ sp_labs <- gsub(pattern = "Anilios_", replacement = "", x = anilios_data$species
 
 # Log max body length against log max number of vertebrae
 
-pdf(file = "output/tbl-vert.pdf", width = 10, height = 7.5)
-plot(max_tbl~max_vert, bty="n", pch = 19,
-     ylab = "Max. total body length (mm)", xlab = "Maximum number of total vertebrae")
+# pdf(file = "output/tbl-vert.pdf", width = 10, height = 7.5)
+
+anilios_tree
+anilios_vert[,c("max_total_vert", "max_tbl")]
+
+phylomorphospace(anilios_tree,anilios_vert[,c("max_total_vert", "max_tbl")], 
[…]

### 2024-07-17 code/03_visualising_data_old.r
add from laptop

@@ -46,16 +46,6 @@ load('data/script_generated_data/subset_mcmctree_shape.rda')
 anilios_tree <- ape::drop.tip(sub_phy, tip = c("Ramphotyphlops_multilineatus","Acutotyphlops_subocularis"))
 anilios_tree$tip.label
 
-dev.off()
-
-b_tree <- sub_phy
-b_tree$tip.label <- gsub(pattern = "Anilios_", replacement = "A. ", b_tree$tip.label)
-b_tree$tip.label <- gsub(pattern = "Acutotyphlops_", replacement = "Acu. ", b_tree$tip.label)
-b_tree$tip.label <- gsub(pattern = "Ramphotyphlops_", replacement = "R. ", b_tree$tip.label)
-
-plotTree(b_tree, ftype="i", lwd = 2, fsize=1, type = "fan")
-
-
 # Summarise linear measurement data ---------------------------------------
 
 genus_species <- paste(linear_df$genus, linear_df$species, sep = "_")
@@ -84,6 +74,13 @@ total_vertebrae <- setNames(vert_df$total_vert, nm = vert_df$species)
 total_vertebrae <- total_vertebrae[which(names(total_vertebrae) %in% sub_phy$tip.label)]
 geiger::name.check(phy = sub_phy, data = total_vertebrae)
 spp<-factor(names(total_vertebrae),untangle(ladderize(sub_phy),"read.tree")$tip.label)
+
+pdf(file = "output/tree_boxplot_totalVert.pdf", width = 11.7, height = 8.3, paper = "a4r")
+plotTree.boxplot(sub_phy,x=total_vertebrae~spp,
+                 args.boxplot = list(xlab="Total number of vertebrae",
+                                     # ylim=c(20,170),
+                                     col="grey99"))
+dev.off()
 
 summary(vert_df$total_vert)
 vert_df[order(vert_df$total_vert), c("species", "total_vert")]
@@ -123,20 +120,6 @@ Vertebra_summary_table_by_sex <- vert_df %>%
 
 Vertebra_summary_table_by_sex
 
-
-## Head length x-ray
-skull_summary <- vert_df %>% 
-  dplyr::group_by(species, sex) %>% 
-  dplyr::filter(sex %in% c("f","m")) %>% 
-  summarise(sample_size = n(),
-            skull_length = mean(headlength_xray),
-            rel_skull = mean(headlength_xray/svl))
-
[…]

### 2024-07-17 code/03_visualising_data.R
phylomorphorspace for tbl v total vert

@@ -7,12 +7,8 @@
 # libraries ---------------------------------------------------------------
 # Load packages with two lines
 
-library(ggplot2)
-library(dplyr)
-library(phytools); 
-library(geiger);
-library(ape); 
-library(geomorph); 
+library(ggplot2); library(dplyr); library(phytools); 
+library(geiger); library(ape); library(geomorph); 
 # library(geiger)
 # library(R.utils)
 library(phylolm)
@@ -272,6 +268,7 @@ slice_df$aspect <- slice_df$total_length/slice_df$midbody_diameter
 vert_data$ratio <- vert_data$mean_tot_vert/ vert_data$mean_total_length
 
 check_data <- geiger::name.check(phy = sub_phy, data = vert_data)
+vert_data_full <- vert_data
 vert_data <- vert_data[which(rownames(vert_data) %notin% check_data$data_not_tree),]
 
 plotTree.barplot(sub_phy, setNames(vert_data$ratio, rownames(vert_data)), 
@@ -412,41 +409,43 @@ max_vert <- anilios_data$max_vert; names(max_vert) <- anilios_data$species
 max_svl <- anilios_data$max_svl; names(max_svl) <- anilios_data$species
 max_tbl <- setNames(anilios_vert$max_tbl, anilios_vert$species)
 
+colnames(anilios_data)
+colnames(vert_data_full)
+
+### Data for species not in the phylogeny
+not_tree_data <- vert_data_full[which(rownames(vert_data_full) %notin% c(anilios_tree$tip.label, "Anilios_sp")),]
+not_tree_labs <- gsub(pattern = "Anilios_", replacement = "", x = not_tree_data$species)
+
+
+## Test pleomerism for species in tree
 pleomerism.pgls <- geomorph::procD.pgls(max_tbl ~ max_vert, phy = anilios_tree)
 summary(pleomerism.pgls)
 coefficients(pleomerism.pgls)
-
 sp_labs <- gsub(pattern = "Anilios_", replacement = "", x = anilios_data$species)
 
+
+### PLOTTING
 # Log max body length against log max number of vertebrae
 
-# pdf(file = "output/tbl-vert.pdf", width = 10, height = 7.5)
-
-anilios_tree
-anilios_vert[,c("max_total_vert", "max_tbl")]
 
 
+### FIGURE 2: PGLS tbl v vert number
 ### Emma suggest plotting as phylomorphospace and add species that are not on the tree overlaid
[…]

### 2024-07-18 code/03_visualising_data.R
add phylomorphospace for figure 3

@@ -432,7 +432,7 @@ sp_labs <- gsub(pattern = "Anilios_", replacement = "", x = anilios_data$species
 ### FIGURE 2: PGLS tbl v vert number
 ### Emma suggest plotting as phylomorphospace and add species that are not on the tree overlaid
 
-pdf(file = "output/tbl-vert.pdf", width = 10, height = 7.5)
+pdf(file = "output/phylomorpho-tbl-vert.pdf", width = 10, height = 7.5)
 
 phylomorphospace(anilios_tree, anilios_vert[,c("max_total_vert", "max_tbl")], 
                  bty="n", label="horizontal", node.size=c(0.4,1.3),
@@ -496,41 +496,43 @@ plot(max_vert ~ aspect_ratio, bty="n", pch = 19,
 text(y = max_vert, x = aspect_ratio, labels = sp_labs, pos = 1, cex = 0.7)
 abline(a = coefficients(ratio_maxvert.pgls)[1], b = coefficients(ratio_maxvert.pgls)[2])
 
-
 # Mean aspect ratio (y) against mean number of vertebrae
 ratio_meanvert.pgls <- geomorph::procD.pgls(mean_vert ~ aspect_ratio, phy = anilios_tree)
 summary(ratio_meanvert.pgls)
 coefficients(ratio_meanvert.pgls)
 
-plot(mean_vert ~ aspect_ratio, bty="n", pch = 19, 
-     ylab = "Mean number of vertebrae", xlab = "Mean aspect ratio")
-text(y = mean_vert, x = aspect_ratio, labels = sp_labs, pos = 1, cex = 0.7)
-abline(a = coefficients(ratio_meanvert.pgls)[1], b = coefficients(ratio_meanvert.pgls)[2])
-
-
-## Combine plots
-pdf(file = "output/vert-aspect.pdf", width = 9, height = 12)
+### Figure 3: COMBINED Phylomorphosapce 
+pdf(file = "output/phylomorpho-vert-aspect-total.pdf", width = 9, height = 12)
 par(mfrow=c(2,1))
-plot(vert_ratio ~ aspect_ratio, bty="n",
-     pch = 19, xlab = "Mean aspect ratio", ylab = "Mean vertebrae ratio",
-     ylim=c(0.4,1.8))
-text(x = aspect_ratio, y = vert_ratio, labels = sp_labs, pos = 1, cex = 0.7)
[…]

### 2024-07-18 manuscript.Rmd
update before sending to Emma on 2024-07-18

@@
   <!-- Note the papers  that are about axial number relating to locomotion -->
-@galisFast2014 - mammals
-@vandammeSpeed2002 - lacertids
-@yamadaAdaptive2009;@forkerBendy2021 - gobies 
-@molnarMorphological2023 - chameleons
+in mammals [@galisFast2014 ], lacertids [@vandammeSpeed2002], gobies [@yamadaAdaptive2009;@forkerBendy2021], and chameloens [@molnarMorphological2023].
 
@@
 * Elongate animals have special properties of vertebral number known as pleomerism, where body size and number of segments in vertebrae are correlated [@lindseyPleomerism1975]. 
-Pleomerism has been observed in eels and fish [@wardEvolution2007;@wardDifferential2014;@reeceEvolutionary2013;], in squamates [@pollyTesting2001;@headDissociation2007;@headEvolution2015;@bergmannVertebral2012;@lindellEvolution1994;@cernanskyVertebral2019], in salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015]
+Pleomerism has been observed in eels and fish [@wardEvolution2007;@wardDifferential2014;@reeceEvolutionary2013], in squamates [@pollyTesting2001;@headDissociation2007;@headEvolution2015;@bergmannVertebral2012;@lindellEvolution1994;@cernanskyVertebral2019], in salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].
 
@@
 
-Scolecophidians, often referred to in English as "blindsnakes", are characterised by their fossorial lifestyle, poor visual system, and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018; @strongDeconstructing2021; @gowerEyetranscriptome2021].
+Scolecophidians, often referred to in English as "blindsnakes", are characterised by their fossorial lifestyle, poor visual system, and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018;@strongDeconstructing2021;@gowerEyetranscriptome2021].
 Scolecophidians comprise five paraphyletic families, including the "core scolecophidians" — a monophyletic clade consisting of Typhlopidae (425 spp), Leptotyphlopidae (143 spp), Gerrhopilidae (29 spp), and Xenotyphlopidae (1 spp) — and the paraphyletic Anomalepididae (20 spp), which is sister to all other "typical snakes" (Alethinophidia) [@mirallesMolecular2018; @burbrinkInterrogating2020; @uetzQuarter2021].
 
-Although all blindsnakes are fossorial, body size and shape vary significantly among species (Tiatragul et al.
-unpublished; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993).
+Although all blindsnakes are fossorial, body size and shape vary significantly among species [@shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993].
 Leptophlopids are generally thought to be more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@herrelBurrowing2021].
@@
 In all PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations.
-
 To visualise the relationship between vertebrae number and intracolumnar vertebral length, for each of the representative species, we plotted the mean vertebrae size (from three repeated measurements) against the vertebra number.
@@
 For the Australian blindsnakes, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig.
-\@ref(fig:fig-tbl-vert)).
+\@ref(fig:fig-tbl-vert) and **Fig S1**).
 This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae, while most species have many more vertebrae than expected for their size.
@@
 We found that vertebrae ratio is positively correlated with aspect ratio (PGLS, slope=-0.0048, intercept=0.73, R^2^=0.17, F=7.09, Z=2.20, P=0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig.
-\@ref(fig:fig-vert-aspect-ratios) and \@ref(fig:fig-vertebral-profile)).
+\@ref(fig:fig-vert-aspect-ratios)).
 We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope=-1.8, intercept=120.95, R^2^=0.69, F=79.39, Z=5.30, P<0.001).
@@
 
-knitr::include_graphics(paste0(WD, '/assets/figures/tbl-vert.jpg'))
+knitr::include_graphics(paste0(WD, '/assets/figures/phylomorpho-tbl-vert.jpg'))
 ```
@@
 Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter.
+(C) Vertebral profile shape of three species representing different vertebrae ratios, and are depicted as a 4° polynomial curve with positive inflection.
 Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.', out.width="124mm", out.height= "165mm"}
 
-knitr::include_graphics(paste0(WD, '/assets/figures/vert-aspect.jpg'))
+knitr::include_graphics(paste0(WD, '/assets/figures/phylomorpho-vert-aspect-total.jpg'))
+# knitr::include_graphics(paste0(WD, '/assets/figures/vert-aspect.jpg')) -->
 ```
@@
 <!-- Fig: Vertebral profile -->
-```{r fig-vertebral-profile, fig.cap= 'Relative vertebral length plotted against vertebra number for three example species with fitted 4° polynomial curves overlaid.', out.width="140mm", out.height= "197mm"}
+<!-- ```{r fig-vertebral-profile, fig.cap= 'Relative vertebral length plotted against vertebra number for three example species with fitted 4° polynomial curves overlaid.', out.width="140mm", out.height= "197mm"}
 
 knitr::include_graphics(paste0(WD, '/assets/figures/vertebral-profile.jpg'))
-```
+``` -->
 
@@
 <!-- Fig: Ratio and ecological variables -->
-```{r fig-vert-ecology, fig.cap= 'Relationship between mean vertebrae ratios and ecological variables.
-(A) Mean vertebrae against species mean annual temperature.
-(B) Mean vertebrae ratio against species max soil bulk density.
+```{r fig-vert-ecology, fig.cap= 'Phylomorphospaces showing relationship between mean vertebrae ratios and mean vertebrae number against ecological variables.
+Mean vertebrae ratio against (A) species mean annual temperature and (B) species max soil bulk density.
 Mean vertebrae ratio is calculated by mean total vertebrae / mean total length for each species.
+Mean vertebrae number against (C) species mean annual temperature and (D) species max soil bulk density.
 Species mean annual temperature and max soil bulk density values are calculated based on occurrence data.
@@
 
-knitr::include_graphics(paste0(WD, '/assets/figures/vert-ecology.jpg'))
+knitr::include_graphics(paste0(WD, '/assets/figures/phylomorpho-vert-ecology.jpg'))
 ```
@@
 
+Discuss amphisbaenians [@gansAmphisbaenia1978] and caecillians [@renousBody1989].
+
 ***[Discuss/speculate why vertebrae ratio is correlated with annual mean temperature and soil bulk density]***
@@
 
+REGIONALISATION Need to mention, plus the heart position aspect; how that’s not been studied here, but it helps to understand regionalisation, which is an important next step in understanding the lack of pleomerism.
+Since the lack of pleomerism in *Hydrophis* sea snakes is due to their elongate forebodies.
+Tamika’s paper on colubrids also shows the various ways elongation occurs, using different regions.
+Hampton 2023 paper on snake regionalisation in greater detail
+
 \newpage

### 2024-07-18 code/03_visualising_data.R

@@ -591,7 +591,6 @@ temp.pgls <- geomorph::procD.pgls(vert_ratio ~ annual_mean_temp, phy = anilios_t
 summary(temp.pgls)
 coefficients(temp.pgls)
 
-
 ### Using geomorph::procD.pgls 
 soil.pgls <- geomorph::procD.pgls(vert_ratio ~ bulk_density, phy = anilios_tree)
 summary(soil.pgls)
@@ -603,45 +602,48 @@ number_temp.pgls <- geomorph::procD.pgls(mean_vert ~ annual_mean_temp, phy = ani
 summary(number_temp.pgls)
 coefficients(number_temp.pgls)
 
-
 number_soil.pgls <- geomorph::procD.pgls(mean_vert ~ bulk_density, phy = anilios_tree)
 summary(number_soil.pgls)
 coefficients(number_soil.pgls)
 
-# Plot results vertebrae ratio against ecological variables
-
-pdf(file = "output/vert-ecology.pdf", width = 12, height = 9)
+# Figure 4: Combined results vertebrae ratio and max total vertebrae against ecological variables
+pdf(file = "output/phylomorpho-vert-ecology.pdf", width = 12, height = 9)
 par(mfrow=c(2,2))
-plot(vert_ratio ~ annual_mean_temp, bty="n", 
-     # cex = width_ratio_cex[names(width_ratio)], 
-     pch = 19,
-     xlab = "", ylab = "Vertebrae ratio")
-text(x = annual_mean_temp, y = vert_ratio, labels = sp_labs, pos = 1, cex = 0.7)
+
+
+### VERT RATIO VS ECOLOGY
+
+## Vert ratio v Annual mean temp
+phylomorphospace(anilios_tree, anilios_data[,c("temp_mean", "ver_rati")], 
+                 bty="n", label="horizontal", node.size=c(0.4,1.3),
+                 ylab = "Vertebrae ratio", xlab = "")
 abline(a = coefficients(temp.pgls)[1], b = coefficients(temp.pgls)[2])
 
-## Soil
-plot(vert_ratio ~ bulk_density, bty="n", pch = 19,
-     xlab = "", ylab = "")
-text(x = bulk_density, y = vert_ratio, labels = sp_labs, pos = 1, cex = 0.7)
+## VERT RATIO VS Soil bulk density
+phylomorphospace(anilios_tree, anilios_data[,c("max_bulk", "ver_rati")], 
[…]

### 2024-08-22 manuscript.Rmd
incorporated Emma's suggestions and inputs

@@
 
-* Axial evolution an important aspect of the vertebrate body plan evolution – facilitates diversity and influences locomotion 
-  <!-- Note the papers  that are about axial number relating to locomotion -->
-in mammals [@galisFast2014 ], lacertids [@vandammeSpeed2002], gobies [@yamadaAdaptive2009;@forkerBendy2021], and chameloens [@molnarMorphological2023].
+Axial evolution is fundamental to vertebrate body plan evolution, facilitating body form changes and influencing modes of locomotion in mammals [@eg1500;@taewcharoenSmall2024;@galisFast2014].
+reptiles [@eg1500; @vandammeSpeed2002;@molnarMorphological2023], amphibians [@eg1500;@bonettEvidence2017;@adlerEvolutionary2022] and fish [e.g., @yamadaAdaptive2009;@forkerBendy2021].
+Variation of the axial skeleton is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised.
+Essentially vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryongenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009]
+Elongate animals are expected to show pleomerism, where body size and number of vertebrae are positively correlated [@lindseyPleomerism1975]. \
+Pleomerism has been observed in eels and fish [@wardEvolution2007;@wardDifferential2014;@reeceEvolutionary2013], in squamates [@pollyTesting2001;@headDissociation2007;@headEvolution2015;@bergmannVertebral2012;@lindellEvolution1994;@cernanskyVertebral2019], in salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].
+However, there are notable exceptions in some taxa [@headDissociation2007;@sherrattMacroevolution2022] indicating it is possible to disassociate segment production with somatic (body) growth, and revealing developmental processes underlying evolutionary patterns.
+Snakes are therefore a good model to study axial evolution because they exhibit great diversity in vertebrae number across species and have a wide range of body sizes and shapes without limb skeleton to confound signals of locomotor adaptation.
+Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattHeterochronic2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024].
+It also has important influence on snake locomotion, where the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997].
+More research is required to understand the role that axial evolution plays in body plan and ecological diversification of snakes.
 
-* Need a reference for mammals? Limited variation of a common theme that is constrained [@asherVariability2011;@naritaEvolution2005]
-@liDivergent2023 found vertebral counts to differ significantly between major groups of mammals and within groups. 
-* Reptiles, on the other hand, are more varied [@headEvolution2015]
-
-* Elongate animals have special properties of vertebral number known as pleomerism, where body size and number of segments in vertebrae are correlated [@lindseyPleomerism1975]. 
-Pleomerism has been observed in eels and fish [@wardEvolution2007;@wardDifferential2014;@reeceEvolutionary2013], in squamates [@pollyTesting2001;@headDissociation2007;@headEvolution2015;@bergmannVertebral2012;@lindellEvolution1994;@cernanskyVertebral2019], in salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].
-
-* Snakes are a good model to study axial evolution because they have a high number of vertebrae and a wide range of body size and shape.
-* In elapid snakes [@sherrattVertebral2019;@sherrattMacroevolution2022;@hamptonInterspecific2019;@hamptonRegionalization2023]; colubrid snakes [@nash-hahnModification2024].
-
-***[Specific background]***
-
-Scolecophidians, often referred to in English as "blindsnakes", are characterised by their fossorial lifestyle, poor visual system, and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018;@strongDeconstructing2021;@gowerEyetranscriptome2021].
+<!-- [Specific background]-->
+Scolecophidians, commonly known in English as "blindsnakes" and "threadsnakes", are characterised by a reduced visual system and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018;@strongDeconstructing2021;@gowerEyetranscriptome2021].
 Scolecophidians comprise five paraphyletic families, including the "core scolecophidians" — a monophyletic clade consisting of Typhlopidae (425 spp), Leptotyphlopidae (143 spp), Gerrhopilidae (29 spp), and Xenotyphlopidae (1 spp) — and the paraphyletic Anomalepididae (20 spp), which is sister to all other "typical snakes" (Alethinophidia) [@mirallesMolecular2018; @burbrinkInterrogating2020; @uetzQuarter2021].
-
-Although all blindsnakes are fossorial, body size and shape vary significantly among species [@shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993].
-Leptophlopids are generally thought to be more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@herrelBurrowing2021].
+Although all blindsnakes are fossorial, body size and shape vary significantly among species [Tiatragul et al.
+unpublished; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993].
+Leptotyphlopids are generally thought to be more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@herrelBurrowing2021].
 These differences in body size and proportions determine how much push force the snakes can generate when burrowing.
@@
 
-***[Knowledge gap]***
+In another predominantly burrowing clade, caecilians, the number of vertebrae is positively correlated with body size [@renousBody1989], suggesting pleomerism is a mechanism for burrowing taxa to diversity in the underground environment.
+Pleomerism has been observed in typhlopidae, however, some notable exceptions indicate different mechanisms for gigantism [@headDissociation2007].
+For instance, Head and Polly [-@headDissociation2007] found that some of the largest species of blindsnakes in the African continent (e.g.
+*Afrotyphlops schlegelii*) have fewer than expected precloacal vertebrae than expected for their size.
+This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae.
+Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001].
+This is reflected in the relatively short tails of blindsnakes [Tiatragul unpublished; @szyndlarIllustrated2023].
+Limited studies on scolecophidian axial skeletons compared to alethinophidia means we have yet to appreciate the diversity that these groups have to offer.
+Given the recent shift in the perception of these earliest diverging clades of snakes (scolecophidians) away from the traditional view that they are "pleisomorphic and morphologically homogenous" [@strongConvergence2022], the scolecophidian axial skeleton also could offer a new perspective on the ecological origin of snakes.
 
@@
 However, information about the relationship between body size and vertebrae number is unknown.
-
-Among typhlopidae, gigantism have evolved multiple times but with different mechanisms [@headDissociation2007].
-For instance, @headDissociation2007 found that some of the largest species of blindsnakes in the African continent (e.g.
-*Afrotyphlops schlegelii*) have fewer than expected precloacal vertebrae than expected for their size.
-This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae.
-
-In another predominantly burrowing clade, caecilians, the number of vertebrae is positively correlated with body size [@renousBody1989], suggesting pleomerism is a mechanism for burrowing taxa to diversity in the underground environment.
-
-<!-- A big review included number of vertebrae [@szyndlarIllustrated2023] for many non-alethinophidian snakes, but  -->
-
-Compared to other snakes, the vertebral column of fossorial snakes tend to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001].
-This is reflected in the relatively short tails of blindsnakes (CITE).
-
-***[What we do in this paper]***
-
 In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes.
@@
 
-To obtain vertebral counts, we X-ray scanned the specimens at various institutions including: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n=87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n=129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (ANU) (n=90), and Queensland X-Ray (n=35).
+To obtain vertebral counts, we X-ray scanned the specimens at various institutions including: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n=87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n=129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (n=90), and Queensland X-Ray (n=35).
 
@@
 
-Additionally, we quantified intracolumnar length of each vertebra for three species representing 1) narrow-and-wide phenotypes (*A.
-grypus*), 2) wide-and-robust phenotypes (*A.
-pinguis*), and 3) intermediate phenotypes (*A.
+Additionally, we quantified intracolumnar length of each vertebra for three species representing (1) narrow-and-wide phenotypes (*A.
+grypus*), (2) wide-and-robust phenotypes (*A.
+pinguis*), and (3) intermediate phenotypes (*A.
 centralis*).
@@
 ## Environmental data
-<!-- NEED TO PUT IN THE CORRECT VALUES HERE -->
 
@@
 ```{r fig-tbl-vert, fig.cap= 'Maximum vertebral counts plotted against maximum total body length for Australian typhlopids.
-Trend line represent PGLS regression for Australian species.', out.width="165mm", out.height= "124mm"}
+Trend line represent PGLS regression for Australian species.
+Phylogeny projected into space using maximum likelihood ancestral state estimation.
+Species not included in the phylogeny are shown in purple.
+', out.width="165mm", out.height= "124mm"}
 
@@
 (C) Vertebral profile shape of three species representing different vertebrae ratios, and are depicted as a 4° polynomial curve with positive inflection.
-Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.', out.width="124mm", out.height= "165mm"}
+Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.
+Phylogeny projected into space using maximum likelihood ancestral state estimation.
+Species not included in the phylogeny are shown in purple.', out.width="124mm", out.height= "165mm"}
 
@@
 Body shape varied with the number of vertebrae in similar ways as previously reported in vipers and colubrids [@lindellEvolution1994]. 
-We can speculate that species that are narrow-and-long may be using pre-existing burrows rather than digging their own burrow.
+We speculate that species that are narrow-and-long may be using pre-existing burrows rather than digging their own burrow.
 Narrow-and-wide phenotypes are common among Leptotyphlopidae, which are thought to use existing burrows rather than dig their own burrow [@herrelBurrowing2021].
@@
 unpublished).
-Species that evolved more narrow-and-long bodies may be converging on the lifestyle with other scolecophidians that have similar phenotypes (i.e.
-leptotyphlopids).
+Species that evolved more narrow-and-long bodies may be converging on the lifestyle with other scolecophidians that have similar phenotypes (i.e., leptotyphlopids).
 On the other hand, robust-and-wide phenotypes common among typhlopid species are expected to be able to generate much more force and make their own burrows [@herrelBurrowing2021].
 
-Discuss amphisbaenians [@gansAmphisbaenia1978] and caecillians [@renousBody1989].
+Discuss amphisbaenians [@gansStudies1960] and caecillians [@renousBody1989].
 
@@
 Both mean annual temperature and soil bulk density were significant predictors of vertebrae ratio but in different directions.
-Mean annual temperature has been shown to be a proxy for several ecological factors such as rainfall and vegetation (CITE).
-However, temperature has also been shown to directly influence multiple axes of reptile development (cite).
-Vertebral development (somitogenesis) and homeotic (*Hox gene*) regulation may thus be affected by variations in incubation conditions [@martin-del-campoHox2018].
+Mean annual temperature has been shown to be a good proxy for ecological factors relevant to squamates [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013].
+However, temperature has also been shown to directly influence multiple axes of reptile development [@nobleDevelopmental2018;@boothInfluence2006;@feinerDevelopmental2018].
+Vertebral development (somitogenesis) and homeotic (*Hox* gene) regulation may thus be affected by variations in incubation conditions [@martin-del-campoHox2018].
 However, we currently lack knowledge about egg-laying behaviour and how temperature affects the development of blindsnake embryos.
@@
 Traditionally, ventral scale counts have been used as a proxy for vertebral number as they correlate for most alethinophidian snakes [@alexanderPattern1966].
-We found that this is (not) the case in blindsnakes.
+We found that this is not the case in blindsnakes.
+
 Blindsnakes are generally small enough to fit in an X-ray scanner, even in portable field X-ray machines.

### 2024-08-23 manuscript.Rmd
discussion paragraph 1

@@
 
-In another predominantly burrowing clade, caecilians, the number of vertebrae is positively correlated with body size [@renousBody1989], suggesting pleomerism is a mechanism for burrowing taxa to diversity in the underground environment.
+<!-- In another predominantly burrowing clade, caecilians, the number of vertebrae is not always correlated with body size [@renousBody1989], suggesting pleomerism is a mechanism for burrowing taxa to diversify in the underground environment.
+-->
+
 Pleomerism has been observed in typhlopidae, however, some notable exceptions indicate different mechanisms for gigantism [@headDissociation2007].
@@
 Limited studies on scolecophidian axial skeletons compared to alethinophidia means we have yet to appreciate the diversity that these groups have to offer.
-Given the recent shift in the perception of these earliest diverging clades of snakes (scolecophidians) away from the traditional view that they are "pleisomorphic and morphologically homogenous" [@strongConvergence2022], the scolecophidian axial skeleton also could offer a new perspective on the ecological origin of snakes.
+Given the recent shift in the perception of these earliest diverging clades of snakes (scolecophidians) away from the traditional view that they are "plesiomorphic and morphologically homogenous" [@strongConvergence2022], the scolecophidian axial skeleton also could offer a new perspective on the ecological origin of snakes.
 
@@
 \@ref(fig:fig-tbl-vert) and **Fig S1**).
-This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae, while most species have many more vertebrae than expected for their size.
+This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae.
+Many species have more vertebrae than expected for their size.
 For example, *A.
-grypus* "West" is an intermediate length (relative to other species) and has the most vertebrae, while *A.
+grypus* "West" is an intermediate length (relative to other species) but has the greatest number of vertebrae, while *A.
 torresianus*, which is the longest, has relatively few vertebrae.
@@
 
-***[Big summary]***
+<!-- However, information about the relationship between body size and vertebrae number is unknown.
+In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes.
+The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes?
+(2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes?
+(3) Does vertebral ratio vary with soil compactness or mean annual temperature?
+-->
 
-Results from our study revealed that pleomerism does not apply to Australian typhlopids, indicating a deviation from the positive correlation between body size and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975].
+Our study investigated the presence of pleomerism in Australian blindsnakes and test whether deviations from pleomerism may be explained by body shape or ecological factors.
+Our results revealed that pleomerism does not generally apply to Australian typhlopids, indicating a deviation from the positive correlation between body size and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975].
 Variation in body shape, however, is underpinned by the ratio between vertebrae number to body size.
@@
 This relationship between the proportion of vertebrae to body shape has been found in colubrids and viperids, where more elongated species tend to have a greater number of vertebrae than stouter ones [@lindellEvolution1994].
-Variation in vertebral ratio can also be explained by both mean annual temperature and soil compactness.
+We also found that vertebral ratio correlate positively with mean annual temperature and negatively with soil compactness.
 These results have important implications for understanding the macroevolutionary dynamics of continental-scale radiation of Australian blindsnakes.
 
-***[Discuss why body shape might influence vertebrae ratio or vice-versa]***
+### Absence of pleomerism in Australian blindsnakes
+
+We found a dissociation between vertebrae number and body size in Australian blindsnakes.
+This is consistent with a previous study on blindsnakes from other continents, where exceptionally large species are shown to have fewer than expected pre-cloacal vertebrae for their size [@headDissociation2007].
+Several mechanisms have been suggested for how giant amniotes achieve their large size including prolonging their growth [@ericksonHow1999] and increasing the rate of growth [@ericksonGigantism2004].
+Unforutnately, without life-history data for blindsnakes, we cannot determine which of these mechanisms are responsible.
+Elongation may also be achieved in other ways.
+The regionalisation of vertebrae, where distinct regions of the vertebrae have different evolutionary pathways and morphological diversification [@hamptonRegionalization2023;@hamptonHeart2022], may explain the lack of pleomerism.
+For instance, the absence of pleomerism in *Hydrophis* sea snakes has been attributed to the increase in the numnber of smaller vertebrae in the pre-atrial region compared to larger vertebrae in the post-atrial region [@sherrattMacroevolution2022;@sherrattHeterochronic2019].
+Similarly, colubrids have been also been shown to have different regionalisation patterns in the vertebral column that contributes to elongation in different ways [@nash-hahnModification2024].
+Another important aspect of regionalisation is the position of the heart, which has been shown to be a key marker for delimiting different regions of the vertebrae [@hamptonHeart2022;@hamptonRegionalization2023].
+Future research should investigate the regionalisation of the vertebral column in blindsnakes as this will provide important insights to understanding the lack of pleomerism.
+
+### Correlation between body shape and vertebrae ratio
 
@@
 
-REGIONALISATION Need to mention, plus the heart position aspect; how that’s not been studied here, but it helps to understand regionalisation, which is an important next step in understanding the lack of pleomerism.
-Since the lack of pleomerism in *Hydrophis* sea snakes is due to their elongate forebodies.
-Tamika’s paper on colubrids also shows the various ways elongation occurs, using different regions.
-Hampton 2023 paper on snake regionalisation in greater detail
-
 \newpage

### 2024-08-23 manuscript.Rmd
small adjustments

@@
 
-***[Discuss/speculate why vertebrae ratio is correlated with annual mean temperature and soil bulk density]***
+### Vertebrae ratio and ecological variables
+<!-- Discuss/speculate why vertebrae ratio is correlated with annual mean temperature and soil bulk density]*** -->
 
 Both mean annual temperature and soil bulk density were significant predictors of vertebrae ratio but in different directions.
+
+We found that species that occupy warmer environments has greater vertebrae ratio than those in cooler places.
+
 Mean annual temperature has been shown to be a good proxy for ecological factors relevant to squamates [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013].
-However, temperature has also been shown to directly influence multiple axes of reptile development [@nobleDevelopmental2018;@boothInfluence2006;@feinerDevelopmental2018].
+Temperature also directly influence multiple axes of reptile development [@nobleDevelopmental2018;@boothInfluence2006;@feinerDevelopmental2018].
 Vertebral development (somitogenesis) and homeotic (*Hox* gene) regulation may thus be affected by variations in incubation conditions [@martin-del-campoHox2018].

### 2024-09-18 manuscript.Rmd
added discussions and incorporated suggestions from Emma.

--- 2024-08-23 manuscript.Rmd
@@
 ---
-title: "Dissociation between vertebrae number and body size in Australian blindsnakes"
+title: "Dissociation between number of vertebrae and body size in Australian blindsnakes"
 date: "Version `r Sys.Date()`"
@@
 ```{r packages}
-library(dplyr); library(readr); library(janitor); 
-library(ape); library(phytools);
+library(dplyr); 
 library(kableExtra)
@@
 
-Keywords: Scolecophidia, vertebrae, pleomerism, snakes,
+Keywords: Scolecophidia, vertebrae, pleomerism, snakes, axial
 
@@
 Axial evolution is fundamental to vertebrate body plan evolution, facilitating body form changes and influencing modes of locomotion in mammals [@eg1500;@taewcharoenSmall2024;@galisFast2014].
-reptiles [@eg1500; @vandammeSpeed2002;@molnarMorphological2023], amphibians [@eg1500;@bonettEvidence2017;@adlerEvolutionary2022] and fish [e.g., @yamadaAdaptive2009;@forkerBendy2021].
+reptiles [@eg1500;@vandammeSpeed2002;@molnarMorphological2023], amphibians [@eg1500;@bonettEvidence2017;@adlerEvolutionary2022] and fish [e.g., @yamadaAdaptive2009;@forkerBendy2021].
 Variation of the axial skeleton is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised.
-Essentially vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryongenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009]
-Elongate animals are expected to show pleomerism, where body size and number of vertebrae are positively correlated [@lindseyPleomerism1975]. \
+Essentially vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryogenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009].
+Elongate animals are expected to show pleomerism, where body size and number of vertebrae are positively correlated [@lindseyPleomerism1975].
 Pleomerism has been observed in eels and fish [@wardEvolution2007;@wardDifferential2014;@reeceEvolutionary2013], in squamates [@pollyTesting2001;@headDissociation2007;@headEvolution2015;@bergmannVertebral2012;@lindellEvolution1994;@cernanskyVertebral2019], in salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].
@@
 Scolecophidians comprise five paraphyletic families, including the "core scolecophidians" — a monophyletic clade consisting of Typhlopidae (425 spp), Leptotyphlopidae (143 spp), Gerrhopilidae (29 spp), and Xenotyphlopidae (1 spp) — and the paraphyletic Anomalepididae (20 spp), which is sister to all other "typical snakes" (Alethinophidia) [@mirallesMolecular2018; @burbrinkInterrogating2020; @uetzQuarter2021].
-Although all blindsnakes are fossorial, body size and shape vary significantly among species [Tiatragul et al.
-unpublished; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993].
+Although all blindsnakes are fossorial, body size and shape vary significantly among species [@tiatragulMorphological2024; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993].
 Leptotyphlopids are generally thought to be more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@herrelBurrowing2021].
@@
 Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001].
-This is reflected in the relatively short tails of blindsnakes [Tiatragul unpublished; @szyndlarIllustrated2023].
+This is reflected in the relatively short tails of blindsnakes [@tiatragulMorphological2024;@szyndlarIllustrated2023].
 Limited studies on scolecophidian axial skeletons compared to alethinophidia means we have yet to appreciate the diversity that these groups have to offer.
@@
 
-The Australian blindsnake radiation comprise 47 described species [@tiatragulAustralia2023] with considerably diverse morphology from narrow-and-long (aspect ratio >= 100) to robust-and-wide (aspect ratio < 60).
+The Australian blindsnake radiation comprise 47 described species [@tiatragulAustralia2023] distributed across all major biomes of Australia [@tiatragulPaleoenvironmental2023] with considerably diverse morphology from narrow-and-long (aspect ratio >= 100) to robust-and-wide (aspect ratio < 60).
 However, information about the relationship between body size and vertebrae number is unknown.
@@
 We sourced specimens from South Australian Museum (SAMA), Western Australian Museum (WAM), Queensland Museum (QM), Museum and Art Gallery of the Northern Territory (MAGNT), and a private collection.
-Where possible, we chose relatively large non-fragile specimens from both males and females with well-preserved heads for X-ray.
-Sex for each specimen were determined by inspecting the gonads or assigned based on relative tail lengths following Tiatragul et al.
-(unpublished).
+Where possible, we chose relatively large non-fragile specimens from both males and females with well-preserved heads for X-ray radiography.
+Sex for each specimen were determined by inspecting the gonads or assigned based on relative tail lengths following @tiatragulMorphological2024.
 
@@
 
-We used the Fiji distribution of ImageJ2 v.1.54j [@schindelinFiji2012;@ruedenImageJ22017] to digitally visualise X-ray images and counted the number of vertebrae using Cell Counter module.
+We used the Fiji distribution of ImageJ2 v.1.54j [@schindelinFiji2012;@ruedenImageJ22017] to digitally visualise X-ray images and counted the number of vertebrae using the Cell Counter module.
 Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023].
@@
 
-We used the Australian blindsnake occurrence records presented in Tiatragul et al.
-(unpublished) including a total of 3898 verified records (min=1, median=23, mean=81, max=525 per species).
+We used the Australian blindsnake occurrence records presented in @tiatragulMorphological2024 including a total of 3898 verified records (min=1, median=23, mean=81, max=525 per species).
 These records were compiled from museum databases and verified human observations from Atlas of the Living Australia (ALA).
@@
 Soil bulk density measures the penetrability of soil, which is a proxy for the type of microhabitat species is most likely living in as they are all strictly fossorial.
+Mean annual temperature is a climate proxy and is commonly used to describe niche for squamate reptiles [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013].
+Temperature also directly influence multiple axes of reptile development [@nobleDevelopmental2018;@boothInfluence2006;@feinerDevelopmental2018], which influences vertebrae development [@martin-del-campoHox2018].
 Bulk density values > 1.6 g/cm<sup>3</sup> indicate soil that restrict penetration by plant roots, while values <0.5 indicate looser, more organic soil [@mckenzieSoil2002].
-Mean annual temperature is a climate proxy and is commonly used to describe niche for squamate reptiles [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013].
 
@@
 First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes.
-We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum total body length from Tiatragul et al.
-(unpublished).
-Next, we test whether the vertebrae to body length ratio is related to the mean vertebrae number and mean aspect ratio, defined by total body length / body width, also from Tiatragul et al.
-(unpublished) data.
+We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum total body length from @tiatragulMorphological2024.
+Next, we test whether the vertebrae to body length ratio is related to the mean vertebrae number and mean aspect ratio, defined by total body length / body width, also from @tiatragulMorphological2024.
 Finally, we tested if variation in vertebrae ratio and vertebrae number can be explained by mean annual temperature and soil compactness.
@@
 
-For the Australian blindsnakes, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig.
+For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig.
 \@ref(fig:fig-tbl-vert) and **Fig S1**).
@@
 
-### Relationship between vertebrae ratio, number, and body shape
+### Relationships between vertebrae ratio, number, and aspect ratio
 We found that vertebrae ratio is positively correlated with aspect ratio (PGLS, slope=-0.0048, intercept=0.73, R^2^=0.17, F=7.09, Z=2.20, P=0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig.
@@
 
- <!-- The specific questions we addressed are: (1) How prevalent is pleomerism among blindsnakes?
-(2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes?
-(3) Does vertebral ratio vary with mean annual temperature or soil compactness?
+Our study investigated the presence of pleomerism in Australian blindsnakes and test whether deviations from pleomerism may be explained by body shape or ecological factors.
+Our results revealed that pleomerism does not generally apply to Australian typhlopids, indicating a deviation from the positive correlation between body size and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975].
+Variation in body shape, however, is underpinned by the ratio between vertebrae number to body size.
+We show that species that are narrow-and-long have proportionally more vertebrae compared to species that are more robust-and-wide.
+We also found that vertebral ratio correlates positively with mean annual temperature and negatively with soil compactness.
+These results have important implications for understanding the axial evolution of basal snakes and the macroevolutionary dynamics of Australian blindsnakes.
+
+### Absence of pleomerism in Australian blindsnakes
+
+We found a dissociation between vertebrae number and body size in Australian blindsnake.
+While pleomerism has been documented in most snake groups [@lindellEvolution1994], exceptionally large species ("giant taxa") within boids, pythonids, and blindsnakes such as *Afrotyphlops schlegelii* have been shown to have fewer than expected pre-cloacal vertebrae for their size [@headDissociation2007].
+In line with these observations, some of the largest species in our dataset (*A.
+torresianus* and *A.
+pinguis*) possess lower than average number of vertebrae (Fig.
+\@ref(fig:fig-tbl-vert)).
+The absence of pleomerism suggest that the large size attained by these "giant taxa" must be due to modification of post-somitogenetic somatic growth [@headDissociation2007].
+In other words, species with larger body size do not simply have more vertebrae segments, rather the size of each vertebral segment has been elongated during growth.
+Several mechanisms have been suggested for how giant amniotes achieve their large size including prolonging their growth [@ericksonHow1999] and increasing the rate of growth [@ericksonGigantism2004].
+Unfortunately, without life-history data for blindsnakes, we cannot determine which of these mechanisms are responsible.
+However, elongation may also be achieved in other ways.
+The regionalisation of vertebrae, where distinct regions of the vertebrae have different evolutionary pathways and morphological diversification [@hamptonRegionalization2023;@hamptonHeart2022], may explain the lack of pleomerism in blindsnakes.
+For example, the absence of pleomerism in *Hydrophis* sea snakes has been attributed to the increase in the number of smaller vertebrae in the pre-atrial region compared to larger vertebrae in the post-atrial region [@sherrattMacroevolution2022;@sherrattHeterochronic2019].
+Similarly, colubrids have also been shown to have different regionalisation patterns in the vertebral column that contributes to elongation in different ways [@nash-hahnModification2024].
+Another important aspect of regionalisation is the position of the heart, which has been shown to be a key marker for delimiting different regions of the vertebrae [@hamptonHeart2022;@hamptonRegionalization2023].
+Future research should investigate the regionalisation of the vertebral column in blindsnakes as this will provide important insights to understanding the lack of pleomerism.
+
+The lack of pleomerism among burrowing vertebrates is not uncommon.
+@lindellEvolution1994 found that among Australian elapids and African snakes, burrowing species had less vertebrae than expected from its body size compared to terrestrial and arboreal species.
+The dissociation between vertebrae number and body size has also been documented in caecilians, a primarily fossorial order of amphibians.
+In caecilians, @renousBody1989 found that members of the same genus will vary in size but generally share similar number of vertebrae indicating a lack of pleomerism except for in a few genera such as *Oscaecilia*.
+Another group of highly specialised burrowing reptile is the amphisbaenians [@gansCharacteristics1978;@gansAmphisbaeniansreptiles1969].
+While several studies have documented the number of vertebrae in some species [@bergmannVertebral2012;@gansStudies1960;@gansStudies1971;@gansCharacteristics1978], no study has investigated the correlation between vertebrae number and body size in amphisbaenians.
+We recommend future research to comprehensively investigate the prevalence of pleomerism among burrowing limbless and limb-reduced vertebrates to improve our understanding of the evolution of axial system.
+
+### Correlation between vertebrae ratio, number, and aspect ratio
+
+Aspect ratio (body shape) varied with the number of vertebrae in similar ways as previously reported in fish and other snakes (e.g.
+vipers and colubrids), where species with a more elongated body had relatively greater number of vertebrae than stouter ones [@lindseyPleomerism1975;@lindellEvolution1994].
+Two possible explanations have been proposed by @lindellEvolution1994 including adaptations for locomotion and prey consumption.
+@poughSpecializations1983 showed that in snakes, longer body segments may be advantageous for elongated species that usually move faster, whereas shorter segments are favoured for stouter species that move more slowly.
+In blindsnakes, however, movement speed may not be as important as how much push force they can generate when burrowing.
+Snakes with relatively wider bodies are generally able to produce greater push force for their body size [@herrelBurrowing2021], but the trade-off is species with narrower bodies will be able to use pre-existing small crevices and galleries without needing to expend any force.
+In regards to prey consumption, stout species tend to consume large bulky prey, which is facilitated by having shorter body segment [@poughSpecializations1983;@lindellEvolution1994;@moonFeeding2019].
+Although Australian blindsnakes primarily prey on pupae and larvae of ants and termites, there is some variation in the size of prey consumed [@webbDietary1993].
+In particular, heavy-bodied blindsnake species (e.g.
+*A.
+pinguis*, *A.
+proximus*, and *A.
+ligatus*) fed on larger species of ants compared to more slender-bodied blindsnakes species (e.g.
+*A.
+hamatus*, *A.
+bituberculatus*, *A.
+waitii*, *A.
+weidii*) [@webbDietary1993].
+
+<!-- ** Discuss how longer body allow more force by elongating muscles @navasMorphological2004 ** -->
+
+### Correlations between vertebrae ratio, number, and ecological variables
+<!-- Discuss/speculate why vertebrae ratio is correlated with annual mean temperature and soil bulk density]*** -->
+
+Both mean annual temperature and soil bulk density were significant predictors of vertebrae ratio but not for the raw number of vertebrae.
+While non-statistically significant, we observed a slightly positive relationship between raw number of vertebrae and mean annual temperature.
+Previous studies across different species of rattlesnakes [@klauberRattlesnakes1982;@klauberCorrelation1941] and between populations of garter snakes (*Thamnophis elegans*) [@arnoldQuantitative1988] both reported a similar pattern where species that occupy warmer and dryer environments had more vertebrae than those living in colder and humid coastal areas.
+While the temperature embryos are exposed to during development have been shown to affect vertebral numbers in snakes [@foxMorphological1961;@foxEffect1948;@osgoodEffects1978], @arnoldQuantitative1988 suggested three reasons it is unlikely the explanation for the regional difference in the number of vertebrae observed in *T.
+elegans*.
+First, the observed difference in the number of vertebrae between populations is much greater than the effects of temperature produced lab experiments [@arnoldQuantitative1988;@osgoodEffects1978].
+Second, the number of vertebrae of the F1 hybrid is intermediate between inland and coastal parents [@arnoldQuantitative1988].
+Finally, females can select microhabitats that buffer their developing embryos against extreme climatic conditions [@arnoldQuantitative1988;@brownMaternal2004;@burgerNest1986;@peet-pareFemale2012].
+Although we currently lack knowledge about egg-laying behaviour and how temperature affects the development of blindsnake embryos, we suspect that temperature effects alone is unlikely to explain the interspecific variation in vertebrae number.
+
+In a related study, @tiatragulMorphological2024 found that Australian blindsnake species with smaller absolute body size (SVL) generally occupy areas that areas that warmer than larger species.
+Here, we observe that vertebrae ratio (raw vertebrae number/SVL) is positively correlated with temperature, indicating species that have more vertebrae than expected from their body size are found in warmer places.
+In his work on rattlesnakes, @klauberRattlesnakes1982  noted that smaller-sized subspecies often retain the same number of vertebrae with their ancestors resulting in greater vertebrae ratio.
+An explanation for this pattern is that size is often under greater selection than vertebral number, and there is negligible disadvantage for a snake in having too many vertebrae for their body size [@lindellEvolution1994].
+This makes sense when we also consider the negative relationship between vertebrae ratio and maximum soil bulk density.
+A greater number of vertebrae for body size may afford greater flexibility [@jayneKinematics1986;@bergmannVertebral2012] that enables these species to navigate in pre-existing burrows or termite galleries, typical across northern Australia, without exerting force to dig their own burrow.
+@tiatragulMorphological2024 also found species that occupy areas with more difficult to-penetrate soil are usually narrow-and-long rather than robust-and-wide species that.
+Narrow-and-wide phenotypes are common among leptotyphlopids, which are thought to use existing burrows rather than digging their own [@herrelBurrowing2021].
+In this regard, the Australian typhlopid genus, *Anilios*, is quite exceptional because there is a great variation in body shape [@tiatragulMorphological2024].
+Species that evolved more narrow-and-long bodies may be converging on the behaviour with other scolecophidians that have similar phenotypes (i.e., leptotyphlopids).
+However, more detailed species-level microhabitat observations are required to determine if these species are actually taking advantage of these environments in the ways we speculated.
+
+<!-- ## Miscellaneous  -->
+<!-- Traditionally, ventral scale counts have been used as a proxy for vertebral number as they correlate for most alethinophidian snakes [@alexanderPattern1966].
+We found that this is not the case in blindsnakes.
 -->
 
-<!-- However, information about the relationship between body size and vertebrae number is unknown.
-In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes.
-The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes?
+## Conclusion
+
+<!-- (1) How prevalent is pleomerism among Australian blindsnakes?
 (2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes?
@@
 
-Our study investigated the presence of pleomerism in Australian blindsnakes and test whether deviations from pleomerism may be explained by body shape or ecological factors.
-Our results revealed that pleomerism does not generally apply to Australian typhlopids, indicating a deviation from the positive correlation between body size and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975].
-Variation in body shape, however, is underpinned by the ratio between vertebrae number to body size.
-We show that species that are narrow-and-long have proportionally more vertebrae compared to species that are more robust-and-wide.
-This relationship between the proportion of vertebrae to body shape has been found in colubrids and viperids, where more elongated species tend to have a greater number of vertebrae than stouter ones [@lindellEvolution1994].
-We also found that vertebral ratio correlate positively with mean annual temperature and negatively with soil compactness.
-These results have important implications for understanding the macroevolutionary dynamics of continental-scale radiation of Australian blindsnakes.
+Compared to other genera of blindsnakes, the Australo-Papuan genus *Anilios* show exceptional diversity in body size and shape [@tiatragulMorphological2024;@herrelBurrowing2021], which make them a good group to study the evolution of axial skeleton.
+We found a lack of pleomerism among the genus, with exceptionally large species possessing a smaller number of vertebrae than expected for their body size.
+We also found substantial variation in the number of vertebrae among species, and this variation is correlated in one way or another with ecological variables considered.
+The number of vertebrae is also correlated with body shape with species that are narrow-and-long possessing more vertebrae than expected for body size compared to species that are robust-and-wide.
+This study adds to a growing body of literature on the axial skeleton of snakes, a field that has largely been neglected.
 
-### Absence of pleomerism in Australian blindsnakes
-
-We found a dissociation between vertebrae number and body size in Australian blindsnakes.
-This is consistent with a previous study on blindsnakes from other continents, where exceptionally large species are shown to have fewer than expected pre-cloacal vertebrae for their size [@headDissociation2007].
-Several mechanisms have been suggested for how giant amniotes achieve their large size including prolonging their growth [@ericksonHow1999] and increasing the rate of growth [@ericksonGigantism2004].
-Unforutnately, without life-history data for blindsnakes, we cannot determine which of these mechanisms are responsible.
-Elongation may also be achieved in other ways.
-The regionalisation of vertebrae, where distinct regions of the vertebrae have different evolutionary pathways and morphological diversification [@hamptonRegionalization2023;@hamptonHeart2022], may explain the lack of pleomerism.
-For instance, the absence of pleomerism in *Hydrophis* sea snakes has been attributed to the increase in the numnber of smaller vertebrae in the pre-atrial region compared to larger vertebrae in the post-atrial region [@sherrattMacroevolution2022;@sherrattHeterochronic2019].
-Similarly, colubrids have been also been shown to have different regionalisation patterns in the vertebral column that contributes to elongation in different ways [@nash-hahnModification2024].
-Another important aspect of regionalisation is the position of the heart, which has been shown to be a key marker for delimiting different regions of the vertebrae [@hamptonHeart2022;@hamptonRegionalization2023].
-Future research should investigate the regionalisation of the vertebral column in blindsnakes as this will provide important insights to understanding the lack of pleomerism.
-
-### Correlation between body shape and vertebrae ratio
-
-Body shape varied with the number of vertebrae in similar ways as previously reported in vipers and colubrids [@lindellEvolution1994]. 
-We speculate that species that are narrow-and-long may be using pre-existing burrows rather than digging their own burrow.
-Narrow-and-wide phenotypes are common among Leptotyphlopidae, which are thought to use existing burrows rather than dig their own burrow [@herrelBurrowing2021].
-In this regard, the Australian typhlopid genus, *Anilios*, is quite exceptional because there is a great variation in body shape (Tiatragul et al.
-unpublished).
-Species that evolved more narrow-and-long bodies may be converging on the lifestyle with other scolecophidians that have similar phenotypes (i.e., leptotyphlopids).
-On the other hand, robust-and-wide phenotypes common among typhlopid species are expected to be able to generate much more force and make their own burrows [@herrelBurrowing2021].
-
-Discuss amphisbaenians [@gansStudies1960] and caecillians [@renousBody1989].
-
-***[Discuss/speculate why vertebrae ratio is correlated with annual mean temperature and soil bulk density]***
-
-Both mean annual temperature and soil bulk density were significant predictors of vertebrae ratio but in different directions.
-Mean annual temperature has been shown to be a good proxy for ecological factors relevant to squamates [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013].
-However, temperature has also been shown to directly influence multiple axes of reptile development [@nobleDevelopmental2018;@boothInfluence2006;@feinerDevelopmental2018].
-Vertebral development (somitogenesis) and homeotic (*Hox* gene) regulation may thus be affected by variations in incubation conditions [@martin-del-campoHox2018].
-However, we currently lack knowledge about egg-laying behaviour and how temperature affects the development of blindsnake embryos.
-Common garden experiments across different species of blindsnakes may be able to shed light on whether variation in incubation temperature and consequently, the rate of development may alter the number of vertebrae formation during development.
-<!-- We would expect higher temperature to accelerate development (CITE anolis papers, turtles, etc), so less time for embryonic development.  -->
-
-A previous study on Australian blindsnakes (Tiatragul et al.
-unpublished) found that robust-and-wide species tend to occupy areas in Australia that have easier-to-penetrate soil compared to narrow-and-long species that tend to occupy areas with more difficult-to-penetrate soil.
-A greater number of vertebrae in a narrow-and-long body may afford greater flexibility that enables these species to navigate in pre-existing burrows or termite galleries, typical across northern Australia, without exerting force to dig their own burrow.
-However, more detailed species-level microhabitat observations are required to determine if body shape is related to the environment in which they burrow.
-
-## Miscellaneous 
-Traditionally, ventral scale counts have been used as a proxy for vertebral number as they correlate for most alethinophidian snakes [@alexanderPattern1966].
-We found that this is not the case in blindsnakes.
-
-Blindsnakes are generally small enough to fit in an X-ray scanner, even in portable field X-ray machines.
+<!-- Blindsnakes are generally small enough to fit in an X-ray scanner, even in portable field X-ray machines.
 Collecting more data on this group of basal snakes can help us understand how vertebral morphology evolved in snakes.
+-->
 

### 2024-10-09 manuscript.Rmd
addressed comments from Scott

@@
 ---
-title: "Dissociation between number of vertebrae and body size in Australian blindsnakes"
+# title: "Dissociation between number of vertebrae and body size in Australian blindsnakes"
+title: "Testing the correlates of huge vertebrae number disparity in Australian blindsnakes"
 date: "Version `r Sys.Date()`"
@@
 
-**Authors**: Sarin Tiatragul^1^; Andrew Amey^2^; Scott Hocknull^2^; Scott Keogh^1^; Emma Sherratt^3^
+**Authors**: Sarin Tiatragul^1^; Andrew Amey^2^; Scott Hocknull^2^; J. Scott Keogh^1^; Emma Sherratt^3^
 
@@
 
-<!-- **Author contributions**:  -->
+**CRediT author statement**:
 
-<!-- ST and ES conceptualised the study; ST collected morphology data;  -->
+ST: Conceptualisation, Methodology, Formal analysis, Writing - Original draft, reviewing and editing.
+AM: Resources, Investigation
+SH: Resources, Investigation
+SK: Supervision, Funding acquisition, Writing - reviewing and editing.
+ES: Conceptualisation, Methodology, Writing - reviewing and editing.
+
+<!-- ST and ES conceptualised the study; ST collected morphology data -->
 <!-- ST obtained X-rays from SAM, WAM, and MAGNT. SH and AM provided X-rays Queensland Museum specimens. -->
@@
 
-<!-- **Acknowledgements (including funding information)**: -->
+**Acknowledgements**:
 
-<!-- We to thank museum curators and collection managers from the  -->
-<!-- South Australian Museum  (R Foster, S Donnellan), Western Australian Museum (P Doughty), Queensland Museum (A Amey and P Couper), Australian National Wildlife Collection (T Haff, C Wilson, and L Joseph), Museum and Art Gallery of the Northern Territory (G Dally), Museums Victoria (J Sumner, J Melville, and R Aguilar), Florida Museum of Natural History (J Gray, CM Sheehy, and ZS Randall), and the Natural History Museum Berlin (K Mahlow and F Tillack).
--->
-<!-- SK thanks the Australian Research Council for ongoing support.
-ST PhD scholarship is supported by the Australian Government Research Training Program Scholarship at the Australian National University.
--->
+We thank museum curators and collection managers from the 
+South Australian Museum, Western Australian Museum, Queensland Museum, Australian National Wildlife Collection, Museum and Art Gallery of the Northern Territory, Museums Victoria, and the Florida Museum of Natural History.
+SK thanks the Australian Research Council for ongoing support.
+ST PhD Scholarship was supported by the Australian Government Research Training Program Scholarship at the Australian National University.
 
@@
 
-Keywords: Scolecophidia, vertebrae, pleomerism, snakes, axial
+The evolution of the axial skeleton is important in understanding the evolution of body plans in vertebrates.
+Changes in the vertebral skeleton can influence body form and modes of locomotion.
+Elongate animals are commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism.
+However, some groups of vertebrates show a lack of pleomerism, suggesting it is possible to dissociate the production of segment and somatic growth, but what factors influenced this pattern is not well understood.
+Snakes offer a great model to study axial evolution because the number of vertebrae vary greatly between species and they exhibit a wide range of body size and shape.
+Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology can explain the variation in vertebrae number.
+We obtained X-ray scans and morphometric data for 42 described species of *Anilios* blindsnakes and evaluated their correlation with each other and ecological variables.
+We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebrae ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species.
+Finally, we found that species that have a greater number of vertebrae for their size are generally found in warmer environments and environments with harder-to-penetrate soil substrate.
+Our results extend our understanding of the axial evolution of basal snakes and the macroevolutionary dynamics of Australian blindsnakes.
+
+Keywords: Scolecophidia, vertebrae, pleomerism, skeleton, axial
 
@@
 
-Axial evolution is fundamental to vertebrate body plan evolution, facilitating body form changes and influencing modes of locomotion in mammals [@eg1500;@taewcharoenSmall2024;@galisFast2014].
-reptiles [@eg1500;@vandammeSpeed2002;@molnarMorphological2023], amphibians [@eg1500;@bonettEvidence2017;@adlerEvolutionary2022] and fish [e.g., @yamadaAdaptive2009;@forkerBendy2021].
-Variation of the axial skeleton is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised.
+Evolution of the axial skeleton is fundamental to vertebrate body plan evolution, facilitating body form changes and influencing modes of locomotion in mammals [e.g., @taewcharoenSmall2024;@galisFast2014], reptiles [e.g., @vandammeSpeed2002;@molnarMorphological2023], amphibians [e.g., @bonettEvidence2017;@adlerEvolutionary2022] and fish [e.g., @yamadaAdaptive2009;@forkerBendy2021].
+Axial skeleton evolution is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised.
 Essentially vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryogenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009].
-Elongate animals are expected to show pleomerism, where body size and number of vertebrae are positively correlated [@lindseyPleomerism1975].
-Pleomerism has been observed in eels and fish [@wardEvolution2007;@wardDifferential2014;@reeceEvolutionary2013], in squamates [@pollyTesting2001;@headDissociation2007;@headEvolution2015;@bergmannVertebral2012;@lindellEvolution1994;@cernanskyVertebral2019], in salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].
-However, there are notable exceptions in some taxa [@headDissociation2007;@sherrattMacroevolution2022] indicating it is possible to disassociate segment production with somatic (body) growth, and revealing developmental processes underlying evolutionary patterns.
-Snakes are therefore a good model to study axial evolution because they exhibit great diversity in vertebrae number across species and have a wide range of body sizes and shapes without limb skeleton to confound signals of locomotor adaptation.
-Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattHeterochronic2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024].
-It also has important influence on snake locomotion, where the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997].
-More research is required to understand the role that axial evolution plays in body plan and ecological diversification of snakes.
+Elongate animals are expected to show a positive correlation between maximum adult body length and number of vertebrae, known as pleomerism [@lindseyPleomerism1975].
+While pleomerism was considered the norm [@lindseyPleomerism1975;@wakeComparative1966;@pollyTesting2001], more recent studies have identified exceptions to pleomerism rule in moray eels [@reeceEvolutionary2013], large species of snakes [@headDissociation2007;@lindellEvolution1994], and salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].
+
+The lack of pleomerism in some species suggests it is possible to disassociate segment production with body growth, and thus reveals the developmental processes underlying evolutionary patterns.
+Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in many snakes [@alexanderPattern1966] and so ventral scales have often been used as a proxy for vertebrae number [@lindellEvolution1994;@shineVertebral2000].
+However, this 1:1 correlation is absent in file snakes,*Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not conform to pleomerism.
+Snakes are therefore a good model to study axial evolution because they have great diversity in vertebrae number and a wide range of adult body sizes and shapes but without a limb skeleton to confound signals of locomotor adaptation.
+Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattVertebral2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024].
+The number of vertebrae has also been shown to influence snake locomotion, whereby the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997].
+Despite these recent advances, there have been no study that have quantified the number of vertebrae across all the species in a diverse assemblage to test the prevalence of pleomerism in fossorial snakes.
 
@@
 Although all blindsnakes are fossorial, body size and shape vary significantly among species [@tiatragulMorphological2024; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993].
-Leptotyphlopids are generally thought to be more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@herrelBurrowing2021].
-These differences in body size and proportions determine how much push force the snakes can generate when burrowing.
-Their morphological differences led to the idea that leptyphlopids generally use existing burrows while typhlopids create their own burrow.
+Leptotyphlopids are generally considered more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@hedgesLower2008;@webbDietary2001;@webbPreysize1993;@iturriagaNew2021;@tiatragulMorphological2024].
+These differences in body size and proportions determine how much push force the snakes can generate when burrowing, which led to the idea that leptyphlopids generally use existing burrows while typhlopids create their own burrow [@herrelBurrowing2021].
 
-<!-- In another predominantly burrowing clade, caecilians, the number of vertebrae is not always correlated with body size [@renousBody1989], suggesting pleomerism is a mechanism for burrowing taxa to diversify in the underground environment.
+The relationship between segmentation and body elongation in scolecophidians is not well understood.
+However, a study by Head and Polly [-@headDissociation2007] that included 49 species across two genera typhlopids showed that vertebrae number and body size are positively correlated except when some of the largest species were included in the analysis.
+In particular, the largest species of blindsnakes in the African continent (*Afrotyphlops schlegelii*) have more precloacal vertebrae than expected for their size [@headDissociation2007].
+This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae.
+Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001], which is reflected in the relatively short tails of blindsnakes [@tiatragulMorphological2024;@szyndlarIllustrated2023].
+Therefore, blindsnakes are a good system to test how changes in the number of pre-cloacal vertebrae affects maximum body length.
+
+<!-- Limited studies on scolecophidian axial skeletons compared to alethinophidia means we have yet to appreciate the diversity that these groups have to offer.
+Given the recent shift in the perception of these earliest diverging clades of snakes (scolecophidians) away from the traditional view that they are "plesiomorphic and morphologically homogenous" [@strongConvergence2022], the scolecophidian axial skeleton also could offer a new perspective on the ecological origin of snakes.
 -->
+<!-- In another predominantly burrowing limbless clade, caecilian amphibians demonstrate a lack of pleomerism [@renousBody1989].  -->
+<!-- Burrowing amphisbaenian lizards and some typhlopid blindsnakes have more ventral scales than vertebrae [@alexanderPattern1966] suggesting pleomerism. -->
 
-Pleomerism has been observed in typhlopidae, however, some notable exceptions indicate different mechanisms for gigantism [@headDissociation2007].
-For instance, Head and Polly [-@headDissociation2007] found that some of the largest species of blindsnakes in the African continent (e.g.
-*Afrotyphlops schlegelii*) have fewer than expected precloacal vertebrae than expected for their size.
-This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae.
-Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001].
-This is reflected in the relatively short tails of blindsnakes [@tiatragulMorphological2024;@szyndlarIllustrated2023].
-Limited studies on scolecophidian axial skeletons compared to alethinophidia means we have yet to appreciate the diversity that these groups have to offer.
-Given the recent shift in the perception of these earliest diverging clades of snakes (scolecophidians) away from the traditional view that they are "plesiomorphic and morphologically homogenous" [@strongConvergence2022], the scolecophidian axial skeleton also could offer a new perspective on the ecological origin of snakes.
-
-The Australian blindsnake radiation comprise 47 described species [@tiatragulAustralia2023] distributed across all major biomes of Australia [@tiatragulPaleoenvironmental2023] with considerably diverse morphology from narrow-and-long (aspect ratio >= 100) to robust-and-wide (aspect ratio < 60).
+The Australian blindsnake radiation, *Anilios*, comprise 47 described species distributed across all major biomes of Australia [@tiatragulAustralia2023;@tiatragulPaleoenvironmental2023].
+Species in the genus show considerably diverse morphology from narrow-and-long to robust-and-wide [@tiatragulMorphological2024;@ehmannFamily1993].
 However, information about the relationship between body size and vertebrae number is unknown.
-In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes.
+In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses about how deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes.
 The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes?
-(2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes?
-(3) Does vertebral ratio vary with soil compactness or mean annual temperature?
+(2) Can deviations from pleomerism be explained by body shape in Australian blindsnakes?
+(3) Does vertebral ratio vary with their ecology (e.g., soil compactness and mean annual temperature)?
 
@@
 
-We extracted 37 lineages of Australian blindsnakes from the time-calibrated phylogeny of Australian blindsnakes and outgroups presented in [@tiatragulAustralia2023].
+We extracted 37 lineages of Australian blindsnakes from the time-calibrated phylogeny of Australian blindsnakes and outgroups presented in @tiatragulAustralia2023.
 The subset phylogeny represents 35/47 formally described species of _Anilios_ along with two additional genetically distinct lineages: one eastern population of "_A.
@@
 We collected body size measurements and vertebral counts from museum specimens examined.
-To quantify body size and body width ratios, we measured snout-to-vent length (SVL), tail lengths (TL) and midbody diameter (MBD).
-We measured SVL and TL by stretching the specimen along a ruler to the nearest 0.5 mm.
-For MBD, we used a digital vernier calliper (ABS Coolant Proof Caliper, Mitutoyo, Japan) to the nearest 0.1 mm.
-Total body length (TBL) is the sum of SVL + TL.
-<!-- To obtain  mean body shape ratios [@mosimannSize1970] for each species, we divided each MBD by SVL and log-transformed for further analyses. -->
+Body measurements including snout-to-vent length (SVL), tail lengths (TL) and midbody diameter (MBD) were obtained from @tiatragulMorphological2024.
+SVL and TL were measured to the nearest 1 mm on a ruler.
+MBD was measured using a digital vernier calliper (ABS Coolant Proof Caliper, Mitutoyo, Japan) to the nearest 0.1 mm.
+Total body length (TBL) is the sum of SVL and TL.
+To quantify body shape for each specimen, we calculated the aspect ratio [also referred to as elongation ratio; *sensu* @wardConvergent2004], which is the quotient of TBL and MBD.
 
-To obtain vertebral counts, we X-ray scanned the specimens at various institutions including: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n=87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n=129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (n=90), and Queensland X-Ray (n=35).
+To obtain vertebral counts, we X-ray scanned the specimens on the following machines and institutions: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n=87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n=129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (n=90), and Queensland X-Ray (n=35).
 
@@
 Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023].
-We calculated vertebrae ratio by dividing the total number of vertebrae by total body length.
+We calculated vertebrae ratio by dividing the total number of vertebrae by maximum TBL.
 
@@
 Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra.
-We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R Statistical Environment v.4.3.1 [@rcoreteamLanguage2023].
+We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R Statistical Environment v.4.4.0 [@rcoreteamLanguage2024].
 Measurements were repeated three times to account for measurement errors.
@@
 
-From raster files that contained modelled approximations at the resolution of 3 arc-seconds, we extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html), mean annual temperature from WorldClim v.2 [@fickWorldClim2017] for each georeferenced record using the "extract" function in the R package *`raster`* v.3.6-3 [@hijmansRaster2022].
+We extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html) and mean annual temperature from WorldClim v.2 [@fickWorldClim2017] from raster files that contained modelled approximations at the resolution of 3 arc-seconds.
+For each georeferenced record using the "extract" function in the R package *`raster`* v.3.6-3 [@hijmansRaster2022].
 We then calculated the species maximum soil bulk density and species mean annual temperature.
@@
 
-All statistical analyses were performed in R Statistical Environment.
+All statistical analyses were performed in R Statistical Environment v 4.4.0 [@rcoreteamLanguage2024].
 To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in the R package *`geomorph`* v.4.0.6 [@adamsGeomorph2023].
 First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes.
-We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum total body length from @tiatragulMorphological2024.
-Next, we test whether the vertebrae to body length ratio is related to the mean vertebrae number and mean aspect ratio, defined by total body length / body width, also from @tiatragulMorphological2024.
-Finally, we tested if variation in vertebrae ratio and vertebrae number can be explained by mean annual temperature and soil compactness.
+We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum TBL.
+Next, we tested whether vertebrae ratio is related to the mean vertebrae number and mean aspect ratio.
+Finally, we tested if variation in vertebrae ratio and vertebrae number is correlated with maximum soil bulk density and mean annual temperature.
 In all PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations.
 To visualise the relationship between vertebrae number and intracolumnar vertebral length, for each of the representative species, we plotted the mean vertebrae size (from three repeated measurements) against the vertebra number.
-Each vertebrae profile was overlaid with a fitted 4° polynomial curve.
+Each vertebrae profile was overlaid with a fitted 4° polynomial curve [e.g., @sherrattMacroevolution2022].
 
@@
 
-Total number of vertebrae varies among the 39 species of Australo-Papuan blindsnakes in this study (Fig.
+Total number of vertebrae varies greatly among the 39 species of Australo-Papuan blindsnakes in this study (Fig.
 \@ref(fig:fig-tree-boxplots)), ranging from 146 in *A.
@@
 grypus* (**Table S1**).
-Aspect ratios ranged from 20.8 in the robust-and-wide *A.
+Aspect ratios also vary greatly, ranging from 20.8 in the robust-and-wide *A.
 pinguis* to 169.8 in the narrow-and-long *A.
 longissimus* (**Table S2**).
+Intraspecific variation is also present in both aspect ratio and total number of vertebrae (Fig.
+\@ref(fig:fig-tree-boxplots)).
 
 ```{r fig-tree-boxplots, fig.cap= 'Interspecific variation in aspect ratio and total number of vertebrae among Australo-Papuan blindsnake and close relatives.
-The boxplots represent the distribution of data, with vertical inside representing the median.
+The boxplots represent the distribution of data, with vertical thick line inside representing the median.
 The extent of the box represents the lower (left) and upper (right) quartile, with the whiskers showing 1.5 times the interquartile range.
@@
 
-For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig.
+For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum TBL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig.
 \@ref(fig:fig-tbl-vert) and **Fig S1**).
@@
 grypus* "West" is an intermediate length (relative to other species) but has the greatest number of vertebrae, while *A.
-torresianus*, which is the longest, has relatively few vertebrae.
+torresianus*, which has the greatest maximum TBL, has relatively few vertebrae.
 
 ### Relationships between vertebrae ratio, number, and aspect ratio
-We found that vertebrae ratio is positively correlated with aspect ratio (PGLS, slope=-0.0048, intercept=0.73, R^2^=0.17, F=7.09, Z=2.20, P=0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig.
+We found that vertebrae ratio is positively correlated with mean aspect ratio (PGLS, slope=-0.0048, intercept=0.73, R^2^=0.17, F=7.09, Z=2.20, P=0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig.
 \@ref(fig:fig-vert-aspect-ratios)).
@@
 <!-- Fig. Vertebral counts vs max body length  -->
-```{r fig-tbl-vert, fig.cap= 'Maximum vertebral counts plotted against maximum total body length for Australian typhlopids.
-Trend line represent PGLS regression for Australian species.
+```{r fig-tbl-vert, fig.cap= 'Maximum vertebral counts plotted against maximum total body length (TBL) for Australian typhlopids.
+The trend line represents PGLS regression for Australian species.
 Phylogeny projected into space using maximum likelihood ancestral state estimation.
@@
 ```{r fig-vert-aspect-ratios, fig.cap= 'Species mean aspect ratios against (A) mean vertebrae ratio and (B) mean number of vertebrae of Australian blindsnakes.
-Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter.
+Vertebrae ratio is calculated by total vertebrae/TBL, and aspect ratio is TBL/MBD.
 (C) Vertebral profile shape of three species representing different vertebrae ratios, and are depicted as a 4° polynomial curve with positive inflection.
@@
 \@ref(fig:fig-vert-ecology)A).
-We also found a significant negative relationship between vertebrae ratio and species maximum soil bulk density (PGLS, slope=-1.00, y=2.52, R^2^=0.12, F=4.8, Z=1.7, P=0.043), indicating species that have greater number of vertebrae for body size are generally found in places that have harder to penetrate soil (Fig.
+We also found a significant negative relationship between vertebrae ratio and species maximum soil bulk density (PGLS, slope=-1.00, y=2.52, R^2^=0.12, F=4.8, Z=1.7, P=0.043), indicating species that have greater number of vertebrae for body size are generally found in places that have  harder-to-penetrate soil (Fig.
 \@ref(fig:fig-vert-ecology)B).
@@
 Mean vertebrae ratio against (A) species mean annual temperature and (B) species max soil bulk density.
-Mean vertebrae ratio is calculated by mean total vertebrae / mean total length for each species.
+Mean vertebrae ratio is calculated by mean total vertebrae / mean TBL for each species.
 Mean vertebrae number against (C) species mean annual temperature and (D) species max soil bulk density.
@@
 
-Our study investigated the presence of pleomerism in Australian blindsnakes and test whether deviations from pleomerism may be explained by body shape or ecological factors.
-Our results revealed that pleomerism does not generally apply to Australian typhlopids, indicating a deviation from the positive correlation between body size and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975].
-Variation in body shape, however, is underpinned by the ratio between vertebrae number to body size.
+Our study investigated the presence of pleomerism in Australian blindsnakes and tested whether deviations from pleomerism may be explained by body shape or ecological factors.
+Our results revealed that pleomerism does not generally apply to Australian typhlopids, indicating a deviation from the positive correlation between maximum body length and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975].
+Variation in body shape, however, is related to vertebrae ratio.
 We show that species that are narrow-and-long have proportionally more vertebrae compared to species that are more robust-and-wide.
@@
 
-We found a dissociation between vertebrae number and body size in Australian blindsnake.
-While pleomerism has been documented in most snake groups [@lindellEvolution1994], exceptionally large species ("giant taxa") within boids, pythonids, and blindsnakes such as *Afrotyphlops schlegelii* have been shown to have fewer than expected pre-cloacal vertebrae for their size [@headDissociation2007].
+We found a dissociation between vertebrae number and maximum adult body length in Australian blindsnake.
+While pleomerism has been documented in most snake groups [@lindellEvolution1994], exceptionally large species ("giant taxa") within boids, pythonids, and blindsnakes such as *Afrotyphlops schlegelii* have fewer than expected pre-cloacal vertebrae for their size [@headDissociation2007].
 In line with these observations, some of the largest species in our dataset (*A.
@@
 \@ref(fig:fig-tbl-vert)).
-The absence of pleomerism suggest that the large size attained by these "giant taxa" must be due to modification of post-somitogenetic somatic growth [@headDissociation2007].
+The absence of pleomerism suggest that the large size attained by these "giant taxa" may be due to modification of post-somitogenetic somatic growth [@headDissociation2007].
 In other words, species with larger body size do not simply have more vertebrae segments, rather the size of each vertebral segment has been elongated during growth.
 Several mechanisms have been suggested for how giant amniotes achieve their large size including prolonging their growth [@ericksonHow1999] and increasing the rate of growth [@ericksonGigantism2004].
-Unfortunately, without life-history data for blindsnakes, we cannot determine which of these mechanisms are responsible.
+Unfortunately, without life-history data for blindsnakes, we cannot test if these mechanisms may be responsible.
 However, elongation may also be achieved in other ways.
-The regionalisation of vertebrae, where distinct regions of the vertebrae have different evolutionary pathways and morphological diversification [@hamptonRegionalization2023;@hamptonHeart2022], may explain the lack of pleomerism in blindsnakes.
-For example, the absence of pleomerism in *Hydrophis* sea snakes has been attributed to the increase in the number of smaller vertebrae in the pre-atrial region compared to larger vertebrae in the post-atrial region [@sherrattMacroevolution2022;@sherrattHeterochronic2019].
+The regionalisation of vertebrae, where distinct regions along the axial column exhibit morphological diversification, and evolve relatively independently, may explain the lack of pleomerism in blindsnakes.
+For example, the absence of pleomerism in *Hydrophis* sea snakes has been attributed to the increase in the number of smaller vertebrae in the forebody region compared to larger vertebrae in the hindbody region [@sherrattMacroevolution2022;@sherrattVertebral2019].
 Similarly, colubrids have also been shown to have different regionalisation patterns in the vertebral column that contributes to elongation in different ways [@nash-hahnModification2024].
-Another important aspect of regionalisation is the position of the heart, which has been shown to be a key marker for delimiting different regions of the vertebrae [@hamptonHeart2022;@hamptonRegionalization2023].
+The position of the heart is a key marker for delimiting the regions of the snake axial column [@hamptonHeart2022;@hamptonRegionalization2023].
 Future research should investigate the regionalisation of the vertebral column in blindsnakes as this will provide important insights to understanding the lack of pleomerism.
@@
 The lack of pleomerism among burrowing vertebrates is not uncommon.
-@lindellEvolution1994 found that among Australian elapids and African snakes, burrowing species had less vertebrae than expected from its body size compared to terrestrial and arboreal species.
+@lindellEvolution1994 found that among Australian elapids and African snakes, burrowing species had fewer vertebrae than expected from their body length compared to terrestrial and arboreal species.
 The dissociation between vertebrae number and body size has also been documented in caecilians, a primarily fossorial order of amphibians.
-In caecilians, @renousBody1989 found that members of the same genus will vary in size but generally share similar number of vertebrae indicating a lack of pleomerism except for in a few genera such as *Oscaecilia*.
+In caecilians, @renousBody1989 found that members of the same genus will vary in size but generally share similar number of vertebrae, indicating a lack of pleomerism except for in a few genera such as *Oscaecilia*.
 Another group of highly specialised burrowing reptile is the amphisbaenians [@gansCharacteristics1978;@gansAmphisbaeniansreptiles1969].
-While several studies have documented the number of vertebrae in some species [@bergmannVertebral2012;@gansStudies1960;@gansStudies1971;@gansCharacteristics1978], no study has investigated the correlation between vertebrae number and body size in amphisbaenians.
+While several studies have documented the number of vertebrae in some species [@bergmannVertebral2012;@gansStudies1960;@gansStudies1971;@gansCharacteristics1978;@alexanderPattern1966], no study has investigated the correlation between vertebrae number and body size in amphisbaenians.
+However, they have been shown to deviate from the expected 1:1 ratio of ventral scale number to vertebrae [@alexanderPattern1966], suggesting a lack of pleomerism.
 We recommend future research to comprehensively investigate the prevalence of pleomerism among burrowing limbless and limb-reduced vertebrates to improve our understanding of the evolution of axial system.
@@
 Aspect ratio (body shape) varied with the number of vertebrae in similar ways as previously reported in fish and other snakes (e.g.
-vipers and colubrids), where species with a more elongated body had relatively greater number of vertebrae than stouter ones [@lindseyPleomerism1975;@lindellEvolution1994].
-Two possible explanations have been proposed by @lindellEvolution1994 including adaptations for locomotion and prey consumption.
-@poughSpecializations1983 showed that in snakes, longer body segments may be advantageous for elongated species that usually move faster, whereas shorter segments are favoured for stouter species that move more slowly.
+vipers and colubrids), where species with a more elongated body had relatively more vertebrae than stouter ones [@lindseyPleomerism1975;@lindellEvolution1994;@wardAxial2010].
+Two possible explanations have been proposed by @lindellEvolution1994 for snakes including adaptations for locomotion and prey consumption, which are mirrored in the interpretations from elongate fish [@wardAxial2010].
+Regarding locomotion, @poughSpecializations1983 showed that in snakes, longer body segments may be advantageous for elongated species that usually move faster, whereas shorter segments are favoured for stouter species that move more slowly.
 In blindsnakes, however, movement speed may not be as important as how much push force they can generate when burrowing.
-Snakes with relatively wider bodies are generally able to produce greater push force for their body size [@herrelBurrowing2021], but the trade-off is species with narrower bodies will be able to use pre-existing small crevices and galleries without needing to expend any force.
-In regards to prey consumption, stout species tend to consume large bulky prey, which is facilitated by having shorter body segment [@poughSpecializations1983;@lindellEvolution1994;@moonFeeding2019].
+Snakes with relatively wider bodies are generally able to produce greater push force for their body size [@herrelBurrowing2021], while species with narrower bodies will be able to use pre-existing small crevices and galleries without needing to expend any energy.
+In regards to prey consumption, stout species with large heads (i.e.
+vipers) tend to consume large bulky prey, which could be facilitated by having a stout body as this minimises disruption of physiological functions [@poughSpecializations1983;@lindellEvolution1994;@moonFeeding2019].
 Although Australian blindsnakes primarily prey on pupae and larvae of ants and termites, there is some variation in the size of prey consumed [@webbDietary1993].
@@
 Previous studies across different species of rattlesnakes [@klauberRattlesnakes1982;@klauberCorrelation1941] and between populations of garter snakes (*Thamnophis elegans*) [@arnoldQuantitative1988] both reported a similar pattern where species that occupy warmer and dryer environments had more vertebrae than those living in colder and humid coastal areas.
-While the temperature embryos are exposed to during development have been shown to affect vertebral numbers in snakes [@foxMorphological1961;@foxEffect1948;@osgoodEffects1978], @arnoldQuantitative1988 suggested three reasons it is unlikely the explanation for the regional difference in the number of vertebrae observed in *T.
+While the temperature embryos are exposed to during development has been shown to affect vertebral numbers in snakes [@foxMorphological1961;@foxEffect1948;@osgoodEffects1978], @arnoldQuantitative1988 suggested three reasons it is unlikely the explanation for the regional difference in the number of vertebrae observed in *T.
 elegans*.
@@
 Finally, females can select microhabitats that buffer their developing embryos against extreme climatic conditions [@arnoldQuantitative1988;@brownMaternal2004;@burgerNest1986;@peet-pareFemale2012].
-Although we currently lack knowledge about egg-laying behaviour and how temperature affects the development of blindsnake embryos, we suspect that temperature effects alone is unlikely to explain the interspecific variation in vertebrae number.
+Although we currently lack knowledge about egg-laying behaviour and how temperature affects the development of blindsnake embryos, we suspect that temperature effects alone are unlikely to explain the interspecific variation in vertebrae number.
 
-In a related study, @tiatragulMorphological2024 found that Australian blindsnake species with smaller absolute body size (SVL) generally occupy areas that areas that warmer than larger species.
+In a related study, @tiatragulMorphological2024 found that Australian blindsnake species with smaller absolute body size generally occupy areas that are warmer regions of Australia than while larger species tend to occupy cooler regions.
 Here, we observe that vertebrae ratio (raw vertebrae number/SVL) is positively correlated with temperature, indicating species that have more vertebrae than expected from their body size are found in warmer places.
-In his work on rattlesnakes, @klauberRattlesnakes1982  noted that smaller-sized subspecies often retain the same number of vertebrae with their ancestors resulting in greater vertebrae ratio.
+In his work on rattlesnakes, @klauberRattlesnakes1982  noted that smaller-sized subspecies often retain the same number of vertebrae as their ancestors, which result in the smaller species having relatively greater vertebrae ratio.
 An explanation for this pattern is that size is often under greater selection than vertebral number, and there is negligible disadvantage for a snake in having too many vertebrae for their body size [@lindellEvolution1994].
 This makes sense when we also consider the negative relationship between vertebrae ratio and maximum soil bulk density.
-A greater number of vertebrae for body size may afford greater flexibility [@jayneKinematics1986;@bergmannVertebral2012] that enables these species to navigate in pre-existing burrows or termite galleries, typical across northern Australia, without exerting force to dig their own burrow.
-@tiatragulMorphological2024 also found species that occupy areas with more difficult to-penetrate soil are usually narrow-and-long rather than robust-and-wide species that.
+A greater number of (shorter) vertebrae for body size may afford greater flexibility [@jayneKinematics1986;@bergmannVertebral2012;@johnsonAdaptive1955] that enables these species to navigate in pre-existing burrows or termite galleries, typical across northern Australia, without exerting force to dig their own burrow.
+@tiatragulMorphological2024 also found species that occupy areas with more difficult-to-penetrate soil are usually narrow-and-long rather than robust-and-wide species that occupy areas with easier-to-penetrate soil.
 Narrow-and-wide phenotypes are common among leptotyphlopids, which are thought to use existing burrows rather than digging their own [@herrelBurrowing2021].
-In this regard, the Australian typhlopid genus, *Anilios*, is quite exceptional because there is a great variation in body shape [@tiatragulMorphological2024].
+In this regard, the Australian typhlopid genus *Anilios* is quite exceptional because there is great variation in body shape [@tiatragulMorphological2024].
 Species that evolved more narrow-and-long bodies may be converging on the behaviour with other scolecophidians that have similar phenotypes (i.e., leptotyphlopids).
-However, more detailed species-level microhabitat observations are required to determine if these species are actually taking advantage of these environments in the ways we speculated.
+However, more detailed species-level microhabitat observations are required to determine if these species are actually taking advantage of these environments in the ways we speculate.
 
@@
 
-<!-- (1) How prevalent is pleomerism among Australian blindsnakes?
-(2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes?
-(3) Does vertebral ratio vary with soil compactness or mean annual temperature?
--->
-
-Compared to other genera of blindsnakes, the Australo-Papuan genus *Anilios* show exceptional diversity in body size and shape [@tiatragulMorphological2024;@herrelBurrowing2021], which make them a good group to study the evolution of axial skeleton.
-We found a lack of pleomerism among the genus, with exceptionally large species possessing a smaller number of vertebrae than expected for their body size.
-We also found substantial variation in the number of vertebrae among species, and this variation is correlated in one way or another with ecological variables considered.
-The number of vertebrae is also correlated with body shape with species that are narrow-and-long possessing more vertebrae than expected for body size compared to species that are robust-and-wide.
+Compared to other genera of blindsnakes, the Australo-Papuan genus *Anilios* show exceptional diversity in body size and shape [@tiatragulMorphological2024;@herrelBurrowing2021], which make them a good group to study evolution of the axial skeleton.
+We found a lack of pleomerism in the genus, with exceptionally large species possessing a smaller number of vertebrae than expected for their body size.
+Our findings indicate that body size and shape evolution in this clade has been facilitated by changes to development, whereby the production of vertebral segments during early development is dissociated from somatic growth, likely driven by ecological adaptations.
 This study adds to a growing body of literature on the axial skeleton of snakes, a field that has largely been neglected.
-
-<!-- Blindsnakes are generally small enough to fit in an X-ray scanner, even in portable field X-ray machines.
-Collecting more data on this group of basal snakes can help us understand how vertebral morphology evolved in snakes.
--->
 

### 2024-10-10 manuscript.Rmd
including feedback from Scott

@@
 Elongate animals are expected to show a positive correlation between maximum adult body length and number of vertebrae, known as pleomerism [@lindseyPleomerism1975].
-While pleomerism was considered the norm [@lindseyPleomerism1975;@wakeComparative1966;@pollyTesting2001], more recent studies have identified exceptions to pleomerism rule in moray eels [@reeceEvolutionary2013], large species of snakes [@headDissociation2007;@lindellEvolution1994], and salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].
+While pleomerism was considered the norm [@lindseyPleomerism1975;@wakeComparative1966;@pollyTesting2001], more recent studies have identified exceptions to the pleomerism rule in moray eels [@reeceEvolutionary2013], large species of snakes [@headDissociation2007;@lindellEvolution1994], and salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].
 
@@
 Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in many snakes [@alexanderPattern1966] and so ventral scales have often been used as a proxy for vertebrae number [@lindellEvolution1994;@shineVertebral2000].
-However, this 1:1 correlation is absent in file snakes,*Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not conform to pleomerism.
+However, this 1:1 correlation is absent in file snakes, *Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not conform to pleomerism.
 Snakes are therefore a good model to study axial evolution because they have great diversity in vertebrae number and a wide range of adult body sizes and shapes but without a limb skeleton to confound signals of locomotor adaptation.
@@
 The number of vertebrae has also been shown to influence snake locomotion, whereby the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997].
-Despite these recent advances, there have been no study that have quantified the number of vertebrae across all the species in a diverse assemblage to test the prevalence of pleomerism in fossorial snakes.
+Despite these recent advances, there have been no study that have quantified the number of vertebrae across all the species in a diverse assemblage of fossorial snakes to test for the variation and potential ecological drivers.
 
@@
 Leptotyphlopids are generally considered more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@hedgesLower2008;@webbDietary2001;@webbPreysize1993;@iturriagaNew2021;@tiatragulMorphological2024].
-These differences in body size and proportions determine how much push force the snakes can generate when burrowing, which led to the idea that leptyphlopids generally use existing burrows while typhlopids create their own burrow [@herrelBurrowing2021].
+These differences in body size and proportions determine how much push force the snakes can generate when burrowing, which led to the idea that leptotyphlopids generally use existing burrows while typhlopids create their own burrow [@herrelBurrowing2021].
 
@@
 Total body length (TBL) is the sum of SVL and TL.
-To quantify body shape for each specimen, we calculated the aspect ratio [also referred to as elongation ratio; *sensu* @wardConvergent2004], which is the quotient of TBL and MBD.
+To quantify body shape for each specimen, we calculated the aspect ratio — also referred to as elongation ratio [@wardConvergent2004] — which is the quotient of TBL and MBD.
 
@@
 We used the Fiji distribution of ImageJ2 v.1.54j [@schindelinFiji2012;@ruedenImageJ22017] to digitally visualise X-ray images and counted the number of vertebrae using the Cell Counter module. 
-Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023].
-We calculated vertebrae ratio by dividing the total number of vertebrae by maximum TBL.
+<!-- Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023].  -->
+For each specimen we calculated vertebrae ratio by dividing the total number of vertebrae by TBL.
 
@@
 First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes.
-We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum TBL.
-Next, we tested whether vertebrae ratio is related to the mean vertebrae number and mean aspect ratio.
+We did this by regressing the species maximum number of total vertebrae against maximum TBL.
+Next, we tested whether the species mean aspect ratio is correlated with mean vertebrae ratio (mean vertebrae/mean TBL) and/or mean vertebrae number.
 Finally, we tested if variation in vertebrae ratio and vertebrae number is correlated with maximum soil bulk density and mean annual temperature.
@@
 Here, we observe that vertebrae ratio (raw vertebrae number/SVL) is positively correlated with temperature, indicating species that have more vertebrae than expected from their body size are found in warmer places.
-In his work on rattlesnakes, @klauberRattlesnakes1982  noted that smaller-sized subspecies often retain the same number of vertebrae as their ancestors, which result in the smaller species having relatively greater vertebrae ratio.
+In his work on rattlesnakes, @klauberRattlesnakes1982  noted that smaller-sized subspecies often retain the same number of vertebrae as their larger closest relative, which result in the smaller subspecies having relatively greater vertebrae ratio.
 An explanation for this pattern is that size is often under greater selection than vertebral number, and there is negligible disadvantage for a snake in having too many vertebrae for their body size [@lindellEvolution1994].

### 2024-10-11 manuscript.Rmd
update before sending to co-authors

@@
 # title: "Dissociation between number of vertebrae and body size in Australian blindsnakes"
-title: "Testing the correlates of huge vertebrae number disparity in Australian blindsnakes"
+title: "Testing the correlates of vast vertebrae number disparity in Australian blindsnakes"
 date: "Version `r Sys.Date()`"
@@
 We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebrae ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species.
-Finally, we found that species that have a greater number of vertebrae for their size are generally found in warmer environments and environments with harder-to-penetrate soil substrate.
+We also found that species that have a greater number of vertebrae for their size are generally found in warmer environments and environments with harder-to-penetrate soil substrate.
 Our results extend our understanding of the axial evolution of basal snakes and the macroevolutionary dynamics of Australian blindsnakes.
@@
 The number of vertebrae has also been shown to influence snake locomotion, whereby the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997].
-Despite these recent advances, there have been no study that have quantified the number of vertebrae across all the species in a diverse assemblage of fossorial snakes to test for the variation and potential ecological drivers.
+Despite these recent advances, there have been no studies that have quantified the number of vertebrae across all the species in a diverse assemblage of fossorial snakes to test for the variation and potential ecological drivers.
 
@@
 The relationship between segmentation and body elongation in scolecophidians is not well understood.
-However, a study by Head and Polly [-@headDissociation2007] that included 49 species across two genera typhlopids showed that vertebrae number and body size are positively correlated except when some of the largest species were included in the analysis.
-In particular, the largest species of blindsnakes in the African continent (*Afrotyphlops schlegelii*) have more precloacal vertebrae than expected for their size [@headDissociation2007].
+However, a study by Head and Polly [-@headDissociation2007] that included 49 species across two typhlopids genera showed that vertebrae number and body size are positively correlated except when some of the largest species were included in the analysis.
+In particular, the largest species of blindsnakes on the African continent (*Afrotyphlops schlegelii*) have more precloacal vertebrae than expected for their size [@headDissociation2007].
 This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae.
@@
 The Australian blindsnake radiation, *Anilios*, comprise 47 described species distributed across all major biomes of Australia [@tiatragulAustralia2023;@tiatragulPaleoenvironmental2023].
-Species in the genus show considerably diverse morphology from narrow-and-long to robust-and-wide [@tiatragulMorphological2024;@ehmannFamily1993].
+This radiation displays considerable morphological diversity which ranges from narrow-and-long to robust-and-wide species [@tiatragulMorphological2024;@ehmannFamily1993].
 However, information about the relationship between body size and vertebrae number is unknown.
@@
 \@ref(fig:fig-tbl-vert)).
-The absence of pleomerism suggest that the large size attained by these "giant taxa" may be due to modification of post-somitogenetic somatic growth [@headDissociation2007].
+The absence of pleomerism suggests that the large size attained by these "giant taxa" may be due to modification of post-somitogenetic somatic growth [@headDissociation2007].
 In other words, species with larger body size do not simply have more vertebrae segments, rather the size of each vertebral segment has been elongated during growth.
@@
 In caecilians, @renousBody1989 found that members of the same genus will vary in size but generally share similar number of vertebrae, indicating a lack of pleomerism except for in a few genera such as *Oscaecilia*.
-Another group of highly specialised burrowing reptile is the amphisbaenians [@gansCharacteristics1978;@gansAmphisbaeniansreptiles1969].
-While several studies have documented the number of vertebrae in some species [@bergmannVertebral2012;@gansStudies1960;@gansStudies1971;@gansCharacteristics1978;@alexanderPattern1966], no study has investigated the correlation between vertebrae number and body size in amphisbaenians.
+While several studies have documented the number of vertebrae in the highly specialised burrowing amphisbaenianss [@bergmannVertebral2012; @gansAmphisbaeniansreptiles1969; @gansStudies1960;@gansStudies1971;@gansCharacteristics1978;@alexanderPattern1966], no study has investigated the correlation between vertebrae number and body size this group.
 However, they have been shown to deviate from the expected 1:1 ratio of ventral scale number to vertebrae [@alexanderPattern1966], suggesting a lack of pleomerism.
@@
 Snakes with relatively wider bodies are generally able to produce greater push force for their body size [@herrelBurrowing2021], while species with narrower bodies will be able to use pre-existing small crevices and galleries without needing to expend any energy.
-In regards to prey consumption, stout species with large heads (i.e.
-vipers) tend to consume large bulky prey, which could be facilitated by having a stout body as this minimises disruption of physiological functions [@poughSpecializations1983;@lindellEvolution1994;@moonFeeding2019].
-Although Australian blindsnakes primarily prey on pupae and larvae of ants and termites, there is some variation in the size of prey consumed [@webbDietary1993].
-In particular, heavy-bodied blindsnake species (e.g.
+In regard to prey consumption, @poughSpecializations1983 suggest that stout species with large heads (i.e.
+vipers) tend to consume large bulky prey, which could be facilitated by having a stout body as this minimises disruption of physiological functions.
+However, Australian blindsnakes primarily prey on pupae and larvae of ants and termites, which are relatively small [@webbDietary1993].
+As such, we do not expect prey consumption to influence the relationhship between aspect ratio and vertebrae we observed.
+<!-- In particular, heavy-bodied blindsnake species (e.g.
 *A.
@@
 waitii*, *A.
-weidii*) [@webbDietary1993].
+weidii*).
+-->
 

### 2024-10-11 code/03_visualising_data.R
change comment minor

@@ -398,7 +398,7 @@ summary(sig_ver_rati)
 
 # Question 7 --------------------------------------------------------------
 
-## Correlation between maximum precloacal vertebrae positively correlates with body size
+## Correlation between total vertebrae number positively correlates with body size
 
 anilios_data$max_vert <- anilios_vert$max_total_vert
 anilios_data$mean_vert <- anilios_vert$mean_tot_vert

### 2024-10-25 manuscript.Rmd
add acknowledgement section
incorporate language correction from Andrew Amey.

@@
 ST: Conceptualisation, Methodology, Formal analysis, Writing - Original draft, reviewing and editing.
-AM: Resources, Investigation
+AA: Resources, Investigation
 SH: Resources, Investigation
@@
 <!-- ST and ES conceptualised the study; ST collected morphology data -->
-<!-- ST obtained X-rays from SAM, WAM, and MAGNT. SH and AM provided X-rays Queensland Museum specimens. -->
+<!-- ST obtained X-rays from SAM, WAM, and MAGNT. SH and AA provided X-rays Queensland Museum specimens. -->
 <!-- ST and ES analysed the data; ST prepared the figures; ST, SK and ES wrote the manuscript. -->
@@
 South Australian Museum, Western Australian Museum, Queensland Museum, Australian National Wildlife Collection, Museum and Art Gallery of the Northern Territory, Museums Victoria, and the Florida Museum of Natural History.
+SH and AA thank Queensland X-ray for provision of imaging for Queensland Museum specimens.
 SK thanks the Australian Research Council for ongoing support.
@@
 Changes in the vertebral skeleton can influence body form and modes of locomotion.
-Elongate animals are commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism.
-However, some groups of vertebrates show a lack of pleomerism, suggesting it is possible to dissociate the production of segment and somatic growth, but what factors influenced this pattern is not well understood.
+Elongate animals commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism.
+However, some groups of vertebrates show a lack of pleomerism, suggesting it is possible to dissociate the production of segment and somatic growth, but what factors influence this pattern is not well understood.
 Snakes offer a great model to study axial evolution because the number of vertebrae vary greatly between species and they exhibit a wide range of body size and shape.
-Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology can explain the variation in vertebrae number.
+Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology can explain the variation in the number of vertebrae.
 We obtained X-ray scans and morphometric data for 42 described species of *Anilios* blindsnakes and evaluated their correlation with each other and ecological variables.
-We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebrae ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species.
+We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebra ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species.
 We also found that species that have a greater number of vertebrae for their size are generally found in warmer environments and environments with harder-to-penetrate soil substrate.
@@
 Axial skeleton evolution is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised.
-Essentially vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryogenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009].
+Essentially, vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryogenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009].
 Elongate animals are expected to show a positive correlation between maximum adult body length and number of vertebrae, known as pleomerism [@lindseyPleomerism1975].
-While pleomerism was considered the norm [@lindseyPleomerism1975;@wakeComparative1966;@pollyTesting2001], more recent studies have identified exceptions to the pleomerism rule in moray eels [@reeceEvolutionary2013], large species of snakes [@headDissociation2007;@lindellEvolution1994], and salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].
+While pleomerism was considered the norm [@lindseyPleomerism1975;@wakeComparative1966;@pollyTesting2001], more recent studies have identified exceptions to the pleomerism rule in moray eels [@reeceEvolutionary2013], large species of snakes [@headDissociation2007;@lindellEvolution1994], salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].
 
 The lack of pleomerism in some species suggests it is possible to disassociate segment production with body growth, and thus reveals the developmental processes underlying evolutionary patterns.
-Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in many snakes [@alexanderPattern1966] and so ventral scales have often been used as a proxy for vertebrae number [@lindellEvolution1994;@shineVertebral2000].
+Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in many snakes [@alexanderPattern1966] and so ventral scales have often been used as a proxy for number of vertebrae [@lindellEvolution1994;@shineVertebral2000].
 However, this 1:1 correlation is absent in file snakes, *Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not conform to pleomerism.
-Snakes are therefore a good model to study axial evolution because they have great diversity in vertebrae number and a wide range of adult body sizes and shapes but without a limb skeleton to confound signals of locomotor adaptation.
+Snakes are therefore a good model to study axial evolution because they have great diversity in number of vertebrae and a wide range of adult body sizes and shapes but without a limb skeleton to confound signals of locomotor adaptation.
 Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattVertebral2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024].
 The number of vertebrae has also been shown to influence snake locomotion, whereby the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997].
-Despite these recent advances, there have been no studies that have quantified the number of vertebrae across all the species in a diverse assemblage of fossorial snakes to test for the variation and potential ecological drivers.
+Despite these recent advances, there have been no studies that have quantified the number of vertebrae across all the species in a diverse assemblage of fossorial snakes to test for variation and potential ecological drivers.
 
@@
 The relationship between segmentation and body elongation in scolecophidians is not well understood.
-However, a study by Head and Polly [-@headDissociation2007] that included 49 species across two typhlopids genera showed that vertebrae number and body size are positively correlated except when some of the largest species were included in the analysis.
-In particular, the largest species of blindsnakes on the African continent (*Afrotyphlops schlegelii*) have more precloacal vertebrae than expected for their size [@headDissociation2007].
+However, a study by Head and Polly [-@headDissociation2007] that included 49 species across two typhlopid genera showed that the number of vertebrae and body size are positively correlated except when some of the largest species were included in the analysis.
+In particular, the largest species of blindsnakes on the African continent (i.e.
+*Afrotyphlops schlegelii* species complex) has more precloacal vertebrae than expected for their size [@headDissociation2007].
 This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae.
@@
 
-The Australian blindsnake radiation, *Anilios*, comprise 47 described species distributed across all major biomes of Australia [@tiatragulAustralia2023;@tiatragulPaleoenvironmental2023].
+The Australian blindsnake radiation *Anilios* comprises47 described species distributed across all major biomes of Australia [@tiatragulAustralia2023;@tiatragulPaleoenvironmental2023].
 This radiation displays considerable morphological diversity which ranges from narrow-and-long to robust-and-wide species [@tiatragulMorphological2024;@ehmannFamily1993].
-However, information about the relationship between body size and vertebrae number is unknown.
+However, information about the relationship between body size and number of vertebrae is lacking.
 In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses about how deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes.
@@
 (2) Can deviations from pleomerism be explained by body shape in Australian blindsnakes?
-(3) Does vertebral ratio vary with their ecology (e.g., soil compactness and mean annual temperature)?
+(3) Does vertebral ratio vary with ecology (e.g., soil compactness and mean annual temperature)?
 
@@
 
-We examined 341 specimens representing 44 lineages *Anilios* blindsnakes including 42/47 formally described species [@uetzQuarter2021].
-We additionally included two close relatives, *Ramphotyphlops multilineatus* and *Acutotyphlops subocularis* [@tiatragulAustralia2023].
-Our sample size per species ranged from 1 to 25 specimens (median=7 specimens).
-We sourced specimens from South Australian Museum (SAMA), Western Australian Museum (WAM), Queensland Museum (QM), Museum and Art Gallery of the Northern Territory (MAGNT), and a private collection.
+We examined 341 specimens representing 46 typhlopid lineages including 42 of 47 formally described *Anilios*, two genetically distinct undescribed *Anilios* [@tiatragulAustralia2023;uetzQuarter2021], and their close relatives *Ramphotyphlops multilineatus* (Schlegel, 1839) and *Acutotyphlops subocularis* (Waite, 1897).
+Our sample size per species ranged from 1 to 25 specimens (median = 7 specimens).
+We sourced specimens from the South Australian Museum (SAMA), Western Australian Museum (WAM), Queensland Museum (QM), Museum and Art Gallery of the Northern Territory (MAGNT), and a private collection.
 Where possible, we chose relatively large non-fragile specimens from both males and females with well-preserved heads for X-ray radiography.
-Sex for each specimen were determined by inspecting the gonads or assigned based on relative tail lengths following @tiatragulMorphological2024.
+Sex for each specimen was determined by inspecting the gonads or assigned based on relative tail lengths following @tiatragulMorphological2024.
 
@@
 We extracted 37 lineages of Australian blindsnakes from the time-calibrated phylogeny of Australian blindsnakes and outgroups presented in @tiatragulAustralia2023.
-The subset phylogeny represents 35/47 formally described species of _Anilios_ along with two additional genetically distinct lineages: one eastern population of "_A.
+The subset phylogeny represents 35 of 47 formally described species of _Anilios_ along with two additional genetically distinct lineages: one eastern population of "_A.
 grypus_" and one eastern population of "_A.
@@
 
-We collected body size measurements and vertebral counts from museum specimens examined.
+We collected body size measurements and vertebral counts from museum specimens (**Table S1**).
 Body measurements including snout-to-vent length (SVL), tail lengths (TL) and midbody diameter (MBD) were obtained from @tiatragulMorphological2024.
@@
 
-To obtain vertebral counts, we X-ray scanned the specimens on the following machines and institutions: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n=87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n=129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (n=90), and Queensland X-Ray (n=35).
+To obtain vertebral counts, we X-ray scanned the specimens on the following machines and institutions: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n  =  87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n = 129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (n = 90), and Queensland X-Ray (n = 35).
 
@@
 <!-- Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023].  -->
-For each specimen we calculated vertebrae ratio by dividing the total number of vertebrae by TBL.
+For each specimen we calculated vertebra ratio by dividing the total number of vertebrae by TBL.
 
@@
 
-We used the Australian blindsnake occurrence records presented in @tiatragulMorphological2024 including a total of 3898 verified records (min=1, median=23, mean=81, max=525 per species).
+We used the Australian blindsnake occurrence records presented in @tiatragulMorphological2024 including a total of 3898 verified records (min = 1, median = 23, mean = 81, max = 525 per species).
 These records were compiled from museum databases and verified human observations from Atlas of the Living Australia (ALA).
 
-We selected soil bulk density and annual mean temperature, as we are interested in the potential effects these may have on the distribution of species.
-Soil bulk density measures the penetrability of soil, which is a proxy for the type of microhabitat species is most likely living in as they are all strictly fossorial.
+We selected soil bulk density and annual mean temperature, as we were interested in the potential effects these may have on the distribution of species.
+Soil bulk density measures the penetrability of soil, which is a proxy for the type of microhabitat a species is most likely living in as they are all strictly fossorial.
 Mean annual temperature is a climate proxy and is commonly used to describe niche for squamate reptiles [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013].
-Temperature also directly influence multiple axes of reptile development [@nobleDevelopmental2018;@boothInfluence2006;@feinerDevelopmental2018], which influences vertebrae development [@martin-del-campoHox2018].
+Temperature also directly influences multiple axes of reptile development [@nobleDevelopmental2018;@boothInfluence2006;@feinerDevelopmental2018], which influences vertebra development [@martin-del-campoHox2018].
 Bulk density values > 1.6 g/cm<sup>3</sup> indicate soil that restrict penetration by plant roots, while values <0.5 indicate looser, more organic soil [@mckenzieSoil2002].
-
-We extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html) and mean annual temperature from WorldClim v.2 [@fickWorldClim2017] from raster files that contained modelled approximations at the resolution of 3 arc-seconds.
-For each georeferenced record using the "extract" function in the R package *`raster`* v.3.6-3 [@hijmansRaster2022].
+We extracted soil bulk density values between 0 — 5 cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html) and mean annual temperature from WorldClim v.2 [@fickWorldClim2017] raster files that contained modelled approximations at the resolution of 3 arc-seconds using the "extract" function in the R package *`raster`* v.3.6-3 [@hijmansRaster2022].
 We then calculated the species maximum soil bulk density and species mean annual temperature.
@@
 We did this by regressing the species maximum number of total vertebrae against maximum TBL.
-Next, we tested whether the species mean aspect ratio is correlated with mean vertebrae ratio (mean vertebrae/mean TBL) and/or mean vertebrae number.
-Finally, we tested if variation in vertebrae ratio and vertebrae number is correlated with maximum soil bulk density and mean annual temperature.
+Next, we tested whether the species mean aspect ratio is correlated with mean vertebra ratio (mean number of vertebrae/mean TBL) and mean number of vertebrae.
+Finally, we tested if variation in vertebra ratio and number of vertebrae is correlated with maximum soil bulk density and mean annual temperature.
 In all PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations.
-To visualise the relationship between vertebrae number and intracolumnar vertebral length, for each of the representative species, we plotted the mean vertebrae size (from three repeated measurements) against the vertebra number.
-Each vertebrae profile was overlaid with a fitted 4° polynomial curve [e.g., @sherrattMacroevolution2022].
+To visualise the relationship between number of vertebrae and intracolumnar vertebral length for each of the representative species, we plotted the mean vertebra size (from three repeated measurements) against the vertebra number.
+Each vertebra profile was overlaid with a fitted 4° polynomial curve [e.g., @sherrattMacroevolution2022].
 
@@
 
-Total number of vertebrae varies greatly among the 39 species of Australo-Papuan blindsnakes in this study (Fig.
+Total number of vertebrae varies greatly among the Australo-Papuan blindsnakes in this study (Fig.
 \@ref(fig:fig-tree-boxplots)), ranging from 146 in *A.
@@
 
-```{r fig-tree-boxplots, fig.cap= 'Interspecific variation in aspect ratio and total number of vertebrae among Australo-Papuan blindsnake and close relatives.
+```{r fig-tree-boxplots, fig.cap =  'Interspecific variation in aspect ratio and total number of vertebrae among Australo-Papuan blindsnakes and close relatives.
 The boxplots represent the distribution of data, with vertical thick line inside representing the median.
@@
 Outliers beyond this range are depicted as individual points.
-Some species are represented by only the mean.', out.width="165mm", out.height= "129mm"}
+Some species are represented by only a single data point.', out.width = "165mm", out.height =  "129mm"}
 
@@
 
-For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum TBL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig.
+For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum TBL (PGLS, slope = 0.28, intercept = 256, R^2^ = 0.02, F = 0.74, Z = 0.30, P = 0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig.
 \@ref(fig:fig-tbl-vert) and **Fig S1**).
@@
 
-### Relationships between vertebrae ratio, number, and aspect ratio
-We found that vertebrae ratio is positively correlated with mean aspect ratio (PGLS, slope=-0.0048, intercept=0.73, R^2^=0.17, F=7.09, Z=2.20, P=0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig.
+### Relationships between vertebra ratio, number of vertebrae, and aspect ratio
+We found that vertebra ratio is positively correlated with mean aspect ratio (PGLS, slope = -0.0048, intercept = 0.73, R^2^ = 0.17, F = 7.09, Z = 2.20, P = 0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig.
 \@ref(fig:fig-vert-aspect-ratios)).
-We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope=-1.8, intercept=120.95, R^2^=0.69, F=79.39, Z=5.30, P<0.001).
+We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope = -1.8, intercept = 120.95, R^2^ = 0.69, F = 79.39, Z = 5.30, P<0.001).
 
 <!-- Fig. Vertebral counts vs max body length  -->
-```{r fig-tbl-vert, fig.cap= 'Maximum vertebral counts plotted against maximum total body length (TBL) for Australian typhlopids.
+```{r fig-tbl-vert, fig.cap =  'Maximum vertebral counts plotted against maximum total body length (TBL) for Australian typhlopids.
 The trend line represents PGLS regression for Australian species.
@@
 Species not included in the phylogeny are shown in purple.
-', out.width="165mm", out.height= "124mm"}
+', out.width = "165mm", out.height =  "124mm"}
 
@@
 <!-- Fig: Pleomerism and ratio correlations -->
-```{r fig-vert-aspect-ratios, fig.cap= 'Species mean aspect ratios against (A) mean vertebrae ratio and (B) mean number of vertebrae of Australian blindsnakes.
-Vertebrae ratio is calculated by total vertebrae/TBL, and aspect ratio is TBL/MBD.
-(C) Vertebral profile shape of three species representing different vertebrae ratios, and are depicted as a 4° polynomial curve with positive inflection.
+```{r fig-vert-aspect-ratios, fig.cap =  'Species mean aspect ratios against (A) mean vertebra ratio and (B) mean number of vertebrae of Australian blindsnakes.
+Vertebra ratio is calculated by total vertebrae/TBL, and aspect ratio is TBL/MBD.
+Inset images of vertebral profile shape in (A) represent different vertebra ratios, and are depicted as a 4° polynomial curve with positive inflection.
 Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.
 Phylogeny projected into space using maximum likelihood ancestral state estimation.
-Species not included in the phylogeny are shown in purple.', out.width="124mm", out.height= "165mm"}
+Species not included in the phylogeny are shown in purple.', out.width = "124mm", out.height =  "165mm"}
 
@@
 <!-- Fig: Vertebral profile -->
-<!-- ```{r fig-vertebral-profile, fig.cap= 'Relative vertebral length plotted against vertebra number for three example species with fitted 4° polynomial curves overlaid.', out.width="140mm", out.height= "197mm"}
+<!-- ```{r fig-vertebral-profile, fig.cap =  'Relative vertebral length plotted against vertebra number for three example species with fitted 4° polynomial curves overlaid.', out.width = "140mm", out.height =  "197mm"}
 
@@
 
-### Correlation between vertebrae ratio, number, and ecological variables among Australian blindsnakes
+### Correlation between vertebra ratio, number of vertebrae, and ecology among Australian blindsnakes
 
-We found a positive relationship between vertebrae ratio and species mean annual temperature (PGLS, slope=0.038, y=0.19, R^2^=0.21, F=9.58, Z=2.32, P=0.009), indicating species that have greater number of vertebrae for body size are generally found in places that are warmer (Fig.
+We found a positive relationship between vertebra ratio and species mean annual temperature (PGLS, slope = 0.038, y = 0.19, R^2^ = 0.21, F = 9.58, Z = 2.32, P = 0.009), indicating species that have a greater number of vertebrae for body size are generally found in places that are warmer (Fig.
 \@ref(fig:fig-vert-ecology)A).
-We also found a significant negative relationship between vertebrae ratio and species maximum soil bulk density (PGLS, slope=-1.00, y=2.52, R^2^=0.12, F=4.8, Z=1.7, P=0.043), indicating species that have greater number of vertebrae for body size are generally found in places that have  harder-to-penetrate soil (Fig.
+We also found a significant negative relationship between vertebra ratio and species maximum soil bulk density (PGLS, slope = -1.00, y = 2.52, R^2^ = 0.12, F = 4.8, Z = 1.7, P = 0.043), indicating species that have a greater number of vertebrae for body size are generally found in places that have  harder-to-penetrate soil (Fig.
 \@ref(fig:fig-vert-ecology)B).
-When considering raw mean vertebrae number, we found non-statistically significant positive correlations with both mean annual temperature (PGLS, slope=4.13, y=142.45, R^2^=0.76, F=2.87, Z=1.31, P=0.09) and maximum soil bulk density (PGLS, slope=111.69, y=, R^2^=0.04, F=1.61, Z=0.85, P=0.22; Fig.
+When considering raw mean number of vertebrae, we found non-statistically significant positive correlations with both mean annual temperature (PGLS, slope = 4.13, y = 142.45, R^2^ = 0.76, F = 2.87, Z = 1.31, P = 0.09) and maximum soil bulk density (PGLS, slope = 111.69, y = 68.64, R^2^ = 0.04, F = 1.61, Z = 0.85, P = 0.22; Fig.
 \@ref(fig:fig-vert-ecology)C–D).
@@
 <!-- Fig: Ratio and ecological variables -->
-```{r fig-vert-ecology, fig.cap= 'Phylomorphospaces showing relationship between mean vertebrae ratios and mean vertebrae number against ecological variables.
-Mean vertebrae ratio against (A) species mean annual temperature and (B) species max soil bulk density.
-Mean vertebrae ratio is calculated by mean total vertebrae / mean TBL for each species.
-Mean vertebrae number against (C) species mean annual temperature and (D) species max soil bulk density.
+```{r fig-vert-ecology, fig.cap= 'Phylomorphospaces showing relationship between mean vertebra ratios and mean number of vertebrae against ecological variables.
+Mean vertebra ratio against (A) species mean annual temperature and (B) species max soil bulk density.
+Mean vertebra ratio is calculated by mean total vertebrae / mean TBL for each species.
+Mean number of vertebrae against (C) species mean annual temperature and (D) species max soil bulk density.
 Species mean annual temperature and max soil bulk density values are calculated based on occurrence data.
@@
 Our study investigated the presence of pleomerism in Australian blindsnakes and tested whether deviations from pleomerism may be explained by body shape or ecological factors.
-Our results revealed that pleomerism does not generally apply to Australian typhlopids, indicating a deviation from the positive correlation between maximum body length and vertebrae number reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975].
-Variation in body shape, however, is related to vertebrae ratio.
+Our results revealed that pleomerism does not generally apply to Australian typhlopids, indicating a deviation from the positive correlation between maximum body length and number of vertebrae reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975].
+Variation in body shape, however, is related to vertebra ratio.
 We show that species that are narrow-and-long have proportionally more vertebrae compared to species that are more robust-and-wide.
@@
 
-We found a dissociation between vertebrae number and maximum adult body length in Australian blindsnake.
+We found a dissociation between the number of vertebrae and maximum adult body length in Australian blindsnake.
 While pleomerism has been documented in most snake groups [@lindellEvolution1994], exceptionally large species ("giant taxa") within boids, pythonids, and blindsnakes such as *Afrotyphlops schlegelii* have fewer than expected pre-cloacal vertebrae for their size [@headDissociation2007].
@@
 The absence of pleomerism suggests that the large size attained by these "giant taxa" may be due to modification of post-somitogenetic somatic growth [@headDissociation2007].
-In other words, species with larger body size do not simply have more vertebrae segments, rather the size of each vertebral segment has been elongated during growth.
+In other words, species with larger body size do not simply have more vertebra segments, rather the size of each vertebral segment has been elongated during growth.
 Several mechanisms have been suggested for how giant amniotes achieve their large size including prolonging their growth [@ericksonHow1999] and increasing the rate of growth [@ericksonGigantism2004].
@@
 However, elongation may also be achieved in other ways.
-The regionalisation of vertebrae, where distinct regions along the axial column exhibit morphological diversification, and evolve relatively independently, may explain the lack of pleomerism in blindsnakes.
+The regionalisation of vertebrae, where distinct regions along the axial column exhibit morphological diversification and evolve relatively independently, may explain the lack of pleomerism in blindsnakes.
 For example, the absence of pleomerism in *Hydrophis* sea snakes has been attributed to the increase in the number of smaller vertebrae in the forebody region compared to larger vertebrae in the hindbody region [@sherrattMacroevolution2022;@sherrattVertebral2019].
@@
 
-The lack of pleomerism among burrowing vertebrates is not uncommon.
+Lack of pleomerism among burrowing vertebrates is not uncommon.
 @lindellEvolution1994 found that among Australian elapids and African snakes, burrowing species had fewer vertebrae than expected from their body length compared to terrestrial and arboreal species.
-The dissociation between vertebrae number and body size has also been documented in caecilians, a primarily fossorial order of amphibians.
+The dissociation between number of vertebrae and body size has also been documented in caecilians, a primarily fossorial order of amphibians.
 In caecilians, @renousBody1989 found that members of the same genus will vary in size but generally share similar number of vertebrae, indicating a lack of pleomerism except for in a few genera such as *Oscaecilia*.
-While several studies have documented the number of vertebrae in the highly specialised burrowing amphisbaenianss [@bergmannVertebral2012; @gansAmphisbaeniansreptiles1969; @gansStudies1960;@gansStudies1971;@gansCharacteristics1978;@alexanderPattern1966], no study has investigated the correlation between vertebrae number and body size this group.
+While several studies have documented the number of vertebrae in the highly specialised burrowing amphisbaenianss [@bergmannVertebral2012; @gansAmphisbaeniansreptiles1969; @gansStudies1960;@gansStudies1971;@gansCharacteristics1978;@alexanderPattern1966], no study has investigated the correlation between number of vertebrae and body size in this group.
 However, they have been shown to deviate from the expected 1:1 ratio of ventral scale number to vertebrae [@alexanderPattern1966], suggesting a lack of pleomerism.
@@
 
-### Correlation between vertebrae ratio, number, and aspect ratio
+### Correlation between vertebra ratio, number of vertebrae, and aspect ratio
 
@@
 However, Australian blindsnakes primarily prey on pupae and larvae of ants and termites, which are relatively small [@webbDietary1993].
-As such, we do not expect prey consumption to influence the relationhship between aspect ratio and vertebrae we observed.
+As such, we do not expect prey consumption to influence the relationship between aspect ratio and vertebrae we observed.
 <!-- In particular, heavy-bodied blindsnake species (e.g.
@@
 
-### Correlations between vertebrae ratio, number, and ecological variables
-<!-- Discuss/speculate why vertebrae ratio is correlated with annual mean temperature and soil bulk density]*** -->
+### Correlations between vertebra ratio, number of vertebrae, and ecological variables
+<!-- Discuss/speculate why vertebra ratio is correlated with annual mean temperature and soil bulk density]*** -->
 
-Both mean annual temperature and soil bulk density were significant predictors of vertebrae ratio but not for the raw number of vertebrae.
+Both mean annual temperature and soil bulk density were significant predictors of vertebra ratio but not for the raw number of vertebrae.
 While non-statistically significant, we observed a slightly positive relationship between raw number of vertebrae and mean annual temperature.
 Previous studies across different species of rattlesnakes [@klauberRattlesnakes1982;@klauberCorrelation1941] and between populations of garter snakes (*Thamnophis elegans*) [@arnoldQuantitative1988] both reported a similar pattern where species that occupy warmer and dryer environments had more vertebrae than those living in colder and humid coastal areas.
-While the temperature embryos are exposed to during development has been shown to affect vertebral numbers in snakes [@foxMorphological1961;@foxEffect1948;@osgoodEffects1978], @arnoldQuantitative1988 suggested three reasons it is unlikely the explanation for the regional difference in the number of vertebrae observed in *T.
+While the temperature embryos are exposed to during development has been shown to affect vertebral numbers in snakes [@foxMorphological1961;@foxEffect1948;@osgoodEffects1978], @arnoldQuantitative1988 suggested three reasons it is unlikely this is the explanation for the regional difference in the number of vertebrae observed in *T.
 elegans*.
-First, the observed difference in the number of vertebrae between populations is much greater than the effects of temperature produced lab experiments [@arnoldQuantitative1988;@osgoodEffects1978].
-Second, the number of vertebrae of the F1 hybrid is intermediate between inland and coastal parents [@arnoldQuantitative1988].
+First, the observed difference in the number of vertebrae between populations is much greater than the effects of temperature produced in lab experiments [@arnoldQuantitative1988;@osgoodEffects1978].
+Second, the number of vertebrae of an F1 hybrid is intermediate between inland and coastal parents, suggesting the number of vertebrae is influenced by genetics [@arnoldQuantitative1988].
 Finally, females can select microhabitats that buffer their developing embryos against extreme climatic conditions [@arnoldQuantitative1988;@brownMaternal2004;@burgerNest1986;@peet-pareFemale2012].
-Although we currently lack knowledge about egg-laying behaviour and how temperature affects the development of blindsnake embryos, we suspect that temperature effects alone are unlikely to explain the interspecific variation in vertebrae number.
+Although we currently lack knowledge about egg-laying behaviour and how temperature affects the development of blindsnake embryos, we suspect that temperature effects alone are unlikely to explain the interspecific variation in number of vertebrae.
 
-In a related study, @tiatragulMorphological2024 found that Australian blindsnake species with smaller absolute body size generally occupy areas that are warmer regions of Australia than while larger species tend to occupy cooler regions.
-Here, we observe that vertebrae ratio (raw vertebrae number/SVL) is positively correlated with temperature, indicating species that have more vertebrae than expected from their body size are found in warmer places.
-In his work on rattlesnakes, @klauberRattlesnakes1982  noted that smaller-sized subspecies often retain the same number of vertebrae as their larger closest relative, which result in the smaller subspecies having relatively greater vertebrae ratio.
-An explanation for this pattern is that size is often under greater selection than vertebral number, and there is negligible disadvantage for a snake in having too many vertebrae for their body size [@lindellEvolution1994].
-This makes sense when we also consider the negative relationship between vertebrae ratio and maximum soil bulk density.
+In a related study, @tiatragulMorphological2024 found that Australian blindsnake species with smaller absolute body size generally occupy warmer areas than larger species tend to occupy cooler regions.
+Here, we observe that vertebra ratio (raw number of vertebrae/SVL) is positively correlated with temperature, indicating species that have more vertebrae than expected from their body size are found in warmer places.
+In his work on rattlesnakes, @klauberRattlesnakes1982  noted that smaller-sized subspecies often retain the same number of vertebrae as their larger closest relative, which results in the smaller subspecies having a relatively greater vertebra ratio.
+An explanation for this pattern is that size is more likely to be under greater selection than vertebral number, and there is negligible disadvantage for a snake in having too many vertebrae for their body size [@lindellEvolution1994].
+This makes sense when we also consider the negative relationship between vertebra ratio and maximum soil bulk density.
 A greater number of (shorter) vertebrae for body size may afford greater flexibility [@jayneKinematics1986;@bergmannVertebral2012;@johnsonAdaptive1955] that enables these species to navigate in pre-existing burrows or termite galleries, typical across northern Australia, without exerting force to dig their own burrow.
 @tiatragulMorphological2024 also found species that occupy areas with more difficult-to-penetrate soil are usually narrow-and-long rather than robust-and-wide species that occupy areas with easier-to-penetrate soil.
-Narrow-and-wide phenotypes are common among leptotyphlopids, which are thought to use existing burrows rather than digging their own [@herrelBurrowing2021].
-In this regard, the Australian typhlopid genus *Anilios* is quite exceptional because there is great variation in body shape [@tiatragulMorphological2024].
+Narrow-and-long phenotypes are common among leptotyphlopids, which are thought to use existing burrows rather than digging their own [@herrelBurrowing2021].
+In this regard, the Australian typhlopid genus *Anilios* is quite exceptional among blindsnakes because there is great variation in body shape [@tiatragulMorphological2024].
 Species that evolved more narrow-and-long bodies may be converging on the behaviour with other scolecophidians that have similar phenotypes (i.e., leptotyphlopids).
@@
 Compared to other genera of blindsnakes, the Australo-Papuan genus *Anilios* show exceptional diversity in body size and shape [@tiatragulMorphological2024;@herrelBurrowing2021], which make them a good group to study evolution of the axial skeleton.
-We found a lack of pleomerism in the genus, with exceptionally large species possessing a smaller number of vertebrae than expected for their body size.
+We found a lack of pleomerism in the genus, with exceptionally large species possessing a lower number of vertebrae than expected for their body size.
 Our findings indicate that body size and shape evolution in this clade has been facilitated by changes to development, whereby the production of vertebral segments during early development is dissociated from somatic growth, likely driven by ecological adaptations.

### 2024-10-29 manuscript.Rmd
add phylogenetic signal

@@
 
-We examined 341 specimens representing 46 typhlopid lineages including 42 of 47 formally described *Anilios*, two genetically distinct undescribed *Anilios* [@tiatragulAustralia2023;uetzQuarter2021], and their close relatives *Ramphotyphlops multilineatus* (Schlegel, 1839) and *Acutotyphlops subocularis* (Waite, 1897).
-Our sample size per species ranged from 1 to 25 specimens (median = 7 specimens).
+We examined 349 specimens representing 48 typhlopid lineages including 44 *Anilios* [@tiatragulAustralia2023;uetzQuarter2021], *Ramphotyphlops multilineatus* (Schlegel, 1839), *R.
+acuticauda* (Peters, 1878), *R.
+depressus* (Peters, 1880), and *Acutotyphlops subocularis* (Waite, 1897) (**Table S1**).
+Our sample size per species ranged from 1 to 25 specimens (median = 6; mean = 7).
 We sourced specimens from the South Australian Museum (SAMA), Western Australian Museum (WAM), Queensland Museum (QM), Museum and Art Gallery of the Northern Territory (MAGNT), and a private collection.
@@
 
-We collected body size measurements and vertebral counts from museum specimens (**Table S1**).
+We collected body size measurements and vertebral counts from museum specimens.
 Body measurements including snout-to-vent length (SVL), tail lengths (TL) and midbody diameter (MBD) were obtained from @tiatragulMorphological2024.
@@
 We did this by regressing the species maximum number of total vertebrae against maximum TBL.
+We also calculated the degree of phylogenetic signal from these traits separately using the "physignal.z" function in `geomorph`.
 Next, we tested whether the species mean aspect ratio is correlated with mean vertebra ratio (mean number of vertebrae/mean TBL) and mean number of vertebrae.
 Finally, we tested if variation in vertebra ratio and number of vertebrae is correlated with maximum soil bulk density and mean annual temperature.
-In all PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations.
+In the phylogenetic signal and PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations.
 To visualise the relationship between number of vertebrae and intracolumnar vertebral length for each of the representative species, we plotted the mean vertebra size (from three repeated measurements) against the vertebra number.
@@
 tovelli* to 400 in *A.
-grypus* (**Table S1**).
+grypus* (**Table S2**).
 Aspect ratios also vary greatly, ranging from 20.8 in the robust-and-wide *A.
 pinguis* to 169.8 in the narrow-and-long *A.
-longissimus* (**Table S2**).
+longissimus* (**Table S3**).
 Intraspecific variation is also present in both aspect ratio and total number of vertebrae (Fig.
@@
 torresianus*, which has the greatest maximum TBL, has relatively few vertebrae.
+We found phylogenetic signal in both maximum number of total vertebrae (Z = 3.70, P = 0.001, K = 1.2, lambda = 1) and maximum TBL (Z = 3.02, P = 0.002, K = 1.0, lambda = 1).
 
@@
 In his work on rattlesnakes, @klauberRattlesnakes1982 noted that smaller-sized subspecies often retain the same number of vertebrae as their larger closest relative, which results in the smaller subspecies having a relatively greater vertebra ratio.
+We also observed maximum number of vertebrae in our dataset contain phylogenetic signal.
 An explanation for this pattern is that size is more likely to be under greater selection than vertebral number, and there is negligible disadvantage for a snake in having too many vertebrae for their body size [@lindellEvolution1994].

### 2024-10-29 code/00_stats_for_paper.R
update physignal

@@ -8,6 +8,7 @@
 
 library(dplyr)
 library(phytools); 
+library(stringr)
 library(geiger);
 library(ape); 
 library(geomorph); 
@@ -48,7 +49,7 @@ linear_df$species <- genus_species
 # Filter out rows with NA or 0 values in precloacal_vert
 vert_df <- linear_df[!is.na(linear_df$precloacal_vert) & 
                        linear_df$precloacal_vert != 0 & 
-                       linear_df$genus == "Anilios" &       # only Anilios species
+                       # linear_df$genus == "Anilios" &       # only Anilios species
                        linear_df$species != "Anilios_sp",]  # excluding unknown species
 
 # Calculate total vertical length and vertical ratio
@@ -66,6 +67,11 @@ nrow(vert_df)
 # Our data is presented by this many lineages
 unique(vert_df$species)  |> length()
 
+grep(unique(vert_df$species), pattern = "Anilios") |> length()
+grep(unique(vert_df$species), pattern = "Ramphotyphlops") |> length()
+grep(unique(vert_df$species), pattern = "Acutotyphlops") |> length()
+
+
 # Our sample size per species ranged from 
 paste0("sample size ranged from ", min(table(vert_df$species)), " to ", max(table(vert_df$species)))
 
@@ -79,7 +85,21 @@ unique(vert_df$institution)
 ## Number of scans by institution
 table(vert_df$institution)
 
+## Table S1
+# List of examined specimens
+table_s1 <- vert_df %>% 
+  dplyr::select(reg_no, species, institution, sex, total_vert, svl, midbody_diameter, tail_length) %>% 
+  arrange(species)
 
+table_s1$sex <- gsub(pattern = "u", "", table_s1$sex)
+table_s1$sex <- gsub(pattern = "j", "", table_s1$sex)
+table_s1$sex <- gsub(pattern = "?", "", table_s1$sex)
+# table_s1$sex <- gsub(pattern = "m", "male", table_s1$sex)
+# table_s1$sex <- gsub(pattern = "f", "female", table_s1$sex)
+table_s1$species <- gsub(pattern = "_", " ", table_s1$species)
 
+table_s1$institution <- gsub(pattern = "SAM", "SAMA", table_s1$institution)
 
+colnames(table_s1) <- c("Registration number", "Species", "Institution", "Sex", "Total vertebrae", "SVL", "Midbody diameter", "Tail length")
 
+write.csv(x = table_s1, file = "docs/supplement/table_s1.csv", row.names = FALSE)

### 2024-10-29 code/03_visualising_data.R

@@ -42,7 +42,6 @@ load('data/script_generated_data/subset_mcmctree_shape.rda')
 anilios_tree <- ape::drop.tip(sub_phy, tip = c("Ramphotyphlops_multilineatus","Acutotyphlops_subocularis"))
 anilios_tree$tip.label
 
-dev.off()
 
 b_tree <- sub_phy
 b_tree$tip.label <- gsub(pattern = "Anilios_", replacement = "A. ", b_tree$tip.label)
@@ -288,14 +287,14 @@ aspect_v[(which.min(aspect_v))]
 aspect_v[(which.max(aspect_v))]
 
 
-table_s2 <- vert_df %>% 
+table_s3 <- vert_df %>% 
   dplyr::group_by(species) %>% 
   dplyr::summarise(mean_aspect = mean(total_length/midbody_diameter),
                    max_aspect = max(total_length/midbody_diameter),
                    min_aspect = min(total_length/midbody_diameter),
                    sample_n = n()) 
 
-write.csv(table_s2, file = "assets/tables/aspect_ratios_summary", row.names = F)
+write.csv(table_s3, file = "assets/tables/aspect_ratios_summary.csv", row.names = F)
 
 ## This is using old code where we can't adjust the ylim
 
@@ -422,6 +421,21 @@ pleomerism.pgls <- geomorph::procD.pgls(max_tbl ~ max_vert, phy = anilios_tree)
 summary(pleomerism.pgls)
 coefficients(pleomerism.pgls)
 sp_labs <- gsub(pattern = "Anilios_", replacement = "", x = anilios_data$species)
+
+
+pleomerism.pgls <- geomorph::procD.pgls(max_tbl ~ max_vert, phy = anilios_tree)
+## Physignal
+max_tbl.physig <- physignal(A = max_tbl, phy = anilios_tree, iter = 999)
+physignal.z(A = max_tbl, phy = anilios_tree, iter = 999, lambda = "all")
+max_tbl.physig
+
+max_vert.physig <- physignal(A = max_vert, phy = anilios_tree, iter = 999)
+physignal.z(A = max_vert, phy = anilios_tree, iter = 999, lambda = "all")
+
+
+
+fit_pleomerism <- phylolm(max_tbl ~ max_vert, phy = anilios_tree, model="lambda")
+summary(fit_pleomerism)
 
 
 ### PLOTTING

### 2024-10-31 manuscript.Rmd
update one sentence in conclusion

@@
 We found a lack of pleomerism in the genus, with exceptionally large species possessing a lower number of vertebrae than expected for their body size.
-Our findings indicate that body size and shape evolution in this clade has been facilitated by changes to development, whereby the production of vertebral segments during early development is dissociated from somatic growth, likely driven by ecological adaptations.
+Our findings indicate that body size and shape evolution in this clade has been influenced by changes to development, whereby vertebral segmentation during early development is dissociated from somatic growth, likely driven by ecological adaptations.
 This study adds to a growing body of literature on the axial skeleton of snakes, a field that has largely been neglected.

### 2024-11-05 manuscript.Rmd
corrections pre submission

@@
 # title: "Dissociation between number of vertebrae and body size in Australian blindsnakes"
-title: "Testing the correlates of vast vertebrae number disparity in Australian blindsnakes"
+title: "Beyond body length: Uncovering the drivers of vertebra number variation in Australian blindsnakes"
 date: "Version `r Sys.Date()`"
@@
 SH and AA thank Queensland X-ray for provision of imaging for Queensland Museum specimens.
-SK thanks the Australian Research Council for ongoing support.
+JSK thanks the Australian Research Council for ongoing support.
 ST PhD Scholarship was supported by the Australian Government Research Training Program Scholarship at the Australian National University.
@@
 
-The evolution of the axial skeleton is important in understanding the evolution of body plans in vertebrates.
+The axial skeleton is important in understanding the evolution of body plans in vertebrates.
 Changes in the vertebral skeleton can influence body form and modes of locomotion.
 Elongate animals commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism.
-However, some groups of vertebrates show a lack of pleomerism, suggesting it is possible to dissociate the production of segment and somatic growth, but what factors influence this pattern is not well understood.
+However, some groups of vertebrates lack pleomerism, suggesting it is possible to dissociate the production of segmental and somatic growth, but the factors that may influence this pattern are not well understood.
 Snakes offer a great model to study axial evolution because the number of vertebrae vary greatly between species and they exhibit a wide range of body size and shape.
 Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology can explain the variation in the number of vertebrae.
-We obtained X-ray scans and morphometric data for 42 described species of *Anilios* blindsnakes and evaluated their correlation with each other and ecological variables.
+We obtained X-ray scans and morphometric data for 349 museum specimens from 48 lineages of Australo-Papuan blindsnakes and close relatives, and evaluated their correlation with each other and ecological variables.
 We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebra ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species.
 We also found that species that have a greater number of vertebrae for their size are generally found in warmer environments and environments with harder-to-penetrate soil substrate.
-Our results extend our understanding of the axial evolution of basal snakes and the macroevolutionary dynamics of Australian blindsnakes.
+Our results extend our understanding of the axial evolution of snakes and the macroevolutionary dynamics of Australian blindsnakes.
 
@@
 Axial skeleton evolution is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised.
-Essentially, vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryogenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009].
+Vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryogenesis.
+The speed of this process determines the size and number of somites, producing what will become the axial skeleton  [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009].
 Elongate animals are expected to show a positive correlation between maximum adult body length and number of vertebrae, known as pleomerism [@lindseyPleomerism1975].
-While pleomerism was considered the norm [@lindseyPleomerism1975;@wakeComparative1966;@pollyTesting2001], more recent studies have identified exceptions to the pleomerism rule in moray eels [@reeceEvolutionary2013], large species of snakes [@headDissociation2007;@lindellEvolution1994], salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].
+While pleomerism was considered the norm [@lindseyPleomerism1975;@wakeComparative1966;@pollyTesting2001], more recent studies have identified exceptions to the pleomerism rule in moray eels [@reeceEvolutionary2013], some clades of snakes [@headDissociation2007;@lindellEvolution1994;@sherrattVertebral2019], and salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].
 
-The lack of pleomerism in some species suggests it is possible to disassociate segment production with body growth, and thus reveals the developmental processes underlying evolutionary patterns.
-Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in many snakes [@alexanderPattern1966] and so ventral scales have often been used as a proxy for number of vertebrae [@lindellEvolution1994;@shineVertebral2000].
-However, this 1:1 correlation is absent in file snakes, *Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not conform to pleomerism.
+The lack of pleomerism in some species suggests it is possible to disassociate segment production with body growth, and thus reveal the developmental processes underlying evolutionary patterns.
+Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in most snakes [@alexanderPattern1966] and so ventral scales are commonly used as a proxy for number of vertebrae [@lindellEvolution1994;@shineVertebral2000].
+However, this 1:1 correlation is absent in file snakes, *Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not show pleomerism.
 Snakes are therefore a good model to study axial evolution because they have great diversity in number of vertebrae and a wide range of adult body sizes and shapes but without a limb skeleton to confound signals of locomotor adaptation.
-Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattVertebral2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024].
+Recent studies have indicated evolutionary diversity in number of vertebrae is related to ecological diversification in elapid snakes [@sherrattVertebral2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024].
 The number of vertebrae has also been shown to influence snake locomotion, whereby the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997].
-Despite these recent advances, there have been no studies that have quantified the number of vertebrae across all the species in a diverse assemblage of fossorial snakes to test for variation and potential ecological drivers.
+Despite these recent advances, no studies that have quantified the number of vertebrae across all the species in a diverse assemblage of fossorial snakes to test for variation and potential ecological drivers.
 
@@
 
-The Australian blindsnake radiation *Anilios* comprises47 described species distributed across all major biomes of Australia [@tiatragulAustralia2023;@tiatragulPaleoenvironmental2023].
+The Australian blindsnake radiation *Anilios* comprises 47 described species distributed across all major biomes of Australia [@tiatragulAustralia2023;@tiatragulPaleoenvironmental2023].
 This radiation displays considerable morphological diversity which ranges from narrow-and-long to robust-and-wide species [@tiatragulMorphological2024;@ehmannFamily1993].
@@
 
-We examined 349 specimens representing 48 typhlopid lineages including 44 *Anilios* [@tiatragulAustralia2023;uetzQuarter2021], *Ramphotyphlops multilineatus* (Schlegel, 1839), *R.
+We examined 349 specimens representing 48 typhlopid lineages including 44 *Anilios* [@tiatragulAustralia2023;@uetzQuarter2021], *Ramphotyphlops multilineatus* (Schlegel, 1839), *R.
 acuticauda* (Peters, 1878), *R.
@@
 
-To obtain vertebral counts, we X-ray scanned the specimens on the following machines and institutions: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n  =  87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n = 129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (n = 90), and Queensland X-Ray (n = 35).
+To obtain vertebral counts, we X-ray scanned the specimens on the following machines and institutions: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n  =  87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n = 129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (n = 90), and FCR0607 (Fujifilm, Tokyo, Japan) at Queensland X-Ray (n = 35).
 
@@
 Temperature also directly influences multiple axes of reptile development [@nobleDevelopmental2018;@boothInfluence2006;@feinerDevelopmental2018], which influences vertebra development [@martin-del-campoHox2018].
-Bulk density values > 1.6 g/cm<sup>3</sup> indicate soil that restrict penetration by plant roots, while values <0.5 indicate looser, more organic soil [@mckenzieSoil2002].
+Bulk density values > 1.6 g/cm<sup>3</sup> indicate soil that restricts penetration by plant roots, while values <0.5 indicate looser, more organic soil [@mckenzieSoil2002].
 We extracted soil bulk density values between 0 — 5 cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html) and mean annual temperature from WorldClim v.2 [@fickWorldClim2017] raster files that contained modelled approximations at the resolution of 3 arc-seconds using the "extract" function in the R package *`raster`* v.3.6-3 [@hijmansRaster2022].
@@
 We did this by regressing the species maximum number of total vertebrae against maximum TBL.
-We also calculated the degree of phylogenetic signal from these traits separately using the "physignal.z" function in `geomorph`.
+We also calculated the degree of phylogenetic signal from these traits separately using the K metric [@blombergTesting2003] and λ metric [@pagelInferring1999] implemented with the the "physignal.z" function in `geomorph` and "phylosig" function in `phytools` v 2.3 [@revellPhytools2024], respectively.
 Next, we tested whether the species mean aspect ratio is correlated with mean vertebra ratio (mean number of vertebrae/mean TBL) and mean number of vertebrae.
@@
 torresianus*, which has the greatest maximum TBL, has relatively few vertebrae.
-We found phylogenetic signal in both maximum number of total vertebrae (Z = 3.70, P = 0.001, K = 1.2, lambda = 1) and maximum TBL (Z = 3.02, P = 0.002, K = 1.0, lambda = 1).
+We found more phylogenetic signal than expected under Brownnian motion in both maximum number of total vertebrae (Z = 3.70, P = 0.001, K = 1.2; λ = 1.12, log likelihood(λ) = -197.20, LR(λ=0) = 11.82, P = 0.0006) and maximum TBL (Z = 3.02, P = 0.002, K = 1.0; λ = 1.14, log likelihood(λ) = -221.98, LR(λ=0) = 7.42, P = 0.006).
 
@@
 We also found that vertebral ratio correlates positively with mean annual temperature and negatively with soil compactness.
-These results have important implications for understanding the axial evolution of basal snakes and the macroevolutionary dynamics of Australian blindsnakes.
+These results have important implications for understanding the axial evolution of snakes and the macroevolutionary dynamics of Australian blindsnakes.
 
@@
 For example, the absence of pleomerism in *Hydrophis* sea snakes has been attributed to the increase in the number of smaller vertebrae in the forebody region compared to larger vertebrae in the hindbody region [@sherrattMacroevolution2022;@sherrattVertebral2019].
-Similarly, colubrids have also been shown to have different regionalisation patterns in the vertebral column that contributes to elongation in different ways [@nash-hahnModification2024].
+Similarly, colubrids have also been shown to leverage regionalisation in the vertebral column to achieve elongation in different ways [@nash-hahnModification2024].
 The position of the heart is a key marker for delimiting the regions of the snake axial column [@hamptonHeart2022;@hamptonRegionalization2023].
@@
 However, they have been shown to deviate from the expected 1:1 ratio of ventral scale number to vertebrae [@alexanderPattern1966], suggesting a lack of pleomerism.
-We recommend future research to comprehensively investigate the prevalence of pleomerism among burrowing limbless and limb-reduced vertebrates to improve our understanding of the evolution of axial system.
+We recommend future research to comprehensively investigate the prevalence of pleomerism among burrowing limbless and limb-reduced vertebrates to improve our understanding of the evolution of the axial system.
 
@@
 In blindsnakes, however, movement speed may not be as important as how much push force they can generate when burrowing.
-Snakes with relatively wider bodies are generally able to produce greater push force for their body size [@herrelBurrowing2021], while species with narrower bodies will be able to use pre-existing small crevices and galleries without needing to expend any energy.
+Snakes with relatively wider bodies are generally able to produce greater push force for their body size [@herrelBurrowing2021], while species with narrower bodies will be able to use pre-existing small crevices and galleries with a minimum amount of energy.
 In regard to prey consumption, @poughSpecializations1983 suggest that stout species with large heads (i.e.
@@
 
-In a related study, @tiatragulMorphological2024 found that Australian blindsnake species with smaller absolute body size generally occupy warmer areas than larger species tend to occupy cooler regions.
+In a related study, @tiatragulMorphological2024 found that Australian blindsnake species with smaller absolute body size generally occupy warmer areas than larger species, which tend to occupy cooler regions.
 Here, we observe that vertebra ratio (raw number of vertebrae/SVL) is positively correlated with temperature, indicating species that have more vertebrae than expected from their body size are found in warmer places.
 In his work on rattlesnakes, @klauberRattlesnakes1982 noted that smaller-sized subspecies often retain the same number of vertebrae as their larger closest relative, which results in the smaller subspecies having a relatively greater vertebra ratio.
-We also observed maximum number of vertebrae in our dataset contain phylogenetic signal.
+We also observed strong phylogenetic signal in the maximum number of vertebrae in our dataset.
 An explanation for this pattern is that size is more likely to be under greater selection than vertebral number, and there is negligible disadvantage for a snake in having too many vertebrae for their body size [@lindellEvolution1994].

### 2024-11-05 code/03_visualising_data.R
add phylogenetic signal test

@@ -425,18 +425,25 @@ sp_labs <- gsub(pattern = "Anilios_", replacement = "", x = anilios_data$species
 
 pleomerism.pgls <- geomorph::procD.pgls(max_tbl ~ max_vert, phy = anilios_tree)
 ## Physignal
-max_tbl.physig <- physignal(A = max_tbl, phy = anilios_tree, iter = 999)
-physignal.z(A = max_tbl, phy = anilios_tree, iter = 999, lambda = "all")
-max_tbl.physig
+# max_tbl.physig <- physignal(A = max_tbl, phy = anilios_tree, iter = 999)
+# max_tbl.physig
+max_tbl.physigz <- physignal.z(A = max_tbl, phy = anilios_tree, iter = 999, lambda = "mean")
+summary(max_tbl.physigz)
 
-max_vert.physig <- physignal(A = max_vert, phy = anilios_tree, iter = 999)
-physignal.z(A = max_vert, phy = anilios_tree, iter = 999, lambda = "all")
-
-
+# max_vert.physig <- physignal(A = max_vert, phy = anilios_tree, iter = 999)
+max_vert.physigz <- physignal.z(A = max_vert, phy = anilios_tree, iter = 999, lambda = "mean")
+summary(max_vert.physigz)
 
 fit_pleomerism <- phylolm(max_tbl ~ max_vert, phy = anilios_tree, model="lambda")
 summary(fit_pleomerism)
 
+# Using phylosig
+phylosig_maxtbl<-phytools::phylosig(tree = anilios_tree, x = max_tbl, method = "lambda", test=TRUE)
+# phylosig_maxtbl<-phytools::phylosig(tree = anilios_tree, x = log(max_tbl), method = "lambda", test=TRUE)
+phylosig_maxtbl
+
+phylosig_maxvert<-phytools::phylosig(tree = anilios_tree, x = max_vert, method = "lambda", test=TRUE)
+phylosig_maxvert
 
 ### PLOTTING
 # Log max body length against log max number of vertebrae

### 2024-11-06 manuscript.Rmd
prep submit
trim abstract

@@
 Elongate animals commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism.
-However, some groups of vertebrates lack pleomerism, suggesting it is possible to dissociate the production of segmental and somatic growth, but the factors that may influence this pattern are not well understood.
-Snakes offer a great model to study axial evolution because the number of vertebrae vary greatly between species and they exhibit a wide range of body size and shape.
-Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology can explain the variation in the number of vertebrae.
-We obtained X-ray scans and morphometric data for 349 museum specimens from 48 lineages of Australo-Papuan blindsnakes and close relatives, and evaluated their correlation with each other and ecological variables.
+However, some groups of vertebrates lack pleomerism, suggesting it is possible to dissociate the number of vertebrae and somatic growth, but the factors that influence this pattern are not well understood.
+Snakes offer a great model to study axial evolution because they exhibit a wide range of intraspecific variation in vertebra number and interspecific variation in body size, shape and ecology.
+Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology correlate with vertebra number.
+To do so, we assembled a comprehensive eco-morphological dataset including X-ray scans for 48 lineages of Australo-Papuan typhlopid snakes.
 We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebra ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species.
-We also found that species that have a greater number of vertebrae for their size are generally found in warmer environments and environments with harder-to-penetrate soil substrate.
-Our results extend our understanding of the axial evolution of snakes and the macroevolutionary dynamics of Australian blindsnakes.
+We also found that species that have a greater number of vertebrae for their size are found in warmer environments and in harder-to-penetrate soil substrate.
+These results extend our understanding of the axial evolution of snakes and the macroevolutionary dynamics of Australian blindsnakes.
 

### 2024-11-08 manuscript.Rmd
update before initial submission

@@
 SH: Resources, Investigation
-SK: Supervision, Funding acquisition, Writing - reviewing and editing.
+JSK: Supervision, Funding acquisition, Writing - reviewing and editing.
 ES: Conceptualisation, Methodology, Writing - reviewing and editing.
@@
 Snakes offer a great model to study axial evolution because they exhibit a wide range of intraspecific variation in vertebra number and interspecific variation in body size, shape and ecology.
-Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology correlate with vertebra number.
+Here, we investigate the prevalence of pleomerism in a continental radiation of blindsnakes and test whether body shape and ecology correlate with vertebra number.
 To do so, we assembled a comprehensive eco-morphological dataset including X-ray scans for 48 lineages of Australo-Papuan typhlopid snakes.
@@
 For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum TBL (PGLS, slope = 0.28, intercept = 256, R^2^ = 0.02, F = 0.74, Z = 0.30, P = 0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig.
-\@ref(fig:fig-tbl-vert) and **Fig S1**).
+\@ref(fig:fig-tbl-vert)).
 This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae.
@@
 We found that vertebra ratio is positively correlated with mean aspect ratio (PGLS, slope = -0.0048, intercept = 0.73, R^2^ = 0.17, F = 7.09, Z = 2.20, P = 0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig.
-\@ref(fig:fig-vert-aspect-ratios)).
+\@ref(fig:fig-vert-aspect-ratios) and Fig.
+S1).
 We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope = -1.8, intercept = 120.95, R^2^ = 0.69, F = 79.39, Z = 5.30, P<0.001).

### 2024-11-08 code/03_visualising_data.R
change axis title

@@ -637,13 +637,13 @@ par(mfrow=c(2,2))
 ## Vert ratio v Annual mean temp
 phylomorphospace(anilios_tree, anilios_data[,c("temp_mean", "ver_rati")], 
                  bty="n", label="horizontal", node.size=c(0.4,1.3),
-                 ylab = "Vertebrae ratio", xlab = "")
+                 ylab = "Vertebra ratio", xlab = "")
 abline(a = coefficients(temp.pgls)[1], b = coefficients(temp.pgls)[2])
 
 ## VERT RATIO VS Soil bulk density
 phylomorphospace(anilios_tree, anilios_data[,c("max_bulk", "ver_rati")], 
                  bty="n", label="horizontal", node.size=c(0.4,1.3),
-                 ylab = "Vertebrae ratio", xlab = "")
+                 ylab = "Vertebra ratio", xlab = "")
 abline(a = coefficients(soil.pgls)[1], b = coefficients(soil.pgls)[2])
 
 ## VERTEBRAE NUMBER VS ECOLOGY
@@ -654,12 +654,12 @@ anilios_data$aspect <- aspect_ratio[anilios_data$species]
 # vert vs temperature
 phylomorphospace(anilios_tree, anilios_data[,c("temp_mean", "tot_vert")], 
                  bty="n", label="horizontal", node.size=c(0.4,1.3),
-                 ylab = "Mean vertebrae number", xlab = "Mean annual temperature (°C)")
+                 ylab = "Mean vertebra number", xlab = "Mean annual temperature (°C)")
 abline(a = coefficients(number_temp.pgls)[1], b = coefficients(number_temp.pgls)[2])
 
 phylomorphospace(anilios_tree, anilios_data[,c("max_bulk", "tot_vert")], 
                  bty="n", label="horizontal", node.size=c(0.4,1.3),
-                 ylab = "Mean vertebrae number", xlab = expression(Max~soil~bulk~density~(g/cm^3)))
+                 ylab = "Mean vertebra number", xlab = expression(Max~soil~bulk~density~(g/cm^3)))
 abline(a = coefficients(number_soil.pgls)[1], b = coefficients(number_soil.pgls)[2])
 
 dev.off()

### 2024-12-02 manuscript.Rmd
update

@@
 ### Relationships between vertebra ratio, number of vertebrae, and aspect ratio
-We found that vertebra ratio is positively correlated with mean aspect ratio (PGLS, slope = -0.0048, intercept = 0.73, R^2^ = 0.17, F = 7.09, Z = 2.20, P = 0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig.
+We found that vertebra ratio is positively correlated with mean aspect ratio (PGLS, slope = 0.0048, intercept = 0.73, R^2^ = 0.17, F = 7.09, Z = 2.20, P = 0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig.
 \@ref(fig:fig-vert-aspect-ratios) and Fig.
 S1).
-We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope = -1.8, intercept = 120.95, R^2^ = 0.69, F = 79.39, Z = 5.30, P<0.001).
+We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope = 1.8, intercept = 120.95, R^2^ = 0.69, F = 79.39, Z = 5.30, P<0.001).
 

### 2024-12-02 code/00_stats_for_paper.R

@@ -82,8 +82,12 @@ mean(table(vert_df$species))
 ## Institution we measured from
 unique(vert_df$institution)
 
+unique(vert_df$species)
+
+
 ## Number of scans by institution
 table(vert_df$institution)
+sum(table(vert_df$institution))
 
 ## Table S1
 # List of examined specimens
@@ -103,3 +107,6 @@ table_s1$institution <- gsub(pattern = "SAM", "SAMA", table_s1$institution)
 colnames(table_s1) <- c("Registration number", "Species", "Institution", "Sex", "Total vertebrae", "SVL", "Midbody diameter", "Tail length")
 
 write.csv(x = table_s1, file = "docs/supplement/table_s1.csv", row.names = FALSE)
+
+
+sum(table_s1$`Total vertebrae`)

### 2024-12-02 code/03_visualising_data.R

@@ -422,8 +422,6 @@ summary(pleomerism.pgls)
 coefficients(pleomerism.pgls)
 sp_labs <- gsub(pattern = "Anilios_", replacement = "", x = anilios_data$species)
 
-
-pleomerism.pgls <- geomorph::procD.pgls(max_tbl ~ max_vert, phy = anilios_tree)
 ## Physignal
 # max_tbl.physig <- physignal(A = max_tbl, phy = anilios_tree, iter = 999)
 # max_tbl.physig
@@ -460,6 +458,7 @@ phylomorphospace(anilios_tree, anilios_vert[,c("max_total_vert", "max_tbl")],
                  ylab = "Max. total body length (mm)", xlab = "Maximum number of total vertebrae")
 # Add slope from pgls (but this is just for species in tree)
 abline(a = coefficients(pleomerism.pgls)[1], b = coefficients(pleomerism.pgls)[2])
+summary(pleomerism.pgls)[1]
 
 # add species not in tree to overlay text next to it
 points(y = not_tree_data$max_tbl, x = not_tree_data$max_total_vert, cex = 2, col="blue", pch = 20)

### 2025-04-10 manuscript.Rmd
update latest manuscript before acceptance

@@
 Phylogeny projected into space using maximum likelihood ancestral state estimation.
-Species not included in the phylogeny are shown in purple.', out.width = "124mm", out.height =  "165mm"}
+Species not included in the phylogeny are shown in purple.', out.width = "152mm", out.height =  "196mm"}
 
@@
 Species mean annual temperature and max soil bulk density values are calculated based on occurrence data.
-Trend lines represent slopes from PGLS', out.width="165mm", out.height= "117mm"}
+Dashed trend lines represent slopes from PGLS', out.width="165mm", out.height= "117mm"}
 

### 2025-04-10 code/00_stats_for_paper.R
stats for paper update

@@ -75,7 +75,6 @@ grep(unique(vert_df$species), pattern = "Acutotyphlops") |> length()
 # Our sample size per species ranged from 
 paste0("sample size ranged from ", min(table(vert_df$species)), " to ", max(table(vert_df$species)))
 
-# Median | mean of species are:
 paste("median =", median(table(vert_df$species)))
 mean(table(vert_df$species))
 

### 2025-04-10 code/03_visualising_data.R
clean up script

@@ -1,19 +1,15 @@
-# 03_visualising_data.R
-# Putter Tiatragul
-# Feb 2024
-# Visualise relationship between number of vertebrae and other traits
-# 
+## METADATA ===============================================================
+## Filename: 03_visualising_data.R
+## Description: Visualise relationship between number of vertebrae and other traits. And test for correlations
+## 
+## R version: 4.4.0 for Windows
+## Author: Putter Tiatragul (sarin.tiatragul@anu.edu.au)
+##=======================================================================##
 
 # libraries ---------------------------------------------------------------
-# Load packages with two lines
-
 library(ggplot2); library(dplyr); library(phytools); 
 library(geiger); library(ape); library(geomorph); 
-# library(geiger)
-# library(R.utils)
 library(phylolm)
-# library(factoextra)
-
 '%notin%' <- Negate('%in%')
 
 # source ------------------------------------------------------------------
@@ -26,13 +22,9 @@ body_shape_df <- read.csv('data/script_generated_data/blindsnake_logbodyshape_ra
 PC_log_body_ratio_df <- read.csv('data/script_generated_data/blindsnake_sp_pc_logbodyrat.csv')
 PC_log <- read.csv('data/script_generated_data/log_body_shape_ratio_pc_all.csv')
 
-
-# Need in this script
-
 # Full individual data
 linear_df <- read.csv('data/script_generated_data/blindsnake_body_traits.csv')
 linear_df <- linear_df[linear_df$reg_no %notin% c("R.8515","R.127956"),] ### Dropped these samples because mis-id as troglodytes but in Queensland. Probably torresianus or broomi
-
 
 # Summary data
 anilios_data <- read.csv(file = 'data/script_generated_data/anilios_summary_data.csv', row.names = 1)
@@ -42,14 +34,12 @@ load('data/script_generated_data/subset_mcmctree_shape.rda')
 anilios_tree <- ape::drop.tip(sub_phy, tip = c("Ramphotyphlops_multilineatus","Acutotyphlops_subocularis"))
 anilios_tree$tip.label
 
-
 b_tree <- sub_phy
[…]

### 2025-04-10 code/03_vert_profile.R
change name

@@ -1,43 +1,0 @@
-# 03_vert_profile.R
-# 2024-03-25
-# Adapted from Emma's script for Measuring and analysing vertebral length of sea snakes along the axial column
-# October 2021
-
-
-### READ ME! ###
-# Comments are instructions
-# make changes where indicated
-
-##### DO FIRST! ######
-# Set working directory
-# setwd(" ") # paste between the quotation the path to working directory, OR
-
-
-# Data --------------------------------------------------------------------
-# Read in the three sets of replicate measurements for a single specimen
-
-# Ligatus
-lig1 <- read.csv("data/x_ray/lig_R00050_ST_1.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
-lig2 <- read.csv("data/x_ray/lig_R00050_ST_2.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
-lig3 <- read.csv("data/x_ray/lig_R00050_ST_3.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
-
-# Grypus
-gry1 <- read.csv("data/x_ray/gry_R146673_ST_1.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
-gry2 <- read.csv("data/x_ray/gry_R146673_ST_2.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
-gry3 <- read.csv("data/x_ray/gry_R146673_ST_3.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
-
-# Diversus
-div1 <- read.csv("data/x_ray/div_R51126_ST_1.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
-div2 <- read.csv("data/x_ray/div_R51126_ST_2.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
-div3 <- read.csv("data/x_ray/div_R51126_ST_3.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
-
-# Plot  -------------------------------------------------------------------
-source('code/utility/func_vert_profile_plotter.R')
-
-pdf(file = "output/vertebral_profile.pdf", width = 8.3, height = 11.7)
-par(mfrow=c(3,1))
-# vert_raw_plotter(lig1, lig2, lig3, "A. ligatus R00050")
[…]

### 2025-04-10 code/01_vert_profile.R
change name

@@ -1,0 +1,43 @@
+## METADATA ===============================================================
+## Filename: 01_vert_profile.R
+## Description: Adapted from Emma Sherratt's script for Measuring and analysing vertebral length of sea snakes along the axial column
+## March 2024
+## Author: Sarin Tiatragul (sarin.tiatragul@anu.edu.au)
+##=======================================================================##
+
+### READ ME! ###
+# Comments are instructions
+# make changes where indicated
+
+##### DO FIRST! ######
+# Set working directory
+# setwd(" ") # paste between the quotation the path to working directory, OR
+
+# Data --------------------------------------------------------------------
+# Read in the three sets of replicate measurements for a single specimen
+
+# Ligatus
+lig1 <- read.csv("data/x_ray/lig_R00050_ST_1.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+lig2 <- read.csv("data/x_ray/lig_R00050_ST_2.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+lig3 <- read.csv("data/x_ray/lig_R00050_ST_3.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+
+# Grypus
+gry1 <- read.csv("data/x_ray/gry_R146673_ST_1.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+gry2 <- read.csv("data/x_ray/gry_R146673_ST_2.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+gry3 <- read.csv("data/x_ray/gry_R146673_ST_3.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+
+# Diversus
+div1 <- read.csv("data/x_ray/div_R51126_ST_1.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+div2 <- read.csv("data/x_ray/div_R51126_ST_2.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+div3 <- read.csv("data/x_ray/div_R51126_ST_3.csv", header=TRUE, row.names = 1)[,(c("X", "Y"))] 
+
+# Plot  -------------------------------------------------------------------
+source('code/utility/func_vert_profile_plotter.R')
+
+pdf(file = "output/vertebral_profile.pdf", width = 8.3, height = 11.7)
+par(mfrow=c(3,1))
[…]

### 2025-04-10 code/04_model_fitting.R
clean up script, we didn't need this actually

@@ -1,7 +1,9 @@
-# 04_model_fitting.R
-# Sarin Tiatragul 
-# July 2022 modified Feb 2023
-# Combine phenotypic data with environment data and plot
+## METADATA ===============================================================
+## Filename: 04_model_fitting.R
+## Description: Combine phenotypic data with environment data and plot
+## July 2022 modified Feb 2023
+## Author: Sarin Tiatragul (sarin.tiatragul@anu.edu.au)
+##=======================================================================##
 
 # libraries ---------------------------------------------------------------
 # Load packages with two lines
@@ -28,7 +30,6 @@ PC_log <- read.csv('data/script_generated_data/log_body_shape_ratio_pc_all.csv')
 # Load full tree, subsetted tree, and subset 
 load('data/script_generated_data/subset_mcmctree_shape.rda')
 
-
 anilios_tree <- ape::drop.tip(sub_phy, tip = c("Ramphotyphlops_multilineatus","Acutotyphlops_subocularis"))
 
 anilios_data <- read.csv(file = 'data/script_generated_data/anilios_summary_data.csv', row.names = 1)
@@ -49,7 +50,8 @@ vert_df <- linear_df %>%
 
 plot(x = vert_df$total_length, y = vert_df$total_vert, xlab = "Total vertebrae #", ylab = "SVL", bty = "n")
 
-# No clear pattern of vertebrae and total length, indicates there may be some species that elongate their vertebrae and others that shorten their vertebrae relative to the mean. 
+# No clear pattern of vertebrae and total length, indicates there may be some species that elongate their vertebrae
+# and others that shorten their vertebrae relative to the mean. 
 
 # Test for sexual dimorphism ----------------------------------------------
 
@@ -87,14 +89,13 @@ ggplot(filtered_vert_df, aes(x = total_vert, y = total_length, colour = sex)) +
   xlab("Total length (mm)") +
   ylab("# of vertebrae")  
 
-
-
 library(ggtree)
 filtered_vert_df$length_by_vert <- filtered_vert_df$total_length/ filtered_vert_df$total_vert
 
 vert_data <- vert_df %>% dplyr::group_by(species) %>% 
   dplyr::summarise(mean_svl = mean(svl),
                    mean_total_length = mean(total_length),
[…]
